---
title: "SCRATCH-TCR: Repertoire Analysis Report"
author: "Syed Shujaat Ali Zaidi"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    theme: cosmo
    df-print: paged
    css: repertoire_report.css
execute:
  echo: false
  warning: false
  message: false


params:
  # ======================================================
  # Inputs
  # ======================================================
  seurat_rds: "data/seurat_with_consensus_clonotype_clusters.rds"
  export_cells_file: "data/consensus_export_cells.tsv"
  outdir: "Repertoire_Report"

  # Optional inputs
  metadata_file: ""
  previous_summary_file: ""

  # ======================================================
  # Output controls
  # ======================================================
  data_dir: "data"
  tables_dir: "tables"
  figures_dir: "figures"
  save_tables: true
  save_figures: true
  save_updated_seurat: false
  figure_format: "png"
  figure_width: 10
  figure_height: 7
  figure_dpi: 300
  base_size: 12

  # ======================================================
  # Metadata mapping
  # ======================================================
  label_col: ""
  sample_col: "META_SAMPLE"
  patient_col: "META_PATIENT"
  condition_col: "META_TIMECOND"
  timepoint_col: "META_TIMEPOINT"
  batch_col: "META_BATCH"
  clone_id_col: "clone_id"
  clone_size_col: "clone_size"
  paired_tcr_col: "paired_tcr"
  has_tcr_col: "has_tcr"
  consensus_cluster_col: "consensus_cluster"

  label_candidates: "annot;Annotation;celltype;CellType;predicted_labels;predicted.celltype.l2;predicted.celltype.l1;celltypist;celltypist_label;azimuth_labels;seurat_clusters"
  sample_candidates: "META_SAMPLE;sample;orig.ident;sample_id;Sample;SampleID"
  patient_candidates: "META_PATIENT;patient_id;patient;Patient;subject;donor;case_id"
  condition_candidates: "META_TIMECOND;condition;Condition;group;Group;status"
  timepoint_candidates: "META_TIMEPOINT;timepoint;Timepoint;visit;Visit;day;Day;META_TIMECOND"
  batch_candidates: "META_BATCH;batch;Batch;library;Library;run;Run;lane;Lane"
  clone_id_candidates: "clone_id;CTaa;clonotype;clone"
  clone_size_candidates: "clone_size;CloneSize;clone_n"
  paired_tcr_candidates: "paired_tcr;paired;is_paired"
  has_tcr_candidates: "has_tcr;hasTCR;tcr_positive"
  consensus_cluster_candidates: "consensus_cluster;consensus;consensus_id"

  # ======================================================
  # Repertoire / binning controls
  # ======================================================
  min_cells_per_clone_plot: 2
  top_n_shared_clones: 50
  top_n_flux_clones: 30
  shareability_min_n_groups: 2
  min_cells_per_group: 10
  top_n_clusters: 20
  

  # Clone burden / homeostasis bins
  clone_bin_breaks: "0;1e-4;0.001;0.01;0.1;1.0"
  clone_bin_labels: "Rare;Small;Medium;Large;Hyperexpanded"
  use_relative_clone_frequencies: true

  # Similarity / overlap
  overlap_metric: "jaccard"
  diversity_metrics: "shannon;simpson;inv_simpson;richness"
  longitudinal_require_ordered_time: false

  # ======================================================
  # Figure/report toggles
  # ======================================================
  show_diversity_boxplots: true
  show_clone_burden_barplot: true
  show_rank_abundance: true
  show_shareability_barplot: true
  show_shareability_heatmap: true
  show_sample_overlap_heatmap: true
  show_phenotypic_flux: true
  show_longitudinal_clone_tracking: true
  show_consensus_cluster_burden: true
  show_summary_tables: true

  report_label: "Repertoire Analysis"
---


###############################################################################
# Read the integrated Seurat object or exported cell-level table and standardize clonotype, metadata, and annotation fields.
# Compute repertoire summaries including clone frequencies, diversity metrics, clonal burden, and cross-sample shareability.
# Quantify sample-to-sample overlap and identify clonotypes spanning multiple biological groups or phenotypic states.
# Generate publication-quality figures for diversity, burden, shareability, overlap, phenotypic flux, and optional longitudinal tracking.
# Export repertoire summary tables and figures for the final master summary report.
###############################################################################
# setup
```{r}
#| label: setup
suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(forcats)
  library(scales)
  library(glue)
  library(knitr)
  library(kableExtra)
  library(ComplexHeatmap)
  library(circlize)
  library(ggalluvial)
  library(patchwork)
})

options(stringsAsFactors = FALSE)

dir.create(params$outdir, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$data_dir), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$tables_dir), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$figures_dir), recursive = TRUE, showWarnings = FALSE)

`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0 && !all(is.na(a))) a else b

theme_scratch_pub <- function(base_size = 12) {
  theme_bw(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 2, hjust = 0),
      plot.subtitle = element_text(size = base_size, hjust = 0),
      axis.title = element_text(face = "bold"),
      axis.text = element_text(color = "black"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(linewidth = 0.2, color = "grey90"),
      strip.background = element_rect(fill = "grey95", color = "grey80"),
      strip.text = element_text(face = "bold"),
      legend.title = element_text(face = "bold"),
      legend.key = element_blank(),
      plot.caption = element_text(size = base_size - 2, color = "grey40")
    )
}

save_plot_safe <- function(plot_obj, filename,
                           width = params$figure_width,
                           height = params$figure_height,
                           dpi = params$figure_dpi) {
  if (!isTRUE(params$save_figures)) return(invisible(NULL))
  ggplot2::ggsave(
    filename = file.path(params$outdir, params$figures_dir, filename),
    plot = plot_obj,
    width = width,
    height = height,
    dpi = dpi,
    bg = "white",
    limitsize = FALSE
  )
}

save_table_safe <- function(df, filename) {
  if (!isTRUE(params$save_tables)) return(invisible(NULL))
  data.table::fwrite(df, file.path(params$outdir, params$tables_dir, filename), sep = "\t")
}

safe_read_table <- function(path) {
  if (is.null(path) || is.na(path) || path == "" || !file.exists(path)) return(NULL)
  ext <- tolower(tools::file_ext(path))
  if (ext %in% c("tsv", "tab", "txt")) {
    data.table::fread(path, sep = "\t")
  } else {
    data.table::fread(path)
  }
}

require_file <- function(path, msg = NULL) {
  if (!file.exists(path)) stop(msg %||% paste("Missing required file:", path))
  path
}

sanitize_param_string <- function(x) {
  if (is.null(x) || length(x) == 0) return(NULL)
  trimws(gsub("\\u00A0", " ", as.character(x)))
}

normalize_colnames <- function(df) {
  colnames(df) <- trimws(gsub("\\u00A0", " ", colnames(df)))
  df
}

split_candidate_string <- function(x) {
  if (is.null(x) || length(x) == 0 || is.na(x) || x == "") return(character())
  x <- gsub('"', "", as.character(x))
  x <- trimws(unlist(strsplit(x, ";", fixed = TRUE)))
  x[nzchar(x)]
}

parse_numeric_param_vector <- function(x) {
  vals <- split_candidate_string(x)
  as.numeric(vals)
}

parse_character_param_vector <- function(x) {
  split_candidate_string(x)
}

first_existing_col <- function(df, preferred, candidates = character()) {
  df <- normalize_colnames(df)
  preferred <- sanitize_param_string(preferred)

  if (length(candidates) == 1 && is.character(candidates)) {
    candidates <- split_candidate_string(candidates)
  }
  candidates <- trimws(gsub("\\u00A0", " ", as.character(candidates)))

  if (!is.null(preferred) && preferred != "" && preferred %in% colnames(df)) return(preferred)

  hits <- intersect(candidates, colnames(df))
  if (length(hits) > 0) return(hits[[1]])
  NULL
}

safe_n_unique <- function(x) length(unique(na.omit(x)))

calc_shannon <- function(x) {
  x <- x[x > 0]
  p <- x / sum(x)
  -sum(p * log(p))
}

calc_simpson <- function(x) {
  x <- x[x > 0]
  p <- x / sum(x)
  1 - sum(p^2)
}

calc_inv_simpson <- function(x) {
  x <- x[x > 0]
  p <- x / sum(x)
  1 / sum(p^2)
}

calc_richness <- function(x) {
  sum(x > 0)
}

calc_overlap_score <- function(a, b, metric = "jaccard") {
  a <- unique(na.omit(a))
  b <- unique(na.omit(b))
  inter <- length(intersect(a, b))
  union_n <- length(union(a, b))
  if (metric == "jaccard") {
    ifelse(union_n > 0, inter / union_n, NA_real_)
  } else {
    denom <- min(length(a), length(b))
    ifelse(denom > 0, inter / denom, NA_real_)
  }
}

save_rds_safe <- function(obj, filename) {
  saveRDS(obj, file.path(params$outdir, params$data_dir, filename))
}

save_rdata_safe <- function(..., filename) {
  save(..., file = file.path(params$outdir, params$data_dir, filename))
}

```

# loading Inputs
```{r}
#| label: read-inputs
require_file(params$seurat_rds, sprintf("Seurat RDS not found: %s", params$seurat_rds))

seu <- readRDS(params$seurat_rds)
export_cells_in <- safe_read_table(params$export_cells_file)
metadata_tbl <- safe_read_table(params$metadata_file)
previous_summary_tbl <- safe_read_table(params$previous_summary_file)

if (is.null(export_cells_in)) {
  export_cells_in <- seu@meta.data %>% tibble::rownames_to_column("cell_id")
}
```

# Resolving Columns
```{r}
#| label: resolve-columns
df0 <- normalize_colnames(as.data.frame(export_cells_in))

label_col <- first_existing_col(df0, params$label_col, params$label_candidates)
sample_col <- first_existing_col(df0, params$sample_col, params$sample_candidates)
patient_col <- first_existing_col(df0, params$patient_col, params$patient_candidates)
condition_col <- first_existing_col(df0, params$condition_col, params$condition_candidates)
timepoint_col <- first_existing_col(df0, params$timepoint_col, params$timepoint_candidates)
batch_col <- first_existing_col(df0, params$batch_col, params$batch_candidates)
clone_id_col <- first_existing_col(df0, params$clone_id_col, params$clone_id_candidates)
clone_size_col <- first_existing_col(df0, params$clone_size_col, params$clone_size_candidates)
paired_tcr_col <- first_existing_col(df0, params$paired_tcr_col, params$paired_tcr_candidates)
has_tcr_col <- first_existing_col(df0, params$has_tcr_col, params$has_tcr_candidates)
consensus_cluster_col <- first_existing_col(df0, params$consensus_cluster_col, params$consensus_cluster_candidates)

if (is.null(label_col)) stop("Could not resolve an annotation label column.")
if (is.null(sample_col)) stop("Could not resolve a sample column.")
if (is.null(clone_id_col)) stop("Could not resolve a clone ID column.")
if (is.null(clone_size_col)) stop("Could not resolve a clone size column.")
if (is.null(has_tcr_col)) stop("Could not resolve a has_tcr column.")
if (is.null(paired_tcr_col)) stop("Could not resolve a paired_tcr column.")

print(list(
  resolved_label_col = label_col,
  resolved_sample_col = sample_col,
  resolved_patient_col = patient_col,
  resolved_condition_col = condition_col,
  resolved_timepoint_col = timepoint_col,
  resolved_batch_col = batch_col,
  resolved_clone_id_col = clone_id_col,
  resolved_clone_size_col = clone_size_col,
  resolved_paired_tcr_col = paired_tcr_col,
  resolved_has_tcr_col = has_tcr_col,
  resolved_consensus_cluster_col = consensus_cluster_col
))

```

# standardize-export
```{r}
#| label: standardize-export
export_cells <- as.data.frame(export_cells_in) %>%
  mutate(
    sample = as.character(.data[[sample_col]]),
    patient = if (!is.null(patient_col)) as.character(.data[[patient_col]]) else NA_character_,
    condition = if (!is.null(condition_col)) as.character(.data[[condition_col]]) else NA_character_,
    timepoint = if (!is.null(timepoint_col)) as.character(.data[[timepoint_col]]) else NA_character_,
    batch = if (!is.null(batch_col)) as.character(.data[[batch_col]]) else NA_character_,
    annot = as.character(.data[[label_col]]),
    clone_id = as.character(.data[[clone_id_col]]),
    clone_size = suppressWarnings(as.numeric(.data[[clone_size_col]])),
    has_tcr = as.logical(.data[[has_tcr_col]]),
    paired_tcr = as.logical(.data[[paired_tcr_col]]),
    consensus_cluster = if (!is.null(consensus_cluster_col)) as.character(.data[[consensus_cluster_col]]) else NA_character_
  ) %>%
  mutate(
    clone_id = ifelse(is.na(clone_id) | clone_id == "", NA_character_, clone_id),
    clone_size = ifelse(is.na(clone_size), 0, clone_size)
  )
```

# Clone frequencies and burden bins
```{r}
#| label: clone-frequency-and-bins
#| results: asis
clone_bin_breaks <- parse_numeric_param_vector(params$clone_bin_breaks)
clone_bin_labels <- parse_character_param_vector(params$clone_bin_labels)

if (length(clone_bin_breaks) < 2) {
  stop("clone_bin_breaks must contain at least two numeric break values.")
}
if (length(clone_bin_labels) != (length(clone_bin_breaks) - 1)) {
  stop("clone_bin_labels must have exactly length(clone_bin_breaks) - 1 entries.")
}

clone_counts_by_sample <- export_cells %>%
  filter(!is.na(clone_id), has_tcr) %>%
  count(sample, clone_id, name = "n_cells")

sample_totals <- export_cells %>%
  filter(has_tcr) %>%
  count(sample, name = "n_tcr_cells")

clone_counts_by_sample <- clone_counts_by_sample %>%
  left_join(sample_totals, by = "sample") %>%
  mutate(
    clone_freq = ifelse(n_tcr_cells > 0, n_cells / n_tcr_cells, NA_real_),
    clone_bin = cut(
      clone_freq,
      breaks = clone_bin_breaks,
      labels = clone_bin_labels,
      include.lowest = TRUE,
      right = TRUE
    )
  )

save_table_safe(clone_counts_by_sample, "clone_counts_by_sample.tsv")

print(
  kableExtra::kbl(
    clone_counts_by_sample %>% slice_head(n = 20),
    caption = "Example clone frequencies and clone-size bins by sample."
  ) %>%
    kableExtra::kable_styling(
      full_width = TRUE,
      bootstrap_options = c("striped", "hover", "condensed", "responsive")
    )
)

```

# Diversity matrics
```{r}
#| label: diversity-metrics
#| results: asis
diversity_tbl <- clone_counts_by_sample %>%
  group_by(sample) %>%
  summarise(
    n_clones = n(),
    shannon = calc_shannon(n_cells),
    simpson = calc_simpson(n_cells),
    inv_simpson = calc_inv_simpson(n_cells),
    richness = calc_richness(n_cells),
    .groups = "drop"
  )

if (any(!is.na(export_cells$patient))) {
  diversity_by_patient <- clone_counts_by_sample %>%
    left_join(export_cells %>% distinct(sample, patient), by = "sample") %>%
    filter(!is.na(patient)) %>%
    group_by(patient, clone_id) %>%
    summarise(n_cells = sum(n_cells), .groups = "drop") %>%
    group_by(patient) %>%
    summarise(
      n_clones = n(),
      shannon = calc_shannon(n_cells),
      simpson = calc_simpson(n_cells),
      inv_simpson = calc_inv_simpson(n_cells),
      richness = calc_richness(n_cells),
      .groups = "drop"
    )
  save_table_safe(diversity_by_patient, "diversity_by_patient.tsv")
}

if (any(!is.na(export_cells$condition))) {
  diversity_by_condition <- clone_counts_by_sample %>%
    left_join(export_cells %>% distinct(sample, condition), by = "sample") %>%
    filter(!is.na(condition)) %>%
    group_by(condition, clone_id) %>%
    summarise(n_cells = sum(n_cells), .groups = "drop") %>%
    group_by(condition) %>%
    summarise(
      n_clones = n(),
      shannon = calc_shannon(n_cells),
      simpson = calc_simpson(n_cells),
      inv_simpson = calc_inv_simpson(n_cells),
      richness = calc_richness(n_cells),
      .groups = "drop"
    )
  save_table_safe(diversity_by_condition, "diversity_by_condition.tsv")
}

save_table_safe(diversity_tbl, "diversity_by_sample.tsv")

print(
  kableExtra::kbl(
    diversity_tbl,
    caption = "Diversity metrics by sample."
  ) %>%
    kableExtra::kable_styling(
      full_width = TRUE,
      bootstrap_options = c("striped", "hover", "condensed", "responsive")
    )
)

```

# Expansion / contraction and clone burden
```{r}
#| label: clone-burden
#| results: asis
clone_burden_tbl <- clone_counts_by_sample %>%
  count(sample, clone_bin, name = "n_clones") %>%
  left_join(
    clone_counts_by_sample %>% count(sample, name = "n_total_clones"),
    by = "sample"
  ) %>%
  mutate(frac_clones = n_clones / n_total_clones)

save_table_safe(clone_burden_tbl, "clone_burden_by_sample.tsv")

clone_burden_cells_tbl <- clone_counts_by_sample %>%
  group_by(sample, clone_bin) %>%
  summarise(
    n_cells = sum(n_cells),
    .groups = "drop"
  ) %>%
  left_join(
    clone_counts_by_sample %>%
      group_by(sample) %>%
      summarise(total_cells = sum(n_cells), .groups = "drop"),
    by = "sample"
  ) %>%
  mutate(frac_cells = n_cells / total_cells)

save_table_safe(clone_burden_cells_tbl, "clone_burden_cells_by_sample.tsv")

print(
  kableExtra::kbl(
    clone_burden_cells_tbl,
    caption = "Clone burden by sample and clone-size bin."
  ) %>%
    kableExtra::kable_styling(
      full_width = TRUE,
      bootstrap_options = c("striped", "hover", "condensed", "responsive")
    )
)

```

# Rank Abundance
```{r}
#| label: rank-abundance
#| results: asis
clone_rank_tbl <- clone_counts_by_sample %>%
  filter(n_cells >= params$min_cells_per_clone_plot) %>%
  group_by(sample) %>%
  arrange(desc(n_cells), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  ungroup()

save_table_safe(clone_rank_tbl, "clone_rank_abundance.tsv")

print(
  kableExtra::kbl(
    clone_rank_tbl %>% slice_head(n = 30),
    caption = "Top ranked clonotypes by sample."
  ) %>%
    kableExtra::kable_styling(
      full_width = TRUE,
      bootstrap_options = c("striped", "hover", "condensed", "responsive")
    )
)

```

# Shareability
```{r}
#| label: shareability
#| results: asis
shareability_tbl <- export_cells %>%
  filter(!is.na(clone_id), has_tcr) %>%
  distinct(sample, patient, condition, clone_id) %>%
  group_by(clone_id) %>%
  summarise(
    n_samples = safe_n_unique(sample),
    n_patients = safe_n_unique(patient),
    n_conditions = safe_n_unique(condition),
    .groups = "drop"
  ) %>%
  mutate(
    shared_across_samples = n_samples >= params$shareability_min_n_groups,
    shared_across_patients = n_patients >= params$shareability_min_n_groups,
    shared_across_conditions = n_conditions >= params$shareability_min_n_groups
  ) %>%
  arrange(desc(n_samples), desc(n_patients), desc(n_conditions))

save_table_safe(shareability_tbl, "clone_shareability.tsv")

top_shared_clones_tbl <- shareability_tbl %>%
  slice_head(n = params$top_n_shared_clones)
save_table_safe(top_shared_clones_tbl, "top_shared_clones.tsv")

print(
  kableExtra::kbl(
    top_shared_clones_tbl,
    caption = "Top shared clonotypes."
  ) %>%
    kableExtra::kable_styling(
      full_width = TRUE,
      bootstrap_options = c("striped", "hover", "condensed", "responsive")
    )
)

```

# Sample Overlap
```{r}
#| label: sample-overlap
#| results: asis
samples_vec <- sort(unique(na.omit(export_cells$sample)))
sample_clone_sets <- lapply(samples_vec, function(s) {
  export_cells %>%
    filter(sample == s, !is.na(clone_id), has_tcr) %>%
    pull(clone_id) %>%
    unique()
})
names(sample_clone_sets) <- samples_vec

overlap_mat <- matrix(NA_real_, nrow = length(samples_vec), ncol = length(samples_vec))
rownames(overlap_mat) <- samples_vec
colnames(overlap_mat) <- samples_vec

for (i in seq_along(samples_vec)) {
  for (j in seq_along(samples_vec)) {
    overlap_mat[i, j] <- calc_overlap_score(
      sample_clone_sets[[i]],
      sample_clone_sets[[j]],
      metric = params$overlap_metric
    )
  }
}

overlap_df <- as.data.frame(overlap_mat) %>%
  tibble::rownames_to_column("sample")

save_table_safe(overlap_df, "sample_overlap_matrix.tsv")

print(
  kableExtra::kbl(
    overlap_df,
    caption = paste("Sample overlap matrix using", params$overlap_metric, "similarity.")
  ) %>%
    kableExtra::kable_styling(
      full_width = TRUE,
      bootstrap_options = c("striped", "hover", "condensed", "responsive")
    )
)

```

# Phenotypic flux
```{r}
#| label: phenotypic-flux
#| results: asis
flux_clone_candidates <- export_cells %>%
  filter(!is.na(clone_id), clone_size >= params$min_cells_per_clone_plot, has_tcr) %>%
  group_by(clone_id) %>%
  summarise(
    n_states = safe_n_unique(annot),
    total_cells = n(),
    .groups = "drop"
  ) %>%
  filter(n_states > 1) %>%
  arrange(desc(total_cells), desc(n_states)) %>%
  slice_head(n = params$top_n_flux_clones)

flux_tbl <- export_cells %>%
  filter(clone_id %in% flux_clone_candidates$clone_id) %>%
  count(clone_id, sample, annot, name = "n_cells")

save_table_safe(flux_tbl, "phenotypic_flux.tsv")

print(
  kableExtra::kbl(
    flux_tbl %>% slice_head(n = 30),
    caption = "Phenotypic flux table for top expanded multi-state clonotypes."
  ) %>%
    kableExtra::kable_styling(
      full_width = TRUE,
      bootstrap_options = c("striped", "hover", "condensed", "responsive")
    )
)

```

# Longitudinal clone tracking
```{r}
#| label: longitudinal-tracking
#| results: asis
longitudinal_tbl <- NULL
if (any(!is.na(export_cells$timepoint))) {
  longitudinal_tbl <- export_cells %>%
    filter(!is.na(clone_id), !is.na(timepoint), has_tcr) %>%
    count(clone_id, sample, patient, timepoint, name = "n_cells") %>%
    group_by(clone_id) %>%
    summarise(
      n_timepoints = safe_n_unique(timepoint),
      total_cells = sum(n_cells),
      .groups = "drop"
    ) %>%
    filter(n_timepoints > 1) %>%
    arrange(desc(total_cells), desc(n_timepoints))

  save_table_safe(longitudinal_tbl, "longitudinal_clone_tracking_summary.tsv")

  print(
    kableExtra::kbl(
      longitudinal_tbl %>% slice_head(n = 30),
      caption = "Longitudinally recurrent clonotypes."
    ) %>%
      kableExtra::kable_styling(
        full_width = TRUE,
        bootstrap_options = c("striped", "hover", "condensed", "responsive")
      )
  )
} else {
  cat("No longitudinal clone tracking table was produced because no timepoint metadata were available.\n")
}

```

# consensus-cluster-burden
```{r}
#| label: consensus-cluster-burden
#| results: asis
consensus_burden_tbl <- NULL
consensus_cluster_summary <- NULL

if (any(!is.na(export_cells$consensus_cluster))) {
  consensus_burden_tbl <- export_cells %>%
    filter(!is.na(consensus_cluster)) %>%
    count(sample, consensus_cluster, name = "n_cells") %>%
    group_by(sample) %>%
    mutate(frac_cells = n_cells / sum(n_cells)) %>%
    ungroup()

  save_table_safe(consensus_burden_tbl, "consensus_cluster_burden.tsv")

  consensus_cluster_summary <- export_cells %>%
    filter(!is.na(consensus_cluster)) %>%
    count(consensus_cluster, name = "n_cells") %>%
    mutate(frac_cells = n_cells / sum(n_cells)) %>%
    arrange(desc(n_cells))

  save_table_safe(consensus_cluster_summary, "consensus_cluster_summary.tsv")

  print(
    kableExtra::kbl(
      consensus_cluster_summary,
      caption = "Consensus clonotype cluster burden summary."
    ) %>%
      kableExtra::kable_styling(
        full_width = TRUE,
        bootstrap_options = c("striped", "hover", "condensed", "responsive")
      )
  )
} else {
  cat("No consensus cluster burden table was produced because no consensus cluster column was available in the current export table.\n")
}

```



# Report overview
```{r}
#| label: overview
#| results: asis
summary_rollup <- tibble(
  metric = c(
    "Report label",
    "Cells in export table",
    "TCR-positive cells",
    "Paired TCR cells",
    "Unique samples",
    "Unique patients",
    "Unique conditions",
    "Unique timepoints",
    "Unique clonotypes",
    "Shared clonotypes across samples"
  ),
  value = c(
    params$report_label,
    nrow(export_cells),
    sum(export_cells$has_tcr, na.rm = TRUE),
    sum(export_cells$paired_tcr, na.rm = TRUE),
    safe_n_unique(export_cells$sample),
    safe_n_unique(export_cells$patient),
    safe_n_unique(export_cells$condition),
    safe_n_unique(export_cells$timepoint),
    safe_n_unique(export_cells$clone_id),
    sum(shareability_tbl$shared_across_samples, na.rm = TRUE)
  )
)

save_table_safe(summary_rollup, "repertoire_summary_rollup.tsv")

print(
  kableExtra::kbl(
    summary_rollup,
    caption = "High-level overview of repertoire analysis."
  ) %>%
    kableExtra::kable_styling(
      full_width = FALSE,
      bootstrap_options = c("striped", "hover", "condensed")
    )
)


```

# Diversity plots
```{r}
#| label: diversity-plots
if (isTRUE(params$show_diversity_boxplots) && nrow(diversity_tbl) > 0) {
  div_long <- diversity_tbl %>%
    pivot_longer(
      cols = c(shannon, simpson, inv_simpson, richness),
      names_to = "metric",
      values_to = "value"
    )

  p_div <- ggplot(div_long, aes(x = metric, y = value, fill = metric)) +
    geom_boxplot(alpha = 0.85, outlier.size = 0.4) +
    geom_jitter(width = 0.12, size = 1.2, alpha = 0.7) +
    guides(fill = "none") +
    labs(
      title = "Repertoire diversity across samples",
      x = NULL,
      y = "Metric value"
    ) +
    theme_scratch_pub(params$base_size)

  print(p_div)
  save_plot_safe(p_div, glue("diversity_boxplots.{params$figure_format}"))
}
```

# Clone burden plots
```{r}
#| label: clone-burden-plots
if (isTRUE(params$show_clone_burden_barplot) && nrow(clone_burden_cells_tbl) > 0) {

  sample_order <- clone_burden_cells_tbl %>%
    mutate(clone_bin = factor(clone_bin, levels = clone_bin_labels)) %>%
    group_by(sample) %>%
    summarise(
      burden_score = sum(frac_cells * as.numeric(clone_bin), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(burden_score)) %>%
    pull(sample)

  p_burden <- clone_burden_cells_tbl %>%
    mutate(
      sample = factor(sample, levels = sample_order),
      clone_bin = factor(clone_bin, levels = clone_bin_labels)
    ) %>%
    ggplot(aes(x = sample, y = n_cells, fill = clone_bin)) +
    geom_col(position = "fill") +
    coord_flip() +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      title = "Clonal burden across samples",
      subtitle = "Samples ordered by overall expansion burden.",
      x = NULL,
      y = "Fraction of cells",
      fill = "Clone bin"
    ) +
    theme_scratch_pub(params$base_size)

  print(p_burden)
  save_plot_safe(p_burden, glue("clone_burden_by_sample.{params$figure_format}"))
}

```

# Rank-abundance
```{r}
#| label: rank-abundance-plot
if (isTRUE(params$show_rank_abundance) && nrow(clone_rank_tbl) > 0) {
  p_rank <- ggplot(clone_rank_tbl, aes(x = rank, y = n_cells, color = sample)) +
    geom_line(linewidth = 0.9, alpha = 0.8) +
    geom_point(size = 1.2, alpha = 0.9) +
    scale_x_log10() +
    scale_y_log10() +
    labs(
      title = "Clone rank-abundance curve",
      subtitle = glue("Only clones with at least {params$min_cells_per_clone_plot} cells are shown."),
      x = "Clone rank (log10)",
      y = "Cells per clone (log10)",
      color = "Sample"
    ) +
    theme_scratch_pub(params$base_size)

  print (p_rank)
  save_plot_safe(p_rank, glue("clone_rank_abundance.{params$figure_format}"))
}
```

# Shareability
```{r}
#| label: shareability-plots
if (isTRUE(params$show_shareability_barplot) && nrow(shareability_tbl) > 0) {
  share_bar_tbl <- tibble(
    category = c("Shared across samples", "Shared across patients", "Shared across conditions"),
    n_clones = c(
      sum(shareability_tbl$shared_across_samples, na.rm = TRUE),
      sum(shareability_tbl$shared_across_patients, na.rm = TRUE),
      sum(shareability_tbl$shared_across_conditions, na.rm = TRUE)
    )
  )

  p_share <- ggplot(share_bar_tbl, aes(x = category, y = n_clones)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    scale_y_continuous(labels = comma) +
    labs(
      title = "Clone shareability summary",
      x = NULL,
      y = "Number of clones"
    ) +
    theme_scratch_pub(params$base_size)

  print (p_share)
  save_plot_safe(p_share, glue("clone_shareability_barplot.{params$figure_format}"))
}
```

# Shareability heatmap
```{r}
#| label: shareability-heatmap
if (isTRUE(params$show_shareability_heatmap) && nrow(top_shared_clones_tbl) > 0) {
  share_heat_tbl <- export_cells %>%
    filter(clone_id %in% top_shared_clones_tbl$clone_id, !is.na(sample), has_tcr) %>%
    count(clone_id, sample, name = "n_cells") %>%
    pivot_wider(names_from = sample, values_from = n_cells, values_fill = 0) %>%
    as.data.frame()

  rownames(share_heat_tbl) <- share_heat_tbl$clone_id
  share_heat_tbl$clone_id <- NULL
  share_mat <- as.matrix(share_heat_tbl)

  max_mat <- max(share_mat, na.rm = TRUE)
  if (!is.finite(max_mat) || max_mat <= 0) max_mat <- 1

  ht <- ComplexHeatmap::Heatmap(
    share_mat,
    name = "Cells",
    col = circlize::colorRamp2(
      c(0, max_mat / 2, max_mat),
      c("white", "gold", "firebrick")
    ),
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    row_names_side = "left",
    row_names_gp = grid::gpar(fontsize = 5),
    column_names_gp = grid::gpar(fontsize = 8),
    column_names_rot = 45,
    row_dend_width = grid::unit(2.5, "cm"),
    column_dend_height = grid::unit(2.0, "cm"),
    rect_gp = grid::gpar(col = "grey85"),
    column_title = "Top shared clonotypes across samples",
    heatmap_legend_param = list(title = "Cells")
  )
  ComplexHeatmap::draw(ht)

  pdf(
    file.path(params$outdir, params$figures_dir, "shareability_heatmap.pdf"),
    width = 10,
    height = 12
  )
  ComplexHeatmap::draw(ht)
  dev.off()
}


```

# sample-overlap-heatmap
```{r}
#| label: sample-overlap-heatmap
if (isTRUE(params$show_sample_overlap_heatmap) && length(samples_vec) > 1) {
  ht <- Heatmap(
    overlap_mat,
    name = params$overlap_metric,
    col = colorRamp2(c(0, 0.5, 1), c("white", "gold", "firebrick")),
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    row_names_side = "left",
    row_names_gp = grid::gpar(fontsize = 9),
    column_names_gp = grid::gpar(fontsize = 9),
    column_names_rot = 45,
    rect_gp = grid::gpar(col = "grey85"),
    column_title = paste("Sample overlap by", params$overlap_metric),
    heatmap_legend_param = list(title = params$overlap_metric)
  )
  draw(ht)
}
```

# phenotypic-flux-plot
# phenotypic-flux-plot
```{r}
# #| label: phenotypic-flux-plot
# if (isTRUE(params$show_phenotypic_flux) && nrow(flux_tbl) > 0) {
#   flux_plot_tbl <- flux_tbl %>%
#     group_by(clone_id) %>%
#     mutate(total_clone_cells = sum(n_cells)) %>%
#     ungroup() %>%
#     arrange(desc(total_clone_cells)) %>%
#     mutate(clone_id = fct_inorder(clone_id))
# 
#   p_flux <- ggplot(
#     flux_plot_tbl,
#     aes(y = n_cells, axis1 = sample, axis2 = annot)
#   ) +
#     geom_alluvium(aes(fill = clone_id), width = 1/12, alpha = 0.85) +
#     geom_stratum(width = 1/12, fill = "grey85", color = "black") +
#     geom_text(
#       stat = "stratum",
#       aes(
#         label = after_stat(stratum),
#         hjust = ifelse(after_stat(x) == 1, 1, 0)
#       ),
#       nudge_x = ifelse(unique(flux_plot_tbl$sample)[1] == unique(flux_plot_tbl$sample)[1], 0, 0),
#       size = 3
#     ) +
#     scale_x_discrete(limits = c("Sample", "Phenotype"), expand = c(.05, .05)) +
#     guides(
#       fill = guide_legend(
#         ncol = 1,
#         byrow = FALSE
#       )
#     ) +
#     labs(
#       title = "Phenotypic flux of expanded clonotypes",
#       x = NULL,
#       y = "Cells",
#       fill = "clone_id"
#     ) +
#     theme_scratch_pub(params$base_size) +
#     theme(
#       legend.position = "right",
#       legend.direction = "vertical",
#       legend.title = element_text(face = "bold"),
#       legend.text = element_text(size = 8),
#       legend.key.height = unit(0.45, "cm"),
#       legend.key.width = unit(0.45, "cm")
#     )
# 
#   print(p_flux)
# 
#   save_plot_safe(
#     p_flux,
#     glue("phenotypic_flux_alluvial.{params$figure_format}"),
#     width = 14,
#     height = 8
#   )
# }



#| label: phenotypic-flux-plot
if (isTRUE(params$show_phenotypic_flux) && nrow(flux_tbl) > 0) {
  flux_plot_tbl <- flux_tbl %>%
    group_by(clone_id) %>%
    mutate(total_clone_cells = sum(n_cells)) %>%
    ungroup() %>%
    arrange(desc(total_clone_cells)) %>%
    mutate(clone_id = fct_inorder(clone_id))

  p_flux <- ggplot(
    flux_plot_tbl,
    aes(y = n_cells, axis1 = sample, axis2 = annot)
  ) +
    geom_alluvium(aes(fill = clone_id), width = 1/12, alpha = 0.85) +
    geom_stratum(width = 1/12, fill = "grey85", color = "black") +
    geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3) +
    scale_x_discrete(limits = c("Sample", "Phenotype"), expand = c(.05, .05)) +
    guides(
      fill = guide_legend(
        ncol = 1,
        byrow = FALSE
      )
    ) +
    labs(
      title = "Phenotypic flux of expanded clonotypes",
      x = NULL,
      y = "Cells",
      fill = "clone_id"
    ) +
    theme_scratch_pub(params$base_size) +
    theme(
      legend.position = "right",
      legend.direction = "vertical",
      legend.title = element_text(face = "bold"),
      legend.text = element_text(size = 6),
      legend.key.height = unit(0.45, "cm"),
      legend.key.width = unit(0.45, "cm")
    )

  print(p_flux)

  save_plot_safe(
    p_flux,
    glue("phenotypic_flux_alluvial.{params$figure_format}"),
    width = 14,
    height = 8
  )
}



```

# Longitudional Clone Tracking
```{r}
#| label: longitudinal-tracking-plot
#| results: asis
if (isTRUE(params$show_longitudinal_clone_tracking) && !is.null(longitudinal_tbl) && nrow(longitudinal_tbl) > 0) {
  top_longitudinal <- export_cells %>%
    filter(!is.na(clone_id), !is.na(timepoint), has_tcr) %>%
    count(clone_id, timepoint, name = "n_cells") %>%
    group_by(clone_id) %>%
    summarise(
      total_cells = sum(n_cells),
      n_timepoints = safe_n_unique(timepoint),
      .groups = "drop"
    ) %>%
    filter(n_timepoints > 1) %>%
    arrange(desc(total_cells)) %>%
    slice_head(n = params$top_n_flux_clones) %>%
    pull(clone_id)

  long_plot_tbl <- export_cells %>%
    filter(clone_id %in% top_longitudinal, !is.na(timepoint), has_tcr) %>%
    count(clone_id, timepoint, name = "n_cells")

  p_long <- ggplot(long_plot_tbl, aes(x = timepoint, y = n_cells, color = clone_id, group = clone_id)) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2) +
    labs(
      title = "Longitudinal tracking of recurrent clonotypes",
      x = "Timepoint",
      y = "Cells",
      color = "Clone"
    ) +
    theme_scratch_pub(params$base_size)

  print(p_long)
  save_plot_safe(p_long, glue("longitudinal_clone_tracking.{params$figure_format}"), width = 11, height = 7)
} else {
  cat("No longitudinal clone tracking plot was produced because no usable timepoint metadata were available.\n")
}

```

# consensus-burden-plot
```{r}
#| label: consensus-burden-plot
#| results: asis
if (isTRUE(params$show_consensus_cluster_burden) && !is.null(consensus_burden_tbl) && nrow(consensus_burden_tbl) > 0) {
  top_n_cons <- params$top_n_clusters
  if (is.null(top_n_cons) || length(top_n_cons) == 0 || is.na(top_n_cons)) {
    top_n_cons <- 20
  }
  top_n_cons <- as.integer(top_n_cons)

  top_cons <- consensus_burden_tbl %>%
    group_by(consensus_cluster) %>%
    summarise(total_cells = sum(n_cells), .groups = "drop") %>%
    arrange(desc(total_cells)) %>%
    slice_head(n = top_n_cons) %>%
    pull(consensus_cluster)

  p_cons <- consensus_burden_tbl %>%
    filter(consensus_cluster %in% top_cons) %>%
    ggplot(aes(x = sample, y = frac_cells, fill = consensus_cluster)) +
    geom_col(position = "fill") +
    coord_flip() +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      title = "Consensus clonotype burden across samples",
      x = NULL,
      y = "Fraction of cells",
      fill = "Consensus cluster"
    ) +
    theme_scratch_pub(params$base_size)

  print(p_cons)
  save_plot_safe(p_cons, glue("consensus_cluster_burden.{params$figure_format}"), width = 11, height = 8)
} else {
  cat("No consensus burden plot was produced because consensus cluster assignments were not available in the current repertoire input.\n")
}



```

# save-repertoire-objects
```{r}
#| label: save-repertoire-objects
save_rds_safe(export_cells, "repertoire_export_cells.rds")

save_rdata_safe(
  export_cells,
  clone_counts_by_sample,
  diversity_tbl,
  clone_burden_tbl,
  clone_burden_cells_tbl,
  clone_rank_tbl,
  shareability_tbl,
  top_shared_clones_tbl,
  overlap_mat,
  flux_tbl,
  longitudinal_tbl,
  consensus_burden_tbl,
  consensus_cluster_summary,
  summary_rollup,
  filename = "repertoire_analysis_objects.RData"
)
```

# Summary tables display
```{r}
#| label: summary-tables-display
#| results: asis
if (isTRUE(params$show_summary_tables)) {

  print(
    kableExtra::kbl(
      diversity_tbl,
      caption = "Diversity metrics by sample."
    ) %>%
      kableExtra::kable_styling(
        full_width = TRUE,
        bootstrap_options = c("striped", "hover", "condensed", "responsive")
      )
  )

  print(
    kableExtra::kbl(
      top_shared_clones_tbl,
      caption = "Top shared clonotypes."
    ) %>%
      kableExtra::kable_styling(
        full_width = TRUE,
        bootstrap_options = c("striped", "hover", "condensed", "responsive")
      )
  )

  if (!is.null(consensus_cluster_summary) && nrow(consensus_cluster_summary) > 0) {
    top_n_cons <- params$top_n_clusters
    if (is.null(top_n_cons) || length(top_n_cons) == 0 || is.na(top_n_cons)) {
      top_n_cons <- 20
    }
    top_n_cons <- as.integer(top_n_cons)

    print(
      kableExtra::kbl(
        consensus_cluster_summary %>% slice_head(n = top_n_cons),
        caption = "Top consensus clonotype clusters by burden."
      ) %>%
        kableExtra::kable_styling(
          full_width = TRUE,
          bootstrap_options = c("striped", "hover", "condensed", "responsive")
        )
    )
  } else {
    cat("Consensus-cluster summary table was skipped because no consensus clusters were available in the current repertoire input.\n")
  }
}

```

# Warnings
```{r}
#| label: warnings
#| results: asis
warn_tbl <- tibble(
  warning = c(
    "Few unique clonotypes",
    "No patient metadata",
    "No condition metadata",
    "No timepoint metadata",
    "No consensus clusters available",
    "Limited shareability across samples"
  ),
  triggered = c(
    safe_n_unique(export_cells$clone_id) < 20,
    all(is.na(export_cells$patient)),
    all(is.na(export_cells$condition)),
    all(is.na(export_cells$timepoint)),
    all(is.na(export_cells$consensus_cluster)),
    sum(shareability_tbl$shared_across_samples, na.rm = TRUE) < 5
  ),
  interpretation = c(
    "Fewer than 20 unique clonotypes were detected, which may limit repertoire-level interpretation.",
    "Patient-level summaries were skipped because patient metadata were unavailable.",
    "Condition-level summaries were skipped because condition metadata were unavailable.",
    "Longitudinal summaries were skipped because timepoint metadata were unavailable.",
    "Consensus-cluster-aware summaries were limited because no consensus clusters were present.",
    "Very few clones were shared across samples under the current threshold."
  )
) %>%
  filter(triggered)

if (nrow(warn_tbl) == 0) {
  cat("No major automatic warnings were triggered under the current repertoire reporting settings.")
} else {
  print(
    kableExtra::kbl(
      warn_tbl %>% select(-triggered),
      caption = "Automatically generated repertoire analysis warnings and notes."
    ) %>%
      kableExtra::kable_styling(
        full_width = TRUE,
        bootstrap_options = c("striped", "hover", "condensed")
      )
  )
}

```

# session info
```{r}
#| label: session-info
writeLines(capture.output(sessionInfo()), file.path(params$outdir, "sessionInfo.repertoire.txt"))
sessionInfo()
```










