---
title: "SCRATCH-TCR: GIANA Analysis"
author: "Syed Shujaat Ali Zaidi"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    theme: cosmo
    df-print: paged
execute:
  echo: false
  warning: false
  message: false

params:
  # ======================================================
  # Inputs
  # ======================================================
  seurat_rds: "data/seurat_tcells_with_TCR.rds"
  tcr_export_cells_file: "tables/tcr_export_cells_with_embedding.tsv"
  outdir: "GIANA_Report"
  giana_script_name: "GIANA4.py"
  # trbv_fasta: "/opt/tools/GIANA/Imgt_Human_TRBV.fasta"
  trbv_fasta: "Imgt_Human_TRBV.fasta"
  

  # Optional precomputed inputs
  giana_clusters_file: ""
  giana_summary_file: ""
  giana_neighbors_file: ""

  # ======================================================
  # Output controls
  # ======================================================
  data_dir: "data"
  tables_dir: "tables"
  figures_dir: "figures"
  save_tables: true
  save_figures: true
  save_updated_seurat: true
  figure_format: "png"
  figure_width: 10
  figure_height: 7
  figure_dpi: 300
  base_size: 12

  # ======================================================
  # Metadata mapping
  # ======================================================
  label_col: ""
  sample_col: "META_SAMPLE"
  patient_col: "META_PATIENT"
  condition_col: "META_TIMECOND"
  timepoint_col: "META_TIMEPOINT"
  batch_col: "META_BATCH"
  clone_id_col: "clone_id"
  clone_id_candidates: "clone_id;CTaa;clonotype;clone"
  
  clone_size_col: "clone_size"
  clone_size_candidates: "clone_size;CloneSize;clone_n"
  
  paired_tcr_col: "paired_tcr"
  paired_tcr_candidates: "paired_tcr;paired;is_paired"
  
  has_tcr_col: "has_tcr"
  has_tcr_candidates: "has_tcr;hasTCR;tcr_positive"

  clone_match_mode: "cdr3b_from_CTaa"
  derive_clone_key_from_CTaa: true

  label_candidates: "predicted_labels;predicted.celltype.l2;predicted.celltype.l1;azimuth_labels;celltypist;celltypist_label;majority_voting;scType;sctype;sctype_label;celltype;CellType;annot;Annotation;seurat_clusters"
  sample_candidates: "META_SAMPLE;orig.ident;sample;sample_id;Sample;SampleID"
  patient_candidates: "META_PATIENT;patient_id;patient;Patient;subject;donor;case_id"
  condition_candidates: "META_TIMECOND;condition;Condition;group;Group;status"
  timepoint_candidates: "META_TIMEPOINT;timepoint;Timepoint;visit;Visit;day;Day;META_TIMECOND"
  batch_candidates: "META_BATCH;batch;Batch;library;Library;run;Run;lane;Lane"
  # clone_id_candidates: "clone_id;CTaa;clonotype;clone"
  # clone_size_candidates: "clone_size;CloneSize;clone_n"
  # paired_tcr_candidates: "paired_tcr;paired;is_paired"
  # has_tcr_candidates: "has_tcr;hasTCR;tcr_positive"

  # ======================================================
  # Embedding controls
  # ======================================================
  reduction_use: "umap"
  make_umap_if_missing: true
  umap_dims_max: 30
  umap_nfeatures: 3000
  raster_large_umap: true

  # ======================================================
  # Analysis controls
  # ======================================================
  min_cells_per_group: 10
  min_cluster_size_plot: 3
  top_n_clusters: 20
  top_n_states_heatmap: 20
  min_cells_total: 100
  min_unique_clones: 25

  show_cluster_umap: true
  show_cluster_size_barplot: true
  show_annotation_heatmap: true
  show_sample_heatmap: true
  show_summary_tables: true
  
  threshold: 3
  threshold_score: 0
  threshold_vgene: 0
  start_position: 3
  num_threads: 1
  use_trbv: true
  exact_mode: false
  
  giana_python_bin: "/opt/conda/envs/tcrenv/bin/python"
  giana_repo_dir: "/opt/tools/GIANA"

  report_label: "GIANA Analysis"
---
# ==============================================================================
# Read the TCR–GEX integrated Seurat object and import GIANA cluster, summary, and optional neighbor tables.
# Resolve flexible clone-matching strategies and map GIANA clone-level annotations back to Seurat cell metadata.
# Perform merge diagnostics to quantify how many cells and clonotypes received GIANA annotations under the chosen match mode.
# Generate publication-quality visualizations of GIANA clusters and scores across cell states, samples, and existing TCR/GEX integration features.
# Export an updated Seurat object plus downstream-ready tables for consensus clonotype clustering and repertoire analysis.
# ==============================================================================

# setup
```{r}
#| label: setup
suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(forcats)
  library(scales)
  library(glue)
  library(knitr)
  library(kableExtra)
  library(ComplexHeatmap)
  library(circlize)
  library(patchwork)
})

options(stringsAsFactors = FALSE)

dir.create(params$outdir, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$data_dir), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$tables_dir), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$figures_dir), recursive = TRUE, showWarnings = FALSE)

`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0 && !all(is.na(a))) a else b

theme_scratch_pub <- function(base_size = 12) {
  theme_bw(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 2, hjust = 0),
      plot.subtitle = element_text(size = base_size, hjust = 0),
      axis.title = element_text(face = "bold"),
      axis.text = element_text(color = "black"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(linewidth = 0.2, color = "grey90"),
      strip.background = element_rect(fill = "grey95", color = "grey80"),
      strip.text = element_text(face = "bold"),
      legend.title = element_text(face = "bold"),
      legend.key = element_blank(),
      plot.caption = element_text(size = base_size - 2, color = "grey40")
    )
}

save_plot_safe <- function(plot_obj, filename,
                           width = params$figure_width,
                           height = params$figure_height,
                           dpi = params$figure_dpi) {
  if (!isTRUE(params$save_figures)) return(invisible(NULL))
  ggsave(
    filename = file.path(params$outdir, params$figures_dir, filename),
    plot = plot_obj,
    width = width,
    height = height,
    dpi = dpi,
    bg = "white",
    limitsize = FALSE
  )
}

save_table_safe <- function(df, filename) {
  if (!isTRUE(params$save_tables)) return(invisible(NULL))
  fwrite(df, file.path(params$outdir, params$tables_dir, filename), sep = "\t")
}

save_rds_safe <- function(obj, filename) {
  if (!isTRUE(params$save_updated_seurat)) return(invisible(NULL))
  saveRDS(obj, file.path(params$outdir, params$data_dir, filename))
}

require_file <- function(path, msg = NULL) {
  if (!file.exists(path)) stop(msg %||% paste("Missing required file:", path))
  path
}

first_existing_col <- function(df, preferred, candidates = character()) {
  if (!is.null(preferred) && preferred != "" && preferred %in% colnames(df)) return(preferred)
  hits <- intersect(candidates, colnames(df))
  if (length(hits) > 0) return(hits[[1]])
  NULL
}

safe_make_umap <- function(seu, reduction_name = "umap", dims = 1:30, nfeatures = 3000, seed = 1234) {
  if ("RNA" %in% names(seu@assays)) DefaultAssay(seu) <- "RNA"
  npcs <- min(max(dims), max(2, ncol(seu) - 1), 50)
  nf <- min(nfeatures, nrow(seu))

  if (!"pca" %in% Reductions(seu)) {
    seu <- FindVariableFeatures(seu, nfeatures = nf, verbose = FALSE)
    seu <- ScaleData(seu, verbose = FALSE)
    seu <- RunPCA(seu, npcs = npcs, verbose = FALSE)
  }

  use_dims <- dims[dims <= npcs]
  if (length(use_dims) < 2) use_dims <- 1:min(10, npcs)

  seu <- FindNeighbors(seu, dims = use_dims, verbose = FALSE)
  seu <- RunUMAP(seu, dims = use_dims, reduction.name = reduction_name, verbose = FALSE)
  seu
}

extract_cdr3b_from_CTaa <- function(x) {
  x <- as.character(x)
  out <- rep(NA_character_, length(x))
  has_b <- grepl("\\|B:", x)
  out[has_b] <- sub("^.*\\|B:", "", x[has_b])
  only_b <- grepl("^B:", x)
  out[only_b] <- sub("^B:", "", x[only_b])
  out
}

sanitize_param_string <- function(x) {
  if (is.null(x) || length(x) == 0) return(NULL)
  trimws(gsub("\\u00A0", " ", as.character(x)))
}

normalize_colnames <- function(df) {
  colnames(df) <- trimws(gsub("\\u00A0", " ", colnames(df)))
  df
}

split_candidate_string <- function(x) {
  if (is.null(x) || length(x) == 0 || is.na(x) || x == "") return(character())
  x <- gsub('"', "", as.character(x))
  x <- trimws(unlist(strsplit(x, ";", fixed = TRUE)))
  x[nzchar(x)]
}

first_existing_col <- function(df, preferred, candidates = character()) {
  df <- normalize_colnames(df)
  preferred <- sanitize_param_string(preferred)

  if (length(candidates) == 1 && is.character(candidates)) {
    candidates <- split_candidate_string(candidates)
  }
  candidates <- trimws(gsub("\\u00A0", " ", as.character(candidates)))

  if (!is.null(preferred) && preferred != "" && preferred %in% colnames(df)) return(preferred)

  hits <- intersect(candidates, colnames(df))
  if (length(hits) > 0) return(hits[[1]])
  NULL
}

```

# read-inputs
```{r}
#| label: read-inputs
require_file(params$seurat_rds, sprintf("Seurat RDS not found: %s", params$seurat_rds))

giana_python_bin <- sanitize_param_string(params$giana_python_bin)
giana_repo_dir <- sanitize_param_string(params$giana_repo_dir)

if (is.null(giana_python_bin) || giana_python_bin == "") {
  giana_python_bin <- "/opt/conda/envs/tcrenv/bin/python"
}
if (is.null(giana_repo_dir) || giana_repo_dir == "") {
  giana_repo_dir <- "/opt/tools/GIANA"
}

seu <- readRDS(params$seurat_rds)

message("Resolved GIANA python bin: ", giana_python_bin)
message("Resolved GIANA repo dir: ", giana_repo_dir)
```

# resolve-columns
```{r}
#| label: resolve-columns
md <- normalize_colnames(seu@meta.data)
seu@meta.data <- md

label_col <- first_existing_col(md, params$label_col, params$label_candidates)
sample_col <- first_existing_col(md, params$sample_col, params$sample_candidates)
patient_col <- first_existing_col(md, params$patient_col, params$patient_candidates)
condition_col <- first_existing_col(md, params$condition_col, params$condition_candidates)
timepoint_col <- first_existing_col(md, params$timepoint_col, params$timepoint_candidates)
batch_col <- first_existing_col(md, params$batch_col, params$batch_candidates)

clone_id_col <- first_existing_col(md, params$clone_id_col, params$clone_id_candidates)
clone_size_col <- first_existing_col(md, params$clone_size_col, params$clone_size_candidates)
paired_tcr_col <- first_existing_col(md, params$paired_tcr_col, params$paired_tcr_candidates)
has_tcr_col <- first_existing_col(md, params$has_tcr_col, params$has_tcr_candidates)

tcri_score_col_md <- first_existing_col(md, params$tcri_score_col, params$tcri_score_candidates)
tcri_group_col_md <- first_existing_col(md, params$tcri_group_col, params$tcri_group_candidates)
conga_cluster_col_md <- first_existing_col(md, params$conga_cluster_col, params$conga_cluster_candidates_md)
conga_score_col_md <- first_existing_col(md, params$conga_score_col, params$conga_score_candidates_md)
gliph_cluster_col_md <- first_existing_col(md, params$gliph_cluster_col, params$gliph_cluster_candidates_md)
gliph_pattern_col_md <- first_existing_col(md, params$gliph_pattern_col, params$gliph_pattern_candidates_md)
tcrdist_cluster_col_md <- first_existing_col(md, params$tcrdist_cluster_col, params$tcrdist_cluster_candidates_md)
tcrdist_score_col_md <- first_existing_col(md, params$tcrdist_score_col, params$tcrdist_score_candidates_md)

if (is.null(label_col)) stop("Could not resolve an annotation label column in Seurat metadata.")
if (is.null(sample_col)) stop("Could not resolve sample column in Seurat metadata.")
if (is.null(patient_col)) stop("Could not resolve patient column in Seurat metadata.")
if (is.null(condition_col)) stop("Could not resolve condition column in Seurat metadata.")
if (is.null(timepoint_col)) stop("Could not resolve timepoint column in Seurat metadata.")
if (is.null(clone_id_col)) stop("Could not resolve a clone ID column in Seurat metadata.")
if (is.null(clone_size_col)) stop("Could not resolve a clone size column in Seurat metadata.")
if (is.null(has_tcr_col)) stop("Could not resolve a has_tcr column in Seurat metadata.")
if (is.null(paired_tcr_col)) stop("Could not resolve a paired_tcr column in Seurat metadata.")

if (!"META_SAMPLE" %in% colnames(seu@meta.data) && !is.null(sample_col)) {
  seu$META_SAMPLE <- as.character(seu@meta.data[[sample_col]])
}
if (!"META_PATIENT" %in% colnames(seu@meta.data) && !is.null(patient_col)) {
  seu$META_PATIENT <- as.character(seu@meta.data[[patient_col]])
}
if (!"META_TIMECOND" %in% colnames(seu@meta.data) && !is.null(condition_col)) {
  seu$META_TIMECOND <- as.character(seu@meta.data[[condition_col]])
}
if (!"META_TIMEPOINT" %in% colnames(seu@meta.data) && !is.null(timepoint_col)) {
  seu$META_TIMEPOINT <- as.character(seu@meta.data[[timepoint_col]])
}
if (!"META_BATCH" %in% colnames(seu@meta.data) && !is.null(batch_col)) {
  seu$META_BATCH <- as.character(seu@meta.data[[batch_col]])
}

print(list(
  resolved_label_col = label_col,
  resolved_sample_col = sample_col,
  resolved_patient_col = patient_col,
  resolved_condition_col = condition_col,
  resolved_timepoint_col = timepoint_col,
  resolved_batch_col = batch_col,
  resolved_clone_id_col = clone_id_col,
  resolved_clone_size_col = clone_size_col,
  resolved_paired_tcr_col = paired_tcr_col,
  resolved_has_tcr_col = has_tcr_col,
  resolved_tcri_score_col = tcri_score_col_md,
  resolved_tcri_group_col = tcri_group_col_md,
  resolved_conga_cluster_col = conga_cluster_col_md,
  resolved_conga_score_col = conga_score_col_md,
  resolved_gliph_cluster_col = gliph_cluster_col_md,
  resolved_gliph_pattern_col = gliph_pattern_col_md,
  resolved_tcrdist_cluster_col = tcrdist_cluster_col_md,
  resolved_tcrdist_score_col = tcrdist_score_col_md
))
```

# subset-for-giana
```{r}
#| label: subset-for-giana
seu_g <- seu

if (isTRUE(params$subset_to_tcr_positive)) {
  keep <- !is.na(seu_g@meta.data[[has_tcr_col]]) & as.logical(seu_g@meta.data[[has_tcr_col]])
  seu_g <- subset(seu_g, cells = rownames(seu_g@meta.data)[keep])
}

if (isTRUE(params$subset_to_paired_tcr)) {
  keep <- !is.na(seu_g@meta.data[[paired_tcr_col]]) & as.logical(seu_g@meta.data[[paired_tcr_col]])
  seu_g <- subset(seu_g, cells = rownames(seu_g@meta.data)[keep])
}

if (ncol(seu_g) < params$min_cells_total) {
  stop("Too few cells remain for GIANA analysis after filtering: ", ncol(seu_g))
}
```

# build-giana-input
```{r}
#| label: build-giana-input
seu_md <- seu_g@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  mutate(
    clone_id_md = as.character(.data[[clone_id_col]]),
    CDR3 = if (isTRUE(params$derive_clone_key_from_CTaa)) extract_cdr3b_from_CTaa(clone_id_md) else clone_id_md
  ) %>%
  filter(!is.na(CDR3), CDR3 != "")

v_col <- intersect(c("trbv", "v_b_gene", "vb_gene", "TRBV"), colnames(seu_md))
seu_md$TRBV <- if (length(v_col) > 0) as.character(seu_md[[v_col[1]]]) else NA_character_

giana_input <- seu_md %>%
  group_by(clone_id_md, CDR3) %>%
  summarise(
    TRBV = first(na.omit(TRBV)) %||% NA_character_,
    count = n(),
    .groups = "drop"
  ) %>%
  mutate(clone_id = clone_id_md) %>%
  select(CDR3, TRBV, clone_id, count)

if (!isTRUE(params$use_trbv) || all(is.na(giana_input$TRBV))) {
  giana_input <- giana_input %>% select(CDR3, clone_id, count)
}

if (nrow(giana_input) < params$min_unique_clones) {
  stop("Too few unique clonotypes for GIANA: ", nrow(giana_input))
}

input_file <- file.path(params$outdir, params$tables_dir, "giana_input.tsv")
save_table_safe(giana_input, "giana_input.tsv")
```

# run-giana
```{r}
#| label: run-giana
giana_script_name <- sanitize_param_string(params$giana_script_name)
if (is.null(giana_script_name) || giana_script_name == "") {
  giana_script_name <- "GIANA4.py"
}

giana_script <- file.path(giana_repo_dir, giana_script_name)
if (!file.exists(giana_script)) {
  stop("Could not find GIANA script at: ", giana_script)
}

trbv_fasta <- sanitize_param_string(params$trbv_fasta)
if (is.null(trbv_fasta) || trbv_fasta == "") {
  trbv_fasta <- "Imgt_Human_TRBV.fasta"
}

# GIANA expects the FASTA path relative to its own repo dir
trbv_fasta_basename <- basename(trbv_fasta)
trbv_fasta_check <- file.path(giana_repo_dir, trbv_fasta_basename)
if (!file.exists(trbv_fasta_check)) {
  stop("Could not find TRBV fasta file at: ", trbv_fasta_check)
}

out_dir <- file.path(params$outdir, params$data_dir, "giana_raw")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

threshold <- params$threshold
if (is.null(threshold) || length(threshold) == 0 || is.na(threshold)) {
  threshold <- 3
}

threshold_score <- params$threshold_score
if (is.null(threshold_score) || length(threshold_score) == 0 || is.na(threshold_score)) {
  threshold_score <- 0
}

threshold_vgene <- params$threshold_vgene
if (is.null(threshold_vgene) || length(threshold_vgene) == 0 || is.na(threshold_vgene)) {
  threshold_vgene <- 0
}

start_position <- params$start_position
if (is.null(start_position) || length(start_position) == 0 || is.na(start_position)) {
  start_position <- 3
}

num_threads <- params$num_threads
if (is.null(num_threads) || length(num_threads) == 0 || is.na(num_threads)) {
  num_threads <- 1
}

args <- c(
  giana_script,
  "-f", normalizePath(input_file, mustWork = TRUE),
  "-o", normalizePath(out_dir, mustWork = TRUE),
  "-t", as.character(threshold),
  "-S", as.character(threshold_score),
  "-G", as.character(threshold_vgene),
  "-T", as.character(start_position),
  "-N", as.character(num_threads)
)

if (!isTRUE(params$use_trbv) || ncol(giana_input) < 4) {
  args <- c(args, "-v")
}

if (!isTRUE(params$exact_mode)) {
  args <- c(args, "-e")
}

if (!is.null(trbv_fasta_basename) && trbv_fasta_basename != "") {
  args <- c(args, "-V", trbv_fasta_basename)
}

message("Resolved GIANA script: ", giana_script)
message("Resolved TRBV fasta basename: ", trbv_fasta_basename)
message("Resolved TRBV fasta check path: ", trbv_fasta_check)
message("Running GIANA with args: ", paste(shQuote(args), collapse = " "))

giana_log <- system2(
  giana_python_bin,
  args = args,
  stdout = TRUE,
  stderr = TRUE
)

giana_log_file <- file.path(params$outdir, params$tables_dir, "giana_run.log.txt")
writeLines(giana_log, giana_log_file)

status <- attr(giana_log, "status")
if (is.null(status)) status <- 0L
if (status != 0L) {
  stop(
    paste0(
      "GIANA failed with exit status ", status, ". ",
      "See log: ", giana_log_file
    )
  )
}

cand <- list.files(out_dir, pattern = "\\.(txt|tsv)$", full.names = TRUE)
if (length(cand) == 0) {
  stop("GIANA completed but no output table was found in: ", out_dir)
}

cand_info <- file.info(cand)
giana_output_file <- rownames(cand_info)[which.max(cand_info$mtime)]

message("Resolved GIANA output file: ", giana_output_file)

# #| label: run-giana
# giana_script_name <- sanitize_param_string(params$giana_script_name)
# if (is.null(giana_script_name) || giana_script_name == "") {
#   giana_script_name <- "GIANA4.py"
# }
# 
# giana_script <- file.path(giana_repo_dir, giana_script_name)
# if (!file.exists(giana_script)) {
#   stop("Could not find GIANA script at: ", giana_script)
# }
# 
# trbv_fasta <- sanitize_param_string(params$trbv_fasta)
# if (is.null(trbv_fasta) || trbv_fasta == "") {
#   trbv_fasta <- file.path(giana_repo_dir, "Imgt_Human_TRBV.fasta")
# }
# 
# out_dir <- file.path(params$outdir, params$data_dir, "giana_raw")
# dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
# 
# threshold <- params$threshold
# if (is.null(threshold) || length(threshold) == 0 || is.na(threshold)) {
#   threshold <- 3
# }
# 
# threshold_score <- params$threshold_score
# if (is.null(threshold_score) || length(threshold_score) == 0 || is.na(threshold_score)) {
#   threshold_score <- 0
# }
# 
# threshold_vgene <- params$threshold_vgene
# if (is.null(threshold_vgene) || length(threshold_vgene) == 0 || is.na(threshold_vgene)) {
#   threshold_vgene <- 0
# }
# 
# start_position <- params$start_position
# if (is.null(start_position) || length(start_position) == 0 || is.na(start_position)) {
#   start_position <- 3
# }
# 
# num_threads <- params$num_threads
# if (is.null(num_threads) || length(num_threads) == 0 || is.na(num_threads)) {
#   num_threads <- 1
# }
# 
# args <- c(
#   giana_script,
#   "-f", normalizePath(input_file, mustWork = TRUE),
#   "-o", normalizePath(out_dir, mustWork = TRUE),
#   "-t", as.character(threshold),
#   "-S", as.character(threshold_score),
#   "-G", as.character(threshold_vgene),
#   "-T", as.character(start_position),
#   "-N", as.character(num_threads)
# )
# 
# if (!isTRUE(params$use_trbv) || ncol(giana_input) < 4) {
#   args <- c(args, "-v")
# }
# 
# if (!isTRUE(params$exact_mode)) {
#   args <- c(args, "-e")
# }
# 
# if (!is.null(trbv_fasta) && trbv_fasta != "") {
#   if (!file.exists(trbv_fasta)) {
#     stop("Could not find TRBV fasta file at: ", trbv_fasta)
#   }
#   args <- c(args, "-V", normalizePath(trbv_fasta, mustWork = TRUE))
# }
# 
# message("Resolved GIANA script: ", giana_script)
# message("Resolved TRBV fasta: ", trbv_fasta)
# message("Running GIANA with args: ", paste(shQuote(args), collapse = " "))
# 
# giana_log <- system2(
#   giana_python_bin,
#   args = args,
#   stdout = TRUE,
#   stderr = TRUE
# )
# 
# giana_log_file <- file.path(params$outdir, params$tables_dir, "giana_run.log.txt")
# writeLines(giana_log, giana_log_file)
# 
# status <- attr(giana_log, "status")
# if (is.null(status)) status <- 0L
# if (status != 0L) {
#   stop(
#     paste0(
#       "GIANA failed with exit status ", status, ". ",
#       "See log: ", giana_log_file
#     )
#   )
# }
# 
# # GIANA writes a default output name derived from input. Find the newest txt/tsv in out_dir.
# cand <- list.files(out_dir, pattern = "\\.(txt|tsv)$", full.names = TRUE)
# if (length(cand) == 0) {
#   stop("GIANA completed but no output table was found in: ", out_dir)
# }
# 
# cand_info <- file.info(cand)
# giana_output_file <- rownames(cand_info)[which.max(cand_info$mtime)]
# 
# message("Resolved GIANA output file: ", giana_output_file)

```

# read-giana-output
```{r}
#| label: read-giana-output
# GIANA output inserts the cluster-id column between the first and second input columns.
giana_raw <- fread(giana_output_file, header = FALSE, sep = "\t", data.table = FALSE)

if (ncol(giana_raw) < 3) stop("Unexpected GIANA output shape.")

if (isTRUE(params$use_trbv) && ncol(giana_input) >= 4) {
  colnames(giana_raw)[1:4] <- c("CDR3", "giana_cluster", "TRBV", "clone_id")
  if (ncol(giana_raw) >= 5) colnames(giana_raw)[5] <- "count"
} else {
  colnames(giana_raw)[1:3] <- c("CDR3", "giana_cluster", "clone_id")
  if (ncol(giana_raw) >= 4) colnames(giana_raw)[4] <- "count"
}

giana_tbl_std <- giana_raw %>%
  mutate(
    giana_clone_key = as.character(CDR3),
    giana_cluster = as.character(giana_cluster),
    giana_group = as.character(giana_cluster),
    giana_score = if ("count" %in% colnames(.)) suppressWarnings(as.numeric(count)) else NA_real_,
    giana_rank = NA_real_,
    giana_pvalue = NA_real_,
    giana_qvalue = NA_real_,
    giana_cluster_size = as.numeric(ave(seq_along(giana_cluster), giana_cluster, FUN = length)),
    giana_distance = NA_real_
  )

save_table_safe(giana_tbl_std, "giana_clusters.tsv")

# simple neighbor summary within cluster
giana_neighbors_tbl <- giana_tbl_std %>%
  select(giana_cluster, clone_id) %>%
  inner_join(., ., by = "giana_cluster", suffix = c("_source", "_target")) %>%
  filter(clone_id_source != clone_id_target) %>%
  transmute(source = clone_id_source, target = clone_id_target, giana_cluster = giana_cluster) %>%
  distinct()

save_table_safe(giana_neighbors_tbl, "giana_neighbors.tsv")
```

# merge-giana-into-seurat
```{r}
#| label: merge-giana-into-seurat
seu_md_full <- seu@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  mutate(
    clone_id_md = as.character(.data[[clone_id_col]]),
    clone_match_key_cdr3b = if (isTRUE(params$derive_clone_key_from_CTaa)) extract_cdr3b_from_CTaa(clone_id_md) else clone_id_md
  )

merge_df <- seu_md_full %>%
  left_join(
    giana_tbl_std %>%
      select(giana_clone_key, giana_cluster, giana_group, giana_score, giana_rank, giana_pvalue, giana_qvalue, giana_cluster_size, giana_distance) %>%
      distinct(giana_clone_key, .keep_all = TRUE),
    by = c("clone_match_key_cdr3b" = "giana_clone_key")
  )

rownames(merge_df) <- merge_df$cell_id
merge_df$cell_id <- NULL
seu@meta.data <- as.data.frame(merge_df)

save_rds_safe(seu, "seurat_with_GIANA.rds")
```

# ensure-embedding
```{r}
#| label: ensure-embedding
reduction_to_use <- params$reduction_use
if (!(reduction_to_use %in% names(seu@reductions)) && isTRUE(params$make_umap_if_missing)) {
  seu <- safe_make_umap(
    seu,
    reduction_name = params$reduction_use,
    dims = 1:params$umap_dims_max,
    nfeatures = params$umap_nfeatures
  )
}
if (!(reduction_to_use %in% names(seu@reductions))) {
  reduction_to_use <- if ("umap" %in% names(seu@reductions)) "umap" else if ("tsne" %in% names(seu@reductions)) "tsne" else if ("pca" %in% names(seu@reductions)) "pca" else stop("No usable embedding found.")
}
```

# build-export
```{r}
#| label: build-export
embed_df <- as.data.frame(Embeddings(seu, reduction_to_use)) %>%
  tibble::rownames_to_column("cell_id")

export_cells <- seu@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  transmute(
    cell_id,
    sample = if (!is.null(sample_col)) .data[[sample_col]] else NA_character_,
    patient = if (!is.null(patient_col)) .data[[patient_col]] else NA_character_,
    condition = if (!is.null(condition_col)) .data[[condition_col]] else NA_character_,
    timepoint = if (!is.null(timepoint_col)) .data[[timepoint_col]] else NA_character_,
    batch = if (!is.null(batch_col)) .data[[batch_col]] else NA_character_,
    annot = .data[[label_col]],
    clone_id = .data[[clone_id_col]],
    clone_size = suppressWarnings(as.numeric(.data[[clone_size_col]])),
    has_tcr = as.logical(.data[[has_tcr_col]]),
    paired_tcr = as.logical(.data[[paired_tcr_col]]),
    tcri_score = if (!is.null(tcri_score_col_md)) suppressWarnings(as.numeric(.data[[tcri_score_col_md]])) else NA_real_,
    tcri_group = if (!is.null(tcri_group_col_md)) as.character(.data[[tcri_group_col_md]]) else NA_character_,
    conga_cluster = if (!is.null(conga_cluster_col_md)) as.character(.data[[conga_cluster_col_md]]) else NA_character_,
    conga_score = if (!is.null(conga_score_col_md)) suppressWarnings(as.numeric(.data[[conga_score_col_md]])) else NA_real_,
    gliph_cluster = if (!is.null(gliph_cluster_col_md)) as.character(.data[[gliph_cluster_col_md]]) else NA_character_,
    gliph_pattern = if (!is.null(gliph_pattern_col_md)) as.character(.data[[gliph_pattern_col_md]]) else NA_character_,
    tcrdist_cluster = if (!is.null(tcrdist_cluster_col_md)) as.character(.data[[tcrdist_cluster_col_md]]) else NA_character_,
    tcrdist_score = if (!is.null(tcrdist_score_col_md)) suppressWarnings(as.numeric(.data[[tcrdist_score_col_md]])) else NA_real_,
    giana_cluster = as.character(.data[["giana_cluster"]]),
    giana_group = as.character(.data[["giana_group"]]),
    giana_score = suppressWarnings(as.numeric(.data[["giana_score"]])),
    giana_rank = suppressWarnings(as.numeric(.data[["giana_rank"]])),
    giana_pvalue = suppressWarnings(as.numeric(.data[["giana_pvalue"]])),
    giana_qvalue = suppressWarnings(as.numeric(.data[["giana_qvalue"]])),
    giana_cluster_size = suppressWarnings(as.numeric(.data[["giana_cluster_size"]])),
    giana_distance = suppressWarnings(as.numeric(.data[["giana_distance"]]))
  ) %>%
  left_join(embed_df, by = "cell_id")

save_table_safe(export_cells, "giana_export_cells.tsv")
```

# merge-diagnostics
```{r}
#| label: merge-diagnostics
merge_diagnostics <- tibble(
  metric = c(
    "Cells in Seurat",
    "Rows in GIANA results",
    "Cells with GIANA cluster",
    "Cells with GIANA score",
    "Fraction with GIANA cluster"
  ),
  value = c(
    ncol(seu),
    nrow(giana_tbl_std),
    sum(!is.na(export_cells$giana_cluster)),
    sum(!is.na(export_cells$giana_score)),
    sprintf("%.2f%%", 100 * mean(!is.na(export_cells$giana_cluster)))
  )
)
save_table_safe(merge_diagnostics, "giana_merge_diagnostics.tsv")
```

# summary-tables
```{r}
#| label: summary-tables
safe_cor_summary <- function(df, xcol, ycol) {
  if (!(xcol %in% colnames(df)) || !(ycol %in% colnames(df))) return(NULL)

  tmp <- df %>%
    filter(!is.na(.data[[xcol]]), !is.na(.data[[ycol]]))

  if (nrow(tmp) <= 1) return(NULL)

  tibble(
    n_cells = nrow(tmp),
    pearson_cor = cor(tmp[[xcol]], tmp[[ycol]], use = "pairwise.complete.obs", method = "pearson"),
    spearman_cor = cor(tmp[[xcol]], tmp[[ycol]], use = "pairwise.complete.obs", method = "spearman")
  )
}

safe_overlap_summary <- function(df, xcol, ycol) {
  if (!(xcol %in% colnames(df)) || !(ycol %in% colnames(df))) return(NULL)

  tmp <- df %>%
    filter(!is.na(.data[[xcol]]), !is.na(.data[[ycol]]))

  if (nrow(tmp) == 0) return(NULL)

  tibble(
    n_cells = nrow(tmp),
    n_same_nonmissing = sum(tmp[[xcol]] == tmp[[ycol]], na.rm = TRUE)
  )
}

summary_rollup <- tibble(
  metric = c(
    "Report label",
    "Cells in Seurat",
    "Cells with GIANA clusters",
    "Cells with GIANA scores",
    "TCR-positive cells",
    "Paired TCR cells",
    "Unique GIANA clusters",
    "Unique annotations",
    "Unique samples",
    "Selected embedding"
  ),
  value = c(
    params$report_label,
    nrow(export_cells),
    sum(!is.na(export_cells$giana_cluster)),
    sum(!is.na(export_cells$giana_score)),
    sum(export_cells$has_tcr, na.rm = TRUE),
    sum(export_cells$paired_tcr, na.rm = TRUE),
    length(unique(na.omit(export_cells$giana_cluster))),
    length(unique(na.omit(export_cells$annot))),
    length(unique(na.omit(export_cells$sample))),
    reduction_to_use
  )
)
save_table_safe(summary_rollup, "giana_summary_rollup.tsv")

cluster_summary <- export_cells %>%
  filter(!is.na(giana_cluster)) %>%
  count(giana_cluster, name = "n_cells") %>%
  mutate(frac = n_cells / sum(n_cells)) %>%
  arrange(desc(n_cells))
save_table_safe(cluster_summary, "giana_cluster_summary.tsv")

annotation_giana_summary <- export_cells %>%
  filter(!is.na(annot)) %>%
  group_by(annot) %>%
  summarise(
    n_cells = n(),
    n_giana = sum(!is.na(giana_cluster)),
    frac_giana = n_giana / n_cells,
    mean_giana = mean(giana_score, na.rm = TRUE),
    median_giana = median(giana_score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(frac_giana))
save_table_safe(annotation_giana_summary, "annotation_giana_summary.tsv")

sample_giana_summary <- export_cells %>%
  filter(!is.na(sample)) %>%
  group_by(sample) %>%
  summarise(
    n_cells = n(),
    n_giana = sum(!is.na(giana_cluster)),
    frac_giana = n_giana / n_cells,
    mean_giana = mean(giana_score, na.rm = TRUE),
    median_giana = median(giana_score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(frac_giana))
save_table_safe(sample_giana_summary, "sample_giana_summary.tsv")

if ("condition" %in% colnames(export_cells) && any(!is.na(export_cells$condition))) {
  condition_giana_summary <- export_cells %>%
    filter(!is.na(condition)) %>%
    group_by(condition) %>%
    summarise(
      n_cells = n(),
      n_giana = sum(!is.na(giana_cluster)),
      frac_giana = n_giana / n_cells,
      mean_giana = mean(giana_score, na.rm = TRUE),
      median_giana = median(giana_score, na.rm = TRUE),
      .groups = "drop"
    )
  save_table_safe(condition_giana_summary, "condition_giana_summary.tsv")
}

if ("patient" %in% colnames(export_cells) && any(!is.na(export_cells$patient))) {
  patient_giana_summary <- export_cells %>%
    filter(!is.na(patient)) %>%
    group_by(patient) %>%
    summarise(
      n_cells = n(),
      n_giana = sum(!is.na(giana_cluster)),
      frac_giana = n_giana / n_cells,
      mean_giana = mean(giana_score, na.rm = TRUE),
      median_giana = median(giana_score, na.rm = TRUE),
      .groups = "drop"
    )
  save_table_safe(patient_giana_summary, "patient_giana_summary.tsv")
}

cluster_annotation_tbl <- export_cells %>%
  filter(!is.na(giana_cluster), !is.na(annot)) %>%
  count(giana_cluster, annot, name = "n_cells")
save_table_safe(cluster_annotation_tbl, "giana_cluster_annotation_composition.tsv")

cluster_sample_tbl <- export_cells %>%
  filter(!is.na(giana_cluster), !is.na(sample)) %>%
  count(giana_cluster, sample, name = "n_cells")
save_table_safe(cluster_sample_tbl, "giana_cluster_sample_composition.tsv")

tcri_giana_assoc_tbl <- safe_cor_summary(export_cells, "tcri_score", "giana_score")
if (!is.null(tcri_giana_assoc_tbl)) {
  save_table_safe(tcri_giana_assoc_tbl, "tcri_giana_association.tsv")
}

conga_giana_assoc_tbl <- safe_cor_summary(export_cells, "conga_score", "giana_score")
if (!is.null(conga_giana_assoc_tbl)) {
  save_table_safe(conga_giana_assoc_tbl, "conga_giana_association.tsv")
}

gliph_giana_assoc_tbl <- safe_overlap_summary(export_cells, "gliph_cluster", "giana_cluster")
if (!is.null(gliph_giana_assoc_tbl)) {
  save_table_safe(gliph_giana_assoc_tbl, "gliph_giana_overlap_summary.tsv")
}

tcrdist_giana_assoc_tbl <- safe_overlap_summary(export_cells, "tcrdist_cluster", "giana_cluster")
if (!is.null(tcrdist_giana_assoc_tbl)) {
  save_table_safe(tcrdist_giana_assoc_tbl, "tcrdist_giana_overlap_summary.tsv")
}

```

# overview
```{r}
#| label: overview
kable(summary_rollup, caption = "High-level overview of GIANA analysis.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))
```

# score-density
```{r}
#| label: score-density
if (isTRUE(params$show_score_density) && any(!is.na(export_cells$giana_score))) {
  p_density <- export_cells %>%
    filter(!is.na(giana_score)) %>%
    ggplot(aes(x = giana_score)) +
    geom_density(fill = "steelblue", alpha = 0.5, linewidth = 0.8) +
    labs(
      title = "Distribution of GIANA scores",
      subtitle = "Mapped clone-level GIANA support.",
      x = "GIANA score",
      y = "Density"
    ) +
    theme_scratch_pub(params$base_size)

  print(p_density)
  save_plot_safe(p_density, glue("giana_score_density.{params$figure_format}"))
} else {
  cat("No GIANA score-density plot was produced because no non-missing GIANA scores were available.\n")
}
```

# cluster-umap
```{r}
#| label: cluster-umap
#| fig-width: 18
#| fig-height: 12
#| warning: false

if (isTRUE(params$show_cluster_umap)) {
  
  # 1. Prepare Data
  # Map GLIPH clusters to Seurat metadata
  seu$gliph_display <- export_cells$gliph_cluster[match(colnames(seu), export_cells$cell_id)]
  cells_with_gliph <- colnames(seu)[!is.na(seu$gliph_display)]
  
  if (length(cells_with_gliph) > 0) {
    
    # ADVANCED VIZ: High-contrast base plot
    # We use 'na.value = "grey95"' to make non-GLIPH cells fade into the background
    p_base <- DimPlot(
      seu, 
      reduction = reduction_to_use, 
      group.by = "gliph_display", 
      label = TRUE, 
      label.size = 4,
      repel = TRUE,
      raster = isTRUE(params$raster_large_umap),
      pt.size = 0.5
    ) + 
      scale_color_viridis_d(option = "turbo", na.value = "grey95") +
      labs(title = "GLIPH2 Cluster Territorial Distribution",
           subtitle = "Colored by GLIPH2 specificity group") +
      theme_scratch_pub(params$base_size) +
      theme(legend.position = "none") # Label clusters directly on plot to save space

    # ADVANCED VIZ: Density Hotspots
    # This shows where TCR convergence is 'dense' in the transcriptional landscape
    coords <- as.data.frame(Embeddings(seu, reduction_to_use)[cells_with_gliph, ])
    colnames(coords) <- c("dim1", "dim2")

    p_final <- p_base + 
      stat_density_2d(
        data = coords, 
        aes(x = dim1, y = dim2), 
        color = "black", 
        linewidth = 0.2, 
        alpha = 0.4
      ) +
      labs(caption = "Contours indicate high-density regions of GLIPH2-assigned clones.")

    # 2. Rendering & Saving
    # We explicitly check the viewport by using a large height in ggsave
    print(p_final)
    
    save_plot_safe(
      p_final, 
      glue("gliph2_advanced_hotspot_umap.{params$figure_format}"),
      width = 18, 
      height = 12
    )

  } else {
    cat("No GLIPH2 clusters found to visualize on UMAP.")
  }
}


```

# cluster-size-barplot
```{r}
#| label: cluster-size-barplot
if (isTRUE(params$show_cluster_size_barplot) && nrow(cluster_summary) > 0) {
  top_n_clusters_display <- params$top_n_clusters
  if (is.null(top_n_clusters_display) || length(top_n_clusters_display) == 0 || is.na(top_n_clusters_display)) {
    top_n_clusters_display <- 20
  }
  top_n_clusters_display <- as.integer(top_n_clusters_display)

  top_clusters <- cluster_summary %>% slice_head(n = top_n_clusters_display)

  p_cluster_bar <- top_clusters %>%
    mutate(giana_cluster = fct_reorder(as.character(giana_cluster), n_cells)) %>%
    ggplot(aes(x = giana_cluster, y = n_cells)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    scale_y_continuous(labels = comma) +
    labs(title = "Top GIANA clusters", x = NULL, y = "Cells") +
    theme_scratch_pub(params$base_size)

  print(p_cluster_bar)
  save_plot_safe(p_cluster_bar, glue("giana_cluster_size_barplot.{params$figure_format}"))
} else {
  cat("No GIANA cluster-size barplot was produced because no GIANA clusters were available.\n")
}

```

# annotation-heatmap
```{r}
#| label: annotation-heatmap
#| fig-width: 14
#| fig-height: 10

if (isTRUE(params$show_cluster_annotation_heatmap) && nrow(cluster_annotation_tbl) > 0) {
  
  # 1. Prepare Matrix (using the as.numeric fix for sprintf)
  top_clusters_vec <- cluster_summary %>%
    filter(n_cells >= params$min_cluster_size_plot) %>%
    slice_head(n = params$top_n_clusters) %>%
    pull(giana_cluster)

  mat_df <- cluster_annotation_tbl %>%
    filter(giana_cluster %in% top_clusters_vec) %>%
    pivot_wider(names_from = annot, values_from = n_cells, values_fill = 0) %>%
    as.data.frame()

  rownames(mat_df) <- paste0("C", sprintf("%02d", as.numeric(as.character(mat_df$giana_cluster))))
  mat_df$giana_cluster <- NULL
  mat <- as.matrix(mat_df)

  # 2. Advanced: Create Top Annotation (Condition/Group)
  # Summarize the majority condition per cluster
  cond_map <- export_cells %>%
    filter(giana_cluster %in% top_clusters_vec) %>%
    group_by(giana_cluster) %>%
    summarise(top_cond = names(which.max(table(condition))), .groups = "drop") %>%
    mutate(cid = paste0("C", sprintf("%02d", as.numeric(as.character(giana_cluster)))))
  
  # Match annotation to matrix rows
  anno_vec <- cond_map$top_cond[match(rownames(mat), cond_map$cid)]
  
  row_ha = rowAnnotation(
    Majority_Condition = anno_vec,
    col = list(Majority_Condition = c("Pre" = "#66c2a5", "Post" = "#fc8d62", "Normal" = "#8da0cb")), # Custom hex codes
    gp = gpar(col = "white")
  )

  # 3. Draw Advanced Heatmap
  ht <- Heatmap(
    mat,
    name = "Cell Count",
    right_annotation = row_ha,
    col = colorRamp2(c(0, max(mat)/2, max(mat)), c("#f7fbff", "#6baed6", "#08306b")),
    cluster_rows = TRUE, 
    cluster_columns = TRUE,
    column_title = "Cluster-Annotation Mapping with Clinical Metadata",
    row_names_gp = gpar(fontsize = 10, fontface = "bold"),
    column_names_gp = gpar(fontsize = 10),
    rect_gp = gpar(col = "white", lwd = 1)
  )

  draw(ht, heatmap_legend_side = "right")
}

```

# sample-heatmap
```{r}
#| label: sample-heatmap
if (isTRUE(params$show_cluster_sample_heatmap) && nrow(cluster_sample_tbl) > 0) {
  top_n_clusters_display <- params$top_n_clusters
  if (is.null(top_n_clusters_display) || length(top_n_clusters_display) == 0 || is.na(top_n_clusters_display)) {
    top_n_clusters_display <- 20
  }
  top_n_clusters_display <- as.integer(top_n_clusters_display)

  top_clusters <- cluster_summary %>%
    filter(n_cells >= params$min_cluster_size_plot) %>%
    slice_head(n = top_n_clusters_display) %>%
    pull(giana_cluster)

  mat_df <- cluster_sample_tbl %>%
    filter(giana_cluster %in% top_clusters) %>%
    pivot_wider(names_from = sample, values_from = n_cells, values_fill = 0) %>%
    as.data.frame()

  if (nrow(mat_df) > 0 && ncol(mat_df) > 1) {
    rownames(mat_df) <- as.character(mat_df$giana_cluster)
    mat_df$giana_cluster <- NULL
    mat <- as.matrix(mat_df)

    max_mat <- max(mat, na.rm = TRUE)
    if (!is.finite(max_mat) || max_mat <= 0) max_mat <- 1

    ht <- Heatmap(
      mat,
      name = "Cells",
      col = colorRamp2(
        c(0, max_mat / 2, max_mat),
        c("white", "skyblue", "navy")
      ),
      cluster_rows = TRUE,
      cluster_columns = TRUE,
      row_names_side = "left",
      row_names_gp = grid::gpar(fontsize = 8),
      column_names_gp = grid::gpar(fontsize = 9),
      column_names_rot = 45,
      column_title = "Sample composition of GIANA clusters",
      heatmap_legend_param = list(title = "Cells")
    )

    ComplexHeatmap::draw(ht, heatmap_legend_side = "right")
  } else {
    cat("No GIANA sample heatmap was produced because no cluster/sample matrix could be constructed.\n")
  }
} else {
  cat("No GIANA sample heatmap was produced because no non-missing GIANA cluster/sample pairs were available.\n")
}
```

# cluster-composition-plot
```{r}
#| label: cluster-composition-advanced
#| fig-width: 16
#| fig-height: 10

if (nrow(cluster_sample_tbl) > 0) {
  # Normalize counts to proportions within each cluster
  comp_df <- cluster_sample_tbl %>%
    group_by(giana_cluster) %>%
    mutate(prop = n_cells / sum(n_cells)) %>%
    ungroup() %>%
    filter(giana_cluster %in% top_clusters)

  p_comp <- ggplot(comp_df, aes(x = giana_cluster, y = prop, fill = sample)) +
    geom_col(position = "fill", color = "white", linewidth = 0.1) +
    scale_y_continuous(labels = percent_format()) +
    labs(
      title = "GIANA Cluster Composition by Sample",
      subtitle = "Are clusters dominated by a single sample (private) or shared (public)?",
      x = "GIANA Cluster",
      y = "Proportion of Cells",
      fill = "Sample"
    ) +
    theme_scratch_pub(params$base_size) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right",
      legend.text = element_text(size = params$base_size - 4)
    )

  print(p_comp)
  save_plot_safe(p_comp, "giana_cluster_composition_normalized.png")
}
```

# fraction-plots
```{r}
#| label: fraction-plots
plot_frac_by_group <- function(df, group_col, title_txt, filename) {
  tmp <- df %>%
    filter(!is.na(.data[[group_col]])) %>%
    group_by(.data[[group_col]]) %>%
    summarise(
      n_cells = n(),
      n_giana = sum(!is.na(giana_cluster)),
      frac_giana = n_giana / n_cells,
      .groups = "drop"
    ) %>%
    filter(n_cells >= params$min_cells_per_group)

  if (nrow(tmp) == 0) return(NULL)

  p <- tmp %>%
    mutate(.group = fct_reorder(as.character(.data[[group_col]]), frac_giana)) %>%
    ggplot(aes(x = .group, y = frac_giana)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(title = title_txt, x = NULL, y = "Cells with GIANA annotation") +
    theme_scratch_pub(params$base_size)

  print(p)
  save_plot_safe(p, filename)
  invisible(p)
}

any_fraction_plot <- FALSE

if (isTRUE(params$show_giana_by_annotation)) {
  p1 <- plot_frac_by_group(
    export_cells %>% rename(group_annot = annot),
    "group_annot",
    "Fraction of cells with GIANA annotation by cell state",
    glue("giana_fraction_by_annotation.{params$figure_format}")
  )
  any_fraction_plot <- any_fraction_plot || !is.null(p1)
}

if (isTRUE(params$show_giana_by_sample) && any(!is.na(export_cells$sample))) {
  p2 <- plot_frac_by_group(
    export_cells,
    "sample",
    "Fraction of cells with GIANA annotation by sample",
    glue("giana_fraction_by_sample.{params$figure_format}")
  )
  any_fraction_plot <- any_fraction_plot || !is.null(p2)
}

if (isTRUE(params$show_giana_by_condition) && any(!is.na(export_cells$condition))) {
  p3 <- plot_frac_by_group(
    export_cells,
    "condition",
    "Fraction of cells with GIANA annotation by condition",
    glue("giana_fraction_by_condition.{params$figure_format}")
  )
  any_fraction_plot <- any_fraction_plot || !is.null(p3)
}

if (isTRUE(params$show_giana_by_patient) && any(!is.na(export_cells$patient))) {
  p4 <- plot_frac_by_group(
    export_cells,
    "patient",
    "Fraction of cells with GIANA annotation by patient",
    glue("giana_fraction_by_patient.{params$figure_format}")
  )
  any_fraction_plot <- any_fraction_plot || !is.null(p4)
}

if (!any_fraction_plot) {
  cat("No GIANA fraction plots were produced because no grouping variable had enough non-missing cells above the minimum threshold.\n")
}
```

# giana-nj-tree
```{r}
#| label: giana-nj-tree
#| fig-width: 14
#| fig-height: 12

if (isTRUE(params$show_giana_tree) && exists("Mat")) {
  
  # 1. Select the top clusters for visualization (similar to your heatmap)
  # Visualizing 1000s of tips is messy, so we focus on the most expanded/convergent ones
  target_clusters <- cluster_summary %>%
    slice_head(n = 30) %>%
    pull(giana_cluster)
  
  # Filter data for the tree
  tree_df <- export_cells %>%
    filter(giana_cluster %in% target_clusters) %>%
    distinct(clone_id, .keep_all = TRUE) # One tip per clone
  
  # 2. Match with GIANA's Isometric Matrix (Assuming 'Mat' was loaded in read-inputs)
  # We use the Pearson distance formula: $D = \sqrt{1 - r}$
  subset_mat <- Mat[tree_df$clone_id, ]
  subset_mat <- na.omit(subset_mat)
  
  if (nrow(subset_mat) > 5) {
    dist_mat <- as.dist(sqrt(1 - cor(t(subset_mat))))
    
    # 3. Generate the Neighbor Joining Tree
    nj_tree <- ape::nj(dist_mat)
    
    # 4. Advanced Visualization
    # Color tips by 'Condition' or 'Sample' to see if clusters are clinical-specific
    tip_colors <- circlize::colorRamp2(
      c(1, length(unique(tree_df$sample))), 
      c("dodgerblue", "firebrick")
    )(as.numeric(factor(tree_df$sample)))

    # Use ape's plotting but wrap in a PDF/PNG for the report
    plot_path <- file.path(params$outdir, params$figures_dir, glue("giana_nj_tree.{params$figure_format}"))
    
    # Render to device
    png(plot_path, width = 14, height = 12, units = "in", res = 300)
    par(mar = c(1, 1, 3, 10))
    ape::plot.phylo(
      nj_tree, 
      type = "phylogram", 
      tip.color = tip_colors,
      label.offset = 0.02,
      cex = 0.8,
      font = 2,
      main = "Structural Similarity of Top GIANA Clusters"
    )
    ape::add.scale.bar()
    dev.off()
    
    # Display in Quarto
    ape::plot.phylo(nj_tree, type = "phylogram", tip.color = tip_colors, cex = 0.8)
    
  } else {
    cat("Not enough unique clones in top clusters to build a meaningful tree.\n")
  }
}
```

# association-plots
```{r}
#| label: association-plots
any_assoc_plot <- FALSE

if (isTRUE(params$show_tcri_vs_giana) &&
    "tcri_score" %in% colnames(export_cells) &&
    any(!is.na(export_cells$tcri_score)) &&
    any(!is.na(export_cells$giana_score))) {

  assoc_df <- export_cells %>% filter(!is.na(tcri_score), !is.na(giana_score))

  if (nrow(assoc_df) > 1) {
    p_assoc <- assoc_df %>%
      ggplot(aes(x = tcri_score, y = giana_score)) +
      geom_point(alpha = 0.35, size = 1.1) +
      geom_smooth(method = "lm", se = TRUE, color = "firebrick") +
      labs(title = "Association between TCRi and GIANA scores", x = "TCRi score", y = "GIANA score") +
      theme_scratch_pub(params$base_size)

    print(p_assoc)
    save_plot_safe(p_assoc, glue("tcri_vs_giana.{params$figure_format}"))
    any_assoc_plot <- TRUE
  }
}

if (isTRUE(params$show_conga_vs_giana) &&
    "conga_score" %in% colnames(export_cells) &&
    any(!is.na(export_cells$conga_score)) &&
    any(!is.na(export_cells$giana_score))) {

  assoc_df2 <- export_cells %>% filter(!is.na(conga_score), !is.na(giana_score))

  if (nrow(assoc_df2) > 1) {
    p_assoc2 <- assoc_df2 %>%
      ggplot(aes(x = conga_score, y = giana_score)) +
      geom_point(alpha = 0.35, size = 1.1) +
      geom_smooth(method = "lm", se = TRUE, color = "firebrick") +
      labs(title = "Association between CoNGA and GIANA scores", x = "CoNGA score", y = "GIANA score") +
      theme_scratch_pub(params$base_size)

    print(p_assoc2)
    save_plot_safe(p_assoc2, glue("conga_vs_giana.{params$figure_format}"))
    any_assoc_plot <- TRUE
  }
}

if (isTRUE(params$show_clone_size_vs_giana) &&
    any(!is.na(export_cells$clone_size)) &&
    any(!is.na(export_cells$giana_score))) {

  assoc_df3 <- export_cells %>% filter(!is.na(clone_size), clone_size > 0, !is.na(giana_score))

  if (nrow(assoc_df3) > 1) {
    p_clone_assoc <- assoc_df3 %>%
      ggplot(aes(x = clone_size, y = giana_score)) +
      geom_point(alpha = 0.35, size = 1.1) +
      scale_x_log10() +
      geom_smooth(method = "lm", se = TRUE, color = "firebrick") +
      labs(title = "Association between clone size and GIANA score", x = "Clone size (log10 scale)", y = "GIANA score") +
      theme_scratch_pub(params$base_size)

    print(p_clone_assoc)
    save_plot_safe(p_clone_assoc, glue("clone_size_vs_giana.{params$figure_format}"))
    any_assoc_plot <- TRUE
  }
}

if (!any_assoc_plot) {
  cat("No GIANA association plots were produced because there were not enough non-missing paired values for the requested comparisons.\n")
}
```

# neighbor-summary
```{r}
#| label: neighbor-summary
if (isTRUE(params$show_summary_tables) && nrow(giana_neighbors_tbl) > 0) {
  neighbor_summary <- tibble(
    metric = c("Neighbor rows loaded", "Unique source entries", "Unique target entries"),
    value = c(
      nrow(giana_neighbors_tbl),
      length(unique(giana_neighbors_tbl$source)),
      length(unique(giana_neighbors_tbl$target))
    )
  )
  save_table_safe(neighbor_summary, "giana_neighbor_summary.tsv")

  knitr::asis_output(
    kableExtra::kbl(
      neighbor_summary,
      caption = "GIANA neighbor summary."
    ) %>%
      kableExtra::kable_styling(
        full_width = FALSE,
        bootstrap_options = c("striped", "hover", "condensed")
      )
  )
} else {
  cat("No GIANA neighbor summary table was produced because no GIANA neighbor table was available.\n")
}

```

# summary-tables-display
```{r}
#| label: summary-tables-display
#| results: asis

if (isTRUE(params$show_summary_tables)) {

  top_n_clusters_display <- as.integer(params$top_n_clusters %||% 20)
  top_n_states_display <- as.integer(params$top_n_states_heatmap %||% 20)

  # Function to generate styled kables safely
  render_styled_table <- function(df, cap) {
    ktab <- kableExtra::kbl(df, caption = cap) %>%
      kableExtra::kable_styling(
        full_width = FALSE,
        bootstrap_options = c("striped", "hover", "condensed", "responsive")
      )
    return(ktab)
  }

  # Table 1: Top Clusters
  cat("### Top GIANA Clusters\n")
  print(render_styled_table(
    cluster_summary %>% slice_head(n = top_n_clusters_display),
    "Top GIANA clusters by cell count."
  ))
  cat("\n\n<br>\n\n")

  # Table 2: Annotation Rankings
  cat("### Annotation Specificity\n")
  print(render_styled_table(
    annotation_giana_summary %>% slice_head(n = top_n_states_display),
    "Top annotations ranked by GIANA-assigned fraction."
  ))
  cat("\n\n<br>\n\n")

  # Table 3: Diagnostics
  cat("### Merge Diagnostics\n")
  print(render_styled_table(
    merge_diagnostics,
    "GIANA merge diagnostics."
  ))
}

```

# warnings
```{r}
#| label: warnings
patient_missing <- !("patient" %in% colnames(export_cells)) || all(is.na(export_cells$patient))
condition_missing <- !("condition" %in% colnames(export_cells)) || all(is.na(export_cells$condition))

warn_tbl <- tibble(
  warning = c(
    "Few cells with GIANA clusters",
    "Few cells with GIANA scores",
    "No condition metadata in exported table",
    "No patient metadata in exported table"
  ),
  triggered = c(
    mean(!is.na(export_cells$giana_cluster)) < 0.3,
    mean(!is.na(export_cells$giana_score)) < 0.3,
    condition_missing,
    patient_missing
  ),
  interpretation = c(
    "Less than 30% of cells in the Seurat object received a GIANA cluster annotation.",
    "Less than 30% of cells received a GIANA score.",
    "Condition-level summaries were skipped because exported condition metadata were empty after GIANA export.",
    "Patient-level summaries were skipped because exported patient metadata were empty after GIANA export."
  )
) %>%
  filter(triggered)

if (nrow(warn_tbl) == 0) {
  cat("No major automatic warnings were triggered under the current GIANA settings.")
} else {
  knitr::asis_output(
    kableExtra::kbl(
      warn_tbl %>% select(-triggered),
      caption = "Automatically generated GIANA warnings and notes."
    ) %>%
      kableExtra::kable_styling(
        full_width = TRUE,
        bootstrap_options = c("striped", "hover", "condensed")
      )
  )
}

```

# session-info
```{r}
#| label: session-info
writeLines(capture.output(sessionInfo()), file.path(params$outdir, "sessionInfo.giana.txt"))
sessionInfo()
```

