---
title: "SCRATCH-TCR: VDJ QC Analysis"
author: "SCRATCH-TCR"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    theme: cosmo
    df-print: paged
execute:
  echo: false
  warning: false
  message: false
params:
  # ======================================================
  # Inputs
  # ======================================================
  sample_sheet: "vdj_samples.txt"
  metadata_file: ""
  outdir: "VDJ_QC"
  input_annotated_object: ""

  # ======================================================
  # Sample sheet columns
  # sample sheet must minimally contain:
  #   sample, path
  # ======================================================
  sample_sheet_sample_col: "sample"
  sample_sheet_path_col: "path"

  # ======================================================
  # Metadata mapping (optional)
  # ======================================================
  meta_barcode_col: "barcode"
  meta_sample_col: "sample"
  meta_patient_col: "patient_id"
  meta_condition_col: "condition"
  meta_timepoint_col: "timepoint"

  meta_barcode_candidates: !expr c("barcode","cell_id","cell","CB")
  meta_sample_candidates: !expr c("sample","sample_id","orig.ident","library_id")
  meta_patient_candidates: !expr c("patient_id","patient","subject_id")
  meta_condition_candidates: !expr c("condition","group","status","DiseaseStatus")
  meta_timepoint_candidates: !expr c("timepoint","time","visit","day")

  # ======================================================
  # QC parameters
  # ======================================================
  chain_mode: "auto"            # auto / T-AB / T-GD / both
  require_productive: true
  require_high_conf: true
  require_full_length: true
  min_umis: 1
  min_reads: 0
  keep_one_per_chain: true
  keep_paired_only: false
  keep_dual_alpha: true
  keep_dual_beta: true
  cdr3_aa_min: 8
  cdr3_aa_max: 25
  drop_stop_or_frames: true
  top_n_samples_table: 50

  # ======================================================
  # Plot/report settings
  # ======================================================
  figures_dir: "figures"
  tables_dir: "tables"
  save_tables: true
  save_figures: true
  figure_format: "png"
  figure_width: 10
  figure_height: 7
  figure_dpi: 300
  base_size: 12

  top_n_v: 20
  top_n_j: 20
  min_clone_size_plot: 1

  show_reads_umis: true
  show_patient_panels: true
  show_condition_panels: true
  show_timepoint_panels: true
  show_clone_rank_plot: true
  show_vj_heatmaps: true

  report_label: "VDJ QC Analysis"
---
# =========================================================
# SCRATCH-TCR | VDJ QC Analysis
# 1) Read a multi-sample VDJ sample sheet and load Cell Ranger VDJ outputs.
# 2) Standardize contig fields, optionally merge metadata, and infer/filter TCR chain mode.
# 3) Apply explicit QC filters (productive/high-confidence/full-length, reads/UMIs, CDR3 validity/length, one-per-chain, optional paired-only).
# 4) Generate derived QC summaries at contig, cell, and sample levels, including retention, pairing, multi-chain burden, V/J usage, and clone rank-abundance.
# 5) Render a publication-quality QC report and save reusable tables/figures for downstream SCRATCH-TCR modules.

# VDJ_QC_Report.qmd ✔
# TCell_Integration_Report.qmd ✔
# TCRi_Report.qmd ✔patient resolve done
# CoNGA_Report.qmd ✔ patiend resolve done
# GLIPH2_Report.qmd ✔ patient col resolve done
# TCRdist3_Report.qmd  ✔ 
# GIANA_Report.qmd  ✔ 
# Clonotype_Clustering_Consensus_Report.qmd ✔  
# Repertoire_Report.qmd  ✔ 
# Master_Summary_Report.qmd 


# ****IMPORTANT****
# Ensure patient_id and timepoint exist as NA at minimum
# =========================================================



##SETUP
```{r}
#| label: setup
suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(forcats)
  library(ggplot2)
  library(scales)
  library(purrr)
  library(glue)
  library(knitr)
  library(kableExtra)
  library(ComplexHeatmap)
  library(circlize)
  library(patchwork)
  library(RColorBrewer)
})

options(stringsAsFactors = FALSE)

dir.create(params$outdir, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$figures_dir), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$tables_dir), recursive = TRUE, showWarnings = FALSE)

`%||%` <- function(x, y) if (!is.null(x) && length(x) > 0 && !all(is.na(x))) x else y

# theme_scratch_pub <- function(base_size = 12) {
#   theme_bw(base_size = base_size) +
#     theme(
#       plot.title = element_text(face = "bold", size = base_size + 2, hjust = 0),
#       plot.subtitle = element_text(size = base_size, hjust = 0),
#       axis.title = element_text(face = "bold"),
#       axis.text = element_text(color = "black"),
#       panel.grid.minor = element_blank(),
#       panel.grid.major = element_line(linewidth = 0.2, color = "grey90"),
#       strip.background = element_rect(fill = "grey95", color = "grey80"),
#       strip.text = element_text(face = "bold"),
#       legend.title = element_text(face = "bold"),
#       legend.key = element_blank(),
#       plot.caption = element_text(size = base_size - 2, color = "grey40")
#     )
# }
# theme_scratch_pub <- function(base_size = 12) {
#   theme_bw(base_size = base_size) +
#     theme(
#       plot.title = element_text(face = "bold", size = base_size + 2, hjust = 0),
#       plot.subtitle = element_text(size = base_size, hjust = 0),
#       axis.title = element_text(face = "bold", size = base_size - 1),
#       # Shrink Y-axis labels slightly
#       axis.text.y = element_text(color = "black", size = base_size - 3), 
#       # Rotate X-axis labels 45 degrees and shrink
#       axis.text.x = element_text(color = "black", size = base_size - 3, angle = 45, hjust = 1), 
#       panel.grid.minor = element_blank(),
#       panel.grid.major = element_line(linewidth = 0.1, color = "grey90"),
#       strip.background = element_rect(fill = "grey95", color = "grey80"),
#       strip.text = element_text(face = "bold", size = base_size - 2),
#       legend.title = element_text(face = "bold", size = base_size - 2),
#       legend.text = element_text(size = base_size - 3),
#       plot.caption = element_text(size = base_size - 4, color = "grey40")
#     )
# }
theme_scratch_pub <- function(base_size = 12) {
  theme_bw(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 2, hjust = 0),
      plot.subtitle = element_text(size = base_size, hjust = 0),
      axis.title = element_text(face = "bold", size = base_size - 1),
      axis.text.y = element_text(color = "black", size = base_size - 3),
      # Fix for Q2 & Q3: Smaller text and 45-degree rotation for X-axis
      axis.text.x = element_text(color = "black", size = base_size - 3, angle = 45, hjust = 1),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(linewidth = 0.1, color = "grey90"),
      strip.background = element_rect(fill = "grey95", color = "grey80"),
      strip.text = element_text(face = "bold", size = base_size - 2),
      legend.title = element_text(face = "bold", size = base_size - 2),
      legend.text = element_text(size = base_size - 3),
      legend.key = element_blank(),
      plot.caption = element_text(size = base_size - 4, color = "grey40")
    )
}

save_plot_safe <- function(plot_obj, filename,
                           width = params$figure_width,
                           height = params$figure_height,
                           dpi = params$figure_dpi) {
  if (!isTRUE(params$save_figures)) return(invisible(NULL))
  ggsave(
    filename = file.path(params$outdir, params$figures_dir, filename),
    plot = plot_obj,
    width = width,
    height = height,
    dpi = dpi,
    bg = "white",
    limitsize = FALSE
  )
}

save_table_safe <- function(df, filename) {
  if (!isTRUE(params$save_tables)) return(invisible(NULL))
  fwrite(df, file.path(params$outdir, params$tables_dir, filename), sep = "\t")
}

safe_read_table <- function(path) {
  if (is.null(path) || is.na(path) || path == "" || !file.exists(path)) return(NULL)
  ext <- tolower(tools::file_ext(path))
  if (ext %in% c("tsv", "tab", "txt")) {
    fread(path, sep = "\t")
  } else {
    fread(path)
  }
}

first_existing_col <- function(df, preferred, candidates = character()) {
  if (!is.null(preferred) && preferred %in% colnames(df)) return(preferred)
  hits <- intersect(candidates, colnames(df))
  if (length(hits) > 0) return(hits[[1]])
  return(NULL)
}

require_cols <- function(df, cols, object_name = deparse(substitute(df))) {
  miss <- setdiff(cols, colnames(df))
  if (length(miss) > 0) {
    stop(glue("{object_name} is missing required columns: {paste(miss, collapse = ', ')}"))
  }
}

has_col <- function(df, col) {
  !is.null(col) && col %in% colnames(df)
}

to_logical <- function(v) {
  if (is.logical(v)) return(v)
  if (is.numeric(v)) return(v != 0)
  if (is.character(v)) return(tolower(trimws(v)) %in% c("true", "t", "yes", "y", "1"))
  return(as.logical(v))
}

numify <- function(x) suppressWarnings(as.numeric(x))

clean_chain <- function(x) {
  x <- as.character(x)
  x <- toupper(trimws(x))
  x
}

infer_chain_mode <- function(chain_vals) {
  chain_vals <- unique(na.omit(chain_vals))
  has_ab <- any(chain_vals %in% c("TRA","TRB"))
  has_gd <- any(chain_vals %in% c("TRG","TRD"))
  if (has_ab && !has_gd) return("T-AB")
  if (!has_ab && has_gd) return("T-GD")
  if (has_ab && has_gd) return("both")
  "unknown"
}

filter_chain_mode <- function(df, chain_col, mode = "auto") {
  if (!has_col(df, chain_col)) return(df)
  mode_use <- mode
  if (mode_use == "auto") mode_use <- infer_chain_mode(df[[chain_col]])
  if (mode_use == "T-AB") return(df %>% filter(.data[[chain_col]] %in% c("TRA","TRB")))
  if (mode_use == "T-GD") return(df %>% filter(.data[[chain_col]] %in% c("TRG","TRD")))
  df
}

safe_pct <- function(x, denom) ifelse(denom == 0, NA_real_, x / denom)

coalesce_chr <- function(x, y) ifelse(is.na(x) | x == "", y, x)

detect_vdj_file <- function(outs_dir, fname) {
  roots <- c(
    outs_dir,
    file.path(outs_dir, "vdj_t"),
    file.path(outs_dir, "vdj_b")
  )
  fp <- file.path(roots, fname)
  i <- which(file.exists(fp))[1]
  if (is.na(i)) return(NULL)
  fp[[i]]
}

read_samplesheet <- function(path, sample_col = "sample", path_col = "path") {
  stopifnot(file.exists(path))
  df <- safe_read_table(path) %>% as.data.frame()
  require_cols(df, c(sample_col, path_col), "sample_sheet")
  stopifnot(!anyDuplicated(df[[sample_col]]))

  df <- df %>%
    mutate(
      sample = as.character(.data[[sample_col]]),
      raw_path = as.character(.data[[path_col]])
    )

  df$outs <- vapply(df$raw_path, function(p) {
    p <- normalizePath(p, mustWork = FALSE)
    if (dir.exists(file.path(p, "outs"))) {
      file.path(p, "outs")
    } else if (grepl("/outs/?$", p) && dir.exists(p)) {
      p
    } else {
      stop("Could not find 'outs' directory for path: ", p)
    }
  }, character(1))

  df %>% select(sample, outs) %>% distinct()
}

load_vdj_one <- function(sname, outs_dir) {
  contigs_fp <- detect_vdj_file(outs_dir, "filtered_contig_annotations.csv")
  clonotypes_fp <- detect_vdj_file(outs_dir, "clonotypes.csv")
  metrics_fp <- detect_vdj_file(outs_dir, "metrics_summary.csv")

  if (is.null(contigs_fp)) {
    stop("[", sname, "] missing filtered_contig_annotations.csv under ", outs_dir)
  }
  if (is.null(clonotypes_fp)) {
    stop("[", sname, "] missing clonotypes.csv under ", outs_dir)
  }

  contigs <- fread(contigs_fp) %>% as.data.frame()
  clonotypes <- fread(clonotypes_fp) %>% as.data.frame()
  metrics <- NULL
  if (!is.null(metrics_fp) && file.exists(metrics_fp)) {
    metrics <- tryCatch(fread(metrics_fp) %>% as.data.frame(), error = function(e) NULL)
  }

  contigs$sample <- sname
  clonotypes$sample <- sname
  if (!is.null(metrics)) metrics$sample <- sname

  list(
    contigs = contigs,
    clonotypes = clonotypes,
    metrics = metrics
  )
}
```

#read input
```{r}
#| label: read-inputs

sample_sheet <- read_samplesheet(
  params$sample_sheet,
  sample_col = params$sample_sheet_sample_col,
  path_col = params$sample_sheet_path_col
)

metadata_tbl <- safe_read_table(params$metadata_file)

if ((is.null(metadata_tbl) || nrow(metadata_tbl) == 0) &&
    !is.null(params$input_annotated_object) &&
    params$input_annotated_object != "" &&
    file.exists(params$input_annotated_object)) {

  suppressPackageStartupMessages({
    library(Seurat)
    library(tibble)
    library(dplyr)
  })

  seu <- readRDS(params$input_annotated_object)

  metadata_tbl <- seu@meta.data %>%
    tibble::rownames_to_column("cell_id_full") %>%
    mutate(
      sample = as.character(orig.ident),
      barcode = if ("raw_cell_id" %in% colnames(.)) {
        sub("-1$", "", as.character(raw_cell_id))
      } else {
        sub("^[^_]+_", "", cell_id_full) |> sub("-1$", "", x = _)
      },
      patient_id = if ("patient_id" %in% colnames(.)) as.character(patient_id) else NA_character_,
      condition  = if ("condition"  %in% colnames(.)) as.character(condition)  else NA_character_,
      timepoint  = if ("timepoint"  %in% colnames(.)) as.character(timepoint)  else NA_character_
    )

  cat("\nLoaded metadata from annotated object\n")
  print(colnames(metadata_tbl))
  print(head(metadata_tbl[, c("sample","barcode","patient_id","condition","timepoint")], 10))
}

```


##load files VDJ
```{r}
#| label: load-vdj
vdj_list <- lapply(seq_len(nrow(sample_sheet)), function(i) {
  load_vdj_one(sample_sheet$sample[i], sample_sheet$outs[i])
})
names(vdj_list) <- sample_sheet$sample

contigs_all_pre <- rbindlist(lapply(vdj_list, `[[`, "contigs"), use.names = TRUE, fill = TRUE) %>% as.data.frame()
clonotypes_all  <- rbindlist(lapply(vdj_list, `[[`, "clonotypes"), use.names = TRUE, fill = TRUE) %>% as.data.frame()

metrics_list <- lapply(vdj_list, `[[`, "metrics")
metrics_list <- metrics_list[!vapply(metrics_list, is.null, logical(1))]
metrics_all <- if (length(metrics_list) > 0) {
  rbindlist(metrics_list, use.names = TRUE, fill = TRUE) %>% as.data.frame()
} else {
  NULL
}
```

## validate-standardize
```{r}
#| label: validate-and-standardize
need_cols <- c("barcode", "chain", "cdr3", "productive")
require_cols(contigs_all_pre, need_cols, "contigs_all_pre")

contigs_all_pre <- contigs_all_pre %>%
  mutate(
    sample = as.character(sample),
    barcode = as.character(barcode),
    cell_id = as.character(barcode),
    chain = clean_chain(chain),
    productive = if ("productive" %in% names(.)) to_logical(productive) else productive,
    high_confidence = if ("high_confidence" %in% names(.)) to_logical(high_confidence) else high_confidence,
    full_length = if ("full_length" %in% names(.)) to_logical(full_length) else full_length,
    umis = if ("umis" %in% names(.)) numify(umis) else NA_real_,
    reads = if ("reads" %in% names(.)) numify(reads) else NA_real_,
    cdr3 = as.character(cdr3),
    cdr3_nt = if ("cdr3_nt" %in% names(.)) as.character(cdr3_nt) else NA_character_,
    v_gene = if ("v_gene" %in% names(.)) as.character(v_gene) else NA_character_,
    j_gene = if ("j_gene" %in% names(.)) as.character(j_gene) else NA_character_,
    raw_clonotype_id = if ("raw_clonotype_id" %in% names(.)) as.character(raw_clonotype_id) else if ("clonotype_id" %in% names(.)) as.character(clonotype_id) else NA_character_
  )
```

##merge-optional-metadata
```{r}
#| label: merge-optional-metadata

contigs_all_pre$patient_id <- NA_character_
contigs_all_pre$condition  <- NA_character_
contigs_all_pre$timepoint  <- NA_character_

if (!is.null(metadata_tbl) && nrow(metadata_tbl) > 0) {
  message("--- Merging VDJ contigs with GEX metadata ---")

  md_sample_col    <- "sample"
  md_barcode_col   <- "barcode"
  md_patient_col   <- if ("patient_id" %in% colnames(metadata_tbl)) "patient_id" else NULL
  md_condition_col <- if ("condition" %in% colnames(metadata_tbl)) "condition" else NULL
  md_timepoint_col <- if ("timepoint" %in% colnames(metadata_tbl)) "timepoint" else NULL

  metadata_prep <- metadata_tbl %>%
    mutate(
      md_sample  = as.character(.data[[md_sample_col]]),
      md_barcode = sub("-1$", "", as.character(.data[[md_barcode_col]]))
    )

  # barcode + sample level mapping
  barcode_meta_map <- metadata_prep %>%
    transmute(
      sample = md_sample,
      barcode_clean = md_barcode,
      patient_id_join = if (!is.null(md_patient_col)) as.character(.data[[md_patient_col]]) else NA_character_,
      condition_join  = if (!is.null(md_condition_col)) as.character(.data[[md_condition_col]]) else NA_character_,
      timepoint_join  = if (!is.null(md_timepoint_col)) as.character(.data[[md_timepoint_col]]) else NA_character_
    ) %>%
    distinct(sample, barcode_clean, .keep_all = TRUE)

  # sample-level fallback
  sample_meta_map <- metadata_prep %>%
    transmute(
      sample = md_sample,
      patient_id_sample = if (!is.null(md_patient_col)) as.character(.data[[md_patient_col]]) else NA_character_,
      condition_sample  = if (!is.null(md_condition_col)) as.character(.data[[md_condition_col]]) else NA_character_,
      timepoint_sample  = if (!is.null(md_timepoint_col)) as.character(.data[[md_timepoint_col]]) else NA_character_
    ) %>%
    distinct(sample, .keep_all = TRUE)

  contigs_all_pre <- contigs_all_pre %>%
    mutate(
      sample = as.character(sample),
      barcode_clean = sub("-1$", "", as.character(barcode))
    ) %>%
    left_join(barcode_meta_map, by = c("sample", "barcode_clean")) %>%
    left_join(sample_meta_map, by = "sample") %>%
    mutate(
      patient_id = dplyr::coalesce(patient_id_join, patient_id_sample),
      condition  = dplyr::coalesce(condition_join, condition_sample),
      timepoint  = dplyr::coalesce(timepoint_join, timepoint_sample)
    ) %>%
    select(-any_of(c(
      "barcode_clean",
      "patient_id_join", "condition_join", "timepoint_join",
      "patient_id_sample", "condition_sample", "timepoint_sample"
    )))

  message(sprintf(
    "Merged metadata: patient_id non-NA = %d / %d ; timepoint non-NA = %d / %d",
    sum(!is.na(contigs_all_pre$patient_id)), nrow(contigs_all_pre),
    sum(!is.na(contigs_all_pre$timepoint)), nrow(contigs_all_pre)
  ))
}
```

# metadata-debug
```{r}
#| label: metadata-debug
cat("\n===== DEBUG MERGED METADATA =====\n")
print(
  contigs_all_pre %>%
    select(sample, barcode, patient_id, condition, timepoint) %>%
    distinct() %>%
    head(20)
)


cat("\nPer-sample merged metadata summary:\n")
print(
  contigs_all_pre %>%
    group_by(sample) %>%
    summarise(
      patient_vals = paste(unique(na.omit(patient_id)), collapse = ";"),
      condition_vals = paste(unique(na.omit(condition)), collapse = ";"),
      timepoint_vals = paste(unique(na.omit(timepoint)), collapse = ";"),
      .groups = "drop"
    )
)
```
##chain mode
```{r}
#| label: chain-mode
detected_chain_mode <- infer_chain_mode(contigs_all_pre$chain)
effective_chain_mode <- if (params$chain_mode == "auto") detected_chain_mode else params$chain_mode

contigs_all_pre <- filter_chain_mode(contigs_all_pre, "chain", params$chain_mode)
```



##QC-filtering
```{r}
#| label: qc-filtering
contigs_all_pre <- contigs_all_pre %>%
  mutate(
    cdr3_len = nchar(cdr3),
    has_stop = ifelse(is.na(cdr3), NA, str_detect(cdr3, "\\*")),
    has_frame_issue = ifelse(is.na(cdr3_nt), FALSE, nchar(cdr3_nt) %% 3 != 0)
  )

contigs_all_post <- contigs_all_pre %>%
  {
    x <- .
    if (params$require_productive && "productive" %in% names(x)) {
      x <- x %>% filter(productive)
    }
    if (params$require_high_conf && "high_confidence" %in% names(x)) {
      x <- x %>% filter(high_confidence)
    }
    if (params$require_full_length && "full_length" %in% names(x)) {
      x <- x %>% filter(full_length)
    }
    if ("umis" %in% names(x)) {
      x <- x %>% filter(is.na(umis) | umis >= params$min_umis)
    }
    if ("reads" %in% names(x)) {
      x <- x %>% filter(is.na(reads) | reads >= params$min_reads)
    }

    x <- x %>% filter(!is.na(cdr3), cdr3 != "")

    if (isTRUE(params$drop_stop_or_frames)) {
      x <- x %>% filter(!grepl("\\*", cdr3))
      if ("cdr3_nt" %in% names(x)) {
        x <- x %>% filter(is.na(cdr3_nt) | nchar(cdr3_nt) %% 3 == 0)
      }
    }

    x <- x %>%
      filter(
        nchar(cdr3) >= params$cdr3_aa_min,
        nchar(cdr3) <= params$cdr3_aa_max
      )

    if (isTRUE(params$keep_one_per_chain)) {
      order_vars <- intersect(c("umis", "reads"), names(x))
      if (length(order_vars) > 0) {
        x <- x %>%
          group_by(sample, barcode, chain) %>%
          arrange(across(all_of(order_vars), desc), .by_group = TRUE) %>%
          slice(1) %>%
          ungroup()
      } else {
        x <- x %>% distinct(sample, barcode, chain, .keep_all = TRUE)
      }
    }

    x
  }

if (!isTRUE(params$keep_dual_alpha)) {
  contigs_all_post <- contigs_all_post %>%
    group_by(sample, barcode) %>%
    filter(!(chain == "TRA" & row_number() > 1)) %>%
    ungroup()
}

if (!isTRUE(params$keep_dual_beta)) {
  contigs_all_post <- contigs_all_post %>%
    group_by(sample, barcode) %>%
    filter(!(chain == "TRB" & row_number() > 1)) %>%
    ungroup()
}

if (isTRUE(params$keep_paired_only)) {
  if (effective_chain_mode == "T-AB") {
    contigs_all_post <- contigs_all_post %>%
      group_by(sample, barcode) %>%
      filter(any(chain == "TRA") & any(chain == "TRB")) %>%
      ungroup()
  } else if (effective_chain_mode == "T-GD") {
    contigs_all_post <- contigs_all_post %>%
      group_by(sample, barcode) %>%
      filter(any(chain == "TRG") & any(chain == "TRD")) %>%
      ungroup()
  }
}
```

## removed-contigs
```{r}
#| label: removed-contigs
key_cols <- intersect(
  c("sample","barcode","chain","cdr3","cdr3_nt","v_gene","j_gene","raw_clonotype_id"),
  colnames(contigs_all_pre)
)

if (length(key_cols) == 0) {
  removed <- contigs_all_pre[0, , drop = FALSE]
} else {
  removed <- dplyr::anti_join(
    contigs_all_pre,
    contigs_all_post,
    by = key_cols
  )
}
```

## QC-Summary before-after
```{r}
#| label: qc-summary-before-after
qc_summary <- contigs_all_pre %>%
  mutate(stage = "before") %>%
  bind_rows(contigs_all_post %>% mutate(stage = "after")) %>%
  group_by(sample, stage) %>%
  summarise(
    cells_with_any = n_distinct(barcode),
    contigs = n(),
    mean_contigs_per_cell = contigs / cells_with_any,
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = stage,
    values_from = c(cells_with_any, contigs, mean_contigs_per_cell)
  )

before_df <- contigs_all_pre %>% count(sample, name = "contigs_before")
after_df  <- contigs_all_post %>% count(sample, name = "contigs_after")

qc_retention <- before_df %>%
  left_join(after_df, by = "sample") %>%
  mutate(
    contigs_after = replace_na(contigs_after, 0L),
    dropped = pmax(contigs_before - contigs_after, 0L),
    retained_prop = ifelse(contigs_before > 0, contigs_after / contigs_before, NA_real_),
    retained_pct = 100 * retained_prop
  ) %>%
  arrange(desc(contigs_before))
```

### Pairing Summary
```{r}
#| label: pairing-summary
pairing_all <- contigs_all_post %>%
  group_by(sample, cell_id = barcode) %>%
  summarise(
    patient_id = first(na.omit(patient_id)) %||% NA_character_,
    condition = first(na.omit(condition)) %||% NA_character_,
    timepoint = first(na.omit(timepoint)) %||% NA_character_,
    n_prod = n(),
    n_TRA = sum(chain == "TRA"),
    n_TRB = sum(chain == "TRB"),
    n_TRG = sum(chain == "TRG"),
    n_TRD = sum(chain == "TRD"),
    has_TRA = any(chain == "TRA"),
    has_TRB = any(chain == "TRB"),
    has_TRG = any(chain == "TRG"),
    has_TRD = any(chain == "TRD"),
    alpha_ct = sum(chain == "TRA"),
    beta_ct = sum(chain == "TRB"),
    gamma_ct = sum(chain == "TRG"),
    delta_ct = sum(chain == "TRD"),
    cdr3a = paste(unique(cdr3[chain == "TRA"]), collapse = ";"),
    cdr3b = paste(unique(cdr3[chain == "TRB"]), collapse = ";"),
    cdr3g = paste(unique(cdr3[chain == "TRG"]), collapse = ";"),
    cdr3d = paste(unique(cdr3[chain == "TRD"]), collapse = ";"),
    trav = paste(unique(v_gene[chain == "TRA"]), collapse = ";"),
    trbv = paste(unique(v_gene[chain == "TRB"]), collapse = ";"),
    trgv = paste(unique(v_gene[chain == "TRG"]), collapse = ";"),
    trdv = paste(unique(v_gene[chain == "TRD"]), collapse = ";"),
    traj = paste(unique(j_gene[chain == "TRA"]), collapse = ";"),
    trbj = paste(unique(j_gene[chain == "TRB"]), collapse = ";"),
    trgj = paste(unique(j_gene[chain == "TRG"]), collapse = ";"),
    trdj = paste(unique(j_gene[chain == "TRD"]), collapse = ";"),
    clone_id = first(na.omit(raw_clonotype_id)) %||% NA_character_,
    max_reads = suppressWarnings(max(reads, na.rm = TRUE)),
    max_umis = suppressWarnings(max(umis, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    pairing = case_when(
      effective_chain_mode == "T-AB" & has_TRA & has_TRB ~ "Alpha+Beta",
      effective_chain_mode == "T-AB" & has_TRA & !has_TRB ~ "Alpha_only",
      effective_chain_mode == "T-AB" & !has_TRA & has_TRB ~ "Beta_only",
      effective_chain_mode == "T-GD" & has_TRG & has_TRD ~ "Gamma+Delta",
      effective_chain_mode == "T-GD" & has_TRG & !has_TRD ~ "Gamma_only",
      effective_chain_mode == "T-GD" & !has_TRG & has_TRD ~ "Delta_only",
      effective_chain_mode == "both" & (has_TRA & has_TRB) ~ "Alpha+Beta",
      effective_chain_mode == "both" & (has_TRG & has_TRD) ~ "Gamma+Delta",
      TRUE ~ "Other"
    ),
    multi_alpha = alpha_ct > 1,
    multi_beta = beta_ct > 1,
    multi_gamma = gamma_ct > 1,
    multi_delta = delta_ct > 1
  )
```


##CDR3 lengths
```{r}
#| label: cdr3-lengths
cdr3_len_all <- contigs_all_post %>%
  filter(!is.na(cdr3), cdr3 != "") %>%
  mutate(
    len = nchar(cdr3),
    chain_grp = case_when(
      chain == "TRA" ~ "TRA",
      chain == "TRB" ~ "TRB",
      chain == "TRG" ~ "TRG",
      chain == "TRD" ~ "TRD",
      TRUE ~ "Other"
    )
  )
```

##vj-usage
```{r}
#| label: vj-usage-and-clone-rank
v_usage <- NULL
j_usage <- NULL

if (isTRUE(params$show_vj_heatmaps) && any(!is.na(contigs_all_post$v_gene))) {
  v_usage <- contigs_all_post %>%
    filter(!is.na(v_gene), v_gene != "") %>%
    count(sample, v_gene, name = "n") %>%
    group_by(sample) %>%
    mutate(freq = n / sum(n)) %>%
    ungroup()

  top_v <- v_usage %>%
    group_by(v_gene) %>%
    summarise(total = sum(n), .groups = "drop") %>%
    arrange(desc(total)) %>%
    slice_head(n = params$top_n_v) %>%
    pull(v_gene)

  v_usage <- v_usage %>% filter(v_gene %in% top_v)
}

if (isTRUE(params$show_vj_heatmaps) && any(!is.na(contigs_all_post$j_gene))) {
  j_usage <- contigs_all_post %>%
    filter(!is.na(j_gene), j_gene != "") %>%
    count(sample, j_gene, name = "n") %>%
    group_by(sample) %>%
    mutate(freq = n / sum(n)) %>%
    ungroup()

  top_j <- j_usage %>%
    group_by(j_gene) %>%
    summarise(total = sum(n), .groups = "drop") %>%
    arrange(desc(total)) %>%
    slice_head(n = params$top_n_j) %>%
    pull(j_gene)

  j_usage <- j_usage %>% filter(j_gene %in% top_j)
}

clone_rank_tbl <- NULL
if (isTRUE(params$show_clone_rank_plot)) {
  clone_rank_tbl <- pairing_all %>%
    filter(!is.na(clone_id), clone_id != "") %>%
    count(sample, clone_id, name = "n_cells") %>%
    group_by(sample) %>%
    arrange(desc(n_cells), .by_group = TRUE) %>%
    mutate(rank = row_number()) %>%
    ungroup() %>%
    filter(n_cells >= params$min_clone_size_plot)
}
```

##Compact Summary
```{r}
#| label: compact-sample-summary
pairing_rates <- pairing_all %>%
  count(sample, pairing, name = "cells") %>%
  group_by(sample) %>%
  mutate(frac = cells / sum(cells)) %>%
  pivot_wider(names_from = pairing, values_from = c(cells, frac), values_fill = 0) %>%
  ungroup()

sample_qc <- pairing_all %>%
  group_by(sample) %>%
  summarise(
    patient_id = first(na.omit(patient_id)) %||% NA_character_,
    condition = first(na.omit(condition)) %||% NA_character_,
    timepoint = first(na.omit(timepoint)) %||% NA_character_,
    n_cells_total = n(),
    n_cells_paired = sum(pairing %in% c("Alpha+Beta","Gamma+Delta")),
    n_cells_multichain = sum(multi_alpha | multi_beta | multi_gamma | multi_delta, na.rm = TRUE),
    pct_cells_paired = safe_pct(n_cells_paired, n_cells_total),
    pct_cells_multichain = safe_pct(n_cells_multichain, n_cells_total),
    n_unique_clones = n_distinct(clone_id[!is.na(clone_id) & clone_id != ""]),
    median_max_reads = suppressWarnings(median(max_reads[is.finite(max_reads)], na.rm = TRUE)),
    median_max_umis = suppressWarnings(median(max_umis[is.finite(max_umis)], na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  left_join(qc_summary, by = "sample") %>%
  left_join(qc_retention, by = "sample") %>%
  left_join(pairing_rates, by = "sample")
```

##save outputs
```{r}
#| label: save-tables
save_table_safe(sample_sheet, "sample_sheet_resolved.tsv")
save_table_safe(contigs_all_pre, "contigs_before_qc.tsv")
save_table_safe(contigs_all_post, "contigs_after_qc.tsv")
save_table_safe(removed, "contigs_removed_by_qc.tsv")
save_table_safe(qc_summary, "vdj_qc_summary_before_after.tsv")
save_table_safe(qc_retention, "qc_contigs_before_after_summary.tsv")
save_table_safe(pairing_all, "pairing_status_all.tsv")
save_table_safe(sample_qc, "vdj_qc_per_sample_compact.tsv")
save_table_safe(cdr3_len_all, "cdr3_lengths_all.tsv")
save_table_safe(clonotypes_all, "clonotypes_all.tsv")
if (!is.null(metrics_all)) save_table_safe(metrics_all, "vdj_metrics_summary_all.tsv")
if (!is.null(v_usage)) save_table_safe(v_usage, "v_gene_usage.tsv")
if (!is.null(j_usage)) save_table_safe(j_usage, "j_gene_usage.tsv")
if (!is.null(clone_rank_tbl)) save_table_safe(clone_rank_tbl, "clone_rank_abundance.tsv")
```

##report Overview
```{r}
#| label: overview
n_samples <- n_distinct(sample_sheet$sample)
n_contigs_before <- nrow(contigs_all_pre)
n_contigs_after <- nrow(contigs_all_post)
n_cells_after <- n_distinct(pairing_all$cell_id)
pct_retained <- percent(n_contigs_after / max(1, n_contigs_before), accuracy = 0.1)

overview_tbl <- tibble(
  Metric = c(
    "Report label",
    "Resolved samples",
    "Detected chain mode",
    "Effective chain mode",
    "Contigs before QC",
    "Contigs after QC",
    "Retention rate",
    "Cells after QC",
    "Unique clonotypes loaded"
  ),
  Value = c(
    params$report_label,
    comma(n_samples),
    detected_chain_mode,
    effective_chain_mode,
    comma(n_contigs_before),
    comma(n_contigs_after),
    pct_retained,
    comma(n_cells_after),
    comma(nrow(clonotypes_all))
  )
)

kable(overview_tbl, caption = "High-level overview of VDJ QC analysis.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover","condensed"))
```



##sample level Summary
```{r}
#| label: sample-summary-table
print(colnames(sample_qc))

top_n_samples_table <- params$top_n_samples_table
if (is.null(top_n_samples_table) || is.na(top_n_samples_table)) {
  top_n_samples_table <- 50
}
top_n_samples_table <- as.integer(top_n_samples_table)

sample_qc_display <- sample_qc %>%
  mutate(
    retained_pct = sprintf("%.1f%%", retained_pct),
    pct_cells_paired = percent(pct_cells_paired, accuracy = 0.1),
    pct_cells_multichain = percent(pct_cells_multichain, accuracy = 0.1)
  )

if ("contigs_after" %in% colnames(sample_qc_display)) {
  sample_qc_display <- sample_qc_display %>%
    arrange(desc(contigs_after))
} else {
  sample_qc_display <- sample_qc_display %>%
    arrange(desc(n_cells_total))
}

sample_qc_display <- sample_qc_display %>%
  slice_head(n = top_n_samples_table)

# print(
#   kable(sample_qc_display, caption = "Per-sample VDJ QC compact summary.") %>%
#     kable_styling(
#       full_width = TRUE,
#       bootstrap_options = c("striped","hover","condensed","responsive")
#     )
# )
kable(sample_qc_display, caption = "Per-sample VDJ QC compact summary.") %>%
  kable_styling(full_width = TRUE, bootstrap_options = c("striped","hover","condensed","responsive"))

# #| label: sample-summary-table
# print(colnames(sample_qc))
# 
# sample_qc_display <- sample_qc %>%
#   mutate(
#     retained_pct = sprintf("%.1f%%", retained_pct),
#     pct_cells_paired = percent(pct_cells_paired, accuracy = 0.1),
#     pct_cells_multichain = percent(pct_cells_multichain, accuracy = 0.1)
#   )
# 
# if ("contigs_after" %in% colnames(sample_qc_display)) {
#   sample_qc_display <- sample_qc_display %>%
#     arrange(desc(contigs_after))
# } else {
#   sample_qc_display <- sample_qc_display %>%
#     arrange(desc(n_cells_total))
# }
# 
# sample_qc_display <- sample_qc_display %>%
#   slice_head(n = params$top_n_samples_table)
# 
# kable(sample_qc_display, caption = "Per-sample VDJ QC compact summary.") %>%
#   kable_styling(full_width = TRUE, bootstrap_options = c("striped","hover","condensed","responsive"))
```

## Before/After QC overview
```{r}
#| label: before-after-figures
p_before_counts <- qc_retention %>%
  ggplot(aes(x = reorder(sample, contigs_before), y = contigs_before)) +
  geom_col(fill = "grey50") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Contigs Before QC", x = NULL, y = "Contigs") +
  theme_scratch_pub(params$base_size)

p_after_counts <- qc_retention %>%
  ggplot(aes(x = reorder(sample, contigs_before), y = contigs_after)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Contigs After QC", x = NULL, y = "Contigs") +
  theme_scratch_pub(params$base_size)

p_retained <- qc_retention %>%
  ggplot(aes(x = reorder(sample, retained_prop), y = retained_pct)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  geom_text(aes(label = sprintf("%.1f%%", retained_pct)), hjust = 1.05, size = 3, color = "white") +
  scale_y_continuous(limits = c(0, 100)) +
  labs(title = "Retention After QC", x = NULL, y = "Retained (%)") +
  theme_scratch_pub(params$base_size)

patch_before_after <- p_before_counts + p_after_counts + p_retained + plot_layout(ncol = 3)
patch_before_after
save_plot_safe(patch_before_after, glue("qc_before_after_retention.{params$figure_format}"), width = 16, height = 5)
```

##CDr3 length distribution
```{r}
#| label: cdr3-length
p_len_all <- ggplot(cdr3_len_all, aes(x = len, fill = chain_grp)) +
  geom_histogram(position = "identity", alpha = 0.7, bins = 40) +
  labs(
    title = "CDR3 Length Distribution",
    subtitle = "Post-QC pooled across all samples.",
    x = "CDR3 amino acid length",
    y = "Count",
    fill = "Chain"
  ) +
  theme_scratch_pub(params$base_size)

p_len_all
save_plot_safe(p_len_all, glue("cdr3_length_hist_all.{params$figure_format}"))
```

## cdr3-length-facet
```{r}
#| label: cdr3-length-facet
#| fig-width: 16
#| fig-height: !expr ceiling(dplyr::n_distinct(cdr3_len_all$sample) / 4) * 3

n_samples_facet <- dplyr::n_distinct(cdr3_len_all$sample)
facet_ncol <- min(4, max(1, n_samples_facet))
facet_nrow <- ceiling(n_samples_facet / facet_ncol)

plot_width  <- max(14, facet_ncol * 3.8)
plot_height <- max(16, facet_nrow * 5.5)

p_len_facet <- ggplot(cdr3_len_all, aes(x = len, fill = chain_grp)) +
  geom_histogram(position = "identity", alpha = 0.7, bins = 40) +
  facet_wrap(~ sample, scales = "free_y", ncol = facet_ncol) +
  labs(
    title = "CDR3 Length Distribution by Sample",
    subtitle = "Post-QC distribution faceted by sample.",
    x = "CDR3 amino acid length",
    y = "Count",
    fill = "Chain"
  ) +
  theme_scratch_pub(params$base_size) +
  theme(
    axis.text.x = element_text(size = params$base_size - 4, angle = 45, hjust = 1),
    # INCREASED: Changed from -4 to -2 for better legibility
    axis.text.y = element_text(size = params$base_size - 2), 
    strip.text = element_text(size = params$base_size, face = "bold"),
    plot.margin = margin(15, 15, 15, 15)
  )

print(p_len_facet)
save_plot_safe(p_len_facet, glue("cdr3_length_hist_by_sample.{params$figure_format}"), width = plot_width, height = plot_height)

# p_len_facet <- ggplot(cdr3_len_all, aes(x = len, fill = chain_grp)) +
#   geom_histogram(position = "identity", alpha = 0.7, bins = 40) +
#   facet_wrap(~ sample, scales = "free_y") +
#   labs(
#     title = "CDR3 Length Distribution by Sample",
#     subtitle = "Post-QC distribution faceted by sample.",
#     x = "CDR3 amino acid length",
#     y = "Count",
#     fill = "Chain"
#   ) +
#   theme_scratch_pub(params$base_size)
# 
# p_len_facet
# save_plot_safe(p_len_facet, glue("cdr3_length_hist_by_sample.{params$figure_format}"), width = 12, height = 9)
```

##Pairing Plots
```{r}
#| label: pairing-plot
#| fig-width: 16
#| fig-height: !expr ceiling(dplyr::n_distinct(pairing_all$sample) / 4) * 3

n_samples_pair <- dplyr::n_distinct(pairing_all$sample)
facet_ncol <- min(4, max(1, n_samples_pair))
facet_nrow <- ceiling(n_samples_pair / facet_ncol)

plot_width  <- max(14, facet_ncol * 3.8)
plot_height <- max(16, facet_nrow * 5.5)

p_pair_by <- pairing_all %>%
  count(sample, pairing, name = "n") %>%
  ggplot(aes(x = pairing, y = n, fill = pairing)) +
  geom_col() +
  facet_wrap(~ sample, scales = "free_y", ncol = facet_ncol) +
  labs(
    title = "Pairing Status by Sample",
    subtitle = "Pairing assigned after VDJ QC.",
    x = NULL,
    y = "Cells",
    fill = "Pairing"
  ) +
  theme_scratch_pub(params$base_size) +
  theme(
    axis.text.x = element_text(size = params$base_size - 4, angle = 45, hjust = 1),
    # INCREASED: Changed from -4 to -2
    axis.text.y = element_text(size = params$base_size - 2),
    axis.title.y = element_text(size = params$base_size + 1, face = "bold"),
    strip.text = element_text(size = params$base_size, face = "bold"),
    plot.margin = margin(15, 15, 15, 15)
  )

print(p_pair_by)
save_plot_safe(p_pair_by, glue("pairing_bar_by_sample.{params$figure_format}"), width = plot_width, height = plot_height)

# p_pair_by <- pairing_all %>%
#   count(sample, pairing, name = "n") %>%
#   ggplot(aes(x = pairing, y = n, fill = pairing)) +
#   geom_col() +
#   facet_wrap(~ sample, scales = "free_y") +
#   labs(
#     title = "Pairing Status by Sample",
#     subtitle = "Pairing assigned after VDJ QC.",
#     x = NULL,
#     y = "Cells",
#     fill = "Pairing"
#   ) +
#   theme_scratch_pub(params$base_size) +
#   # theme(axis.text.x = element_text(angle = 45, hjust = 1))
#   theme(
#     axis.text.x = element_text(size = params$base_size - 5, angle = 45, hjust = 1),
#     axis.text.y = element_text(size = params$base_size - 5),
#     strip.text = element_text(size = params$base_size - 2, face = "bold"),
#     plot.margin = margin(10, 10, 10, 10)
#   )
# 
# p_pair_by
# save_plot_safe(p_pair_by, glue("pairing_bar_by_sample.{params$figure_format}"), width = 14, height = 10)
```

##Multichain plot
```{r}
#| label: multichain-plot
#| fig-width: 16
#| fig-height: !expr ceiling(dplyr::n_distinct(pairing_all$sample) / 4) * 3

multi_chain_long <- pairing_all %>%
  pivot_longer(
    cols = c("multi_alpha", "multi_beta", "multi_gamma", "multi_delta"),
    names_to = "type",
    values_to = "flag"
  ) %>%
  mutate(
    type = recode(type, multi_alpha = "Multiple TRA", multi_beta = "Multiple TRB", multi_gamma = "Multiple TRG", multi_delta = "Multiple TRD"),
    flag = factor(flag, levels = c(FALSE, TRUE), labels = c("No", "Yes"))
  ) %>%
  count(sample, type, flag, name = "n")

n_samples_multi <- dplyr::n_distinct(multi_chain_long$sample)
facet_ncol <- min(4, max(1, n_samples_multi))
facet_nrow <- ceiling(n_samples_multi / facet_ncol)

plot_width  <- max(14, facet_ncol * 3.8)
plot_height <- max(16, facet_nrow * 5.5)

p_multi_by <- ggplot(multi_chain_long, aes(x = type, y = n, fill = flag)) +
  geom_col(position = "stack") +
  facet_wrap(~ sample, scales = "free_y", ncol = facet_ncol) +
  labs(
    title = "Multiple Productive Chains",
    subtitle = "Post-QC multiple-chain flags by sample.",
    x = NULL,
    y = "Cells",
    fill = "Flag"
  ) +
  theme_scratch_pub(params$base_size) +
  theme(
    axis.text.x = element_text(size = params$base_size - 4, angle = 45, hjust = 1),
    # INCREASED: Changed from -4 to -2
    axis.text.y = element_text(size = params$base_size - 2),
    axis.title.y = element_text(size = params$base_size + 1, face = "bold"),
    strip.text = element_text(size = params$base_size, face = "bold"),
    plot.margin = margin(15, 15, 15, 15)
  )

print(p_multi_by)
save_plot_safe(p_multi_by, glue("multiple_chains_by_sample.{params$figure_format}"), width = plot_width, height = plot_height)


# multi_chain_long <- pairing_all %>%
#   pivot_longer(
#     cols = c("multi_alpha", "multi_beta", "multi_gamma", "multi_delta"),
#     names_to = "type",
#     values_to = "flag"
#   ) %>%
#   mutate(
#     type = recode(
#       type,
#       multi_alpha = "Multiple TRA",
#       multi_beta = "Multiple TRB",
#       multi_gamma = "Multiple TRG",
#       multi_delta = "Multiple TRD"
#     ),
#     flag = factor(flag, levels = c(FALSE, TRUE), labels = c("No", "Yes"))
#   ) %>%
#   count(sample, type, flag, name = "n")
# 
# p_multi_by <- ggplot(multi_chain_long, aes(x = type, y = n, fill = flag)) +
#   geom_col(position = "stack") +
#   facet_wrap(~ sample, scales = "free_y") +
#   labs(
#     title = "Multiple Productive Chains",
#     subtitle = "Post-QC multiple-chain flags by sample.",
#     x = NULL,
#     y = "Cells",
#     fill = "Flag"
#   ) +
#   theme_scratch_pub(params$base_size) +
#   # theme(axis.text.x = element_text(angle = 45, hjust = 1))
#   theme(
#     axis.text.x = element_text(size = params$base_size - 5, angle = 45, hjust = 1),
#     axis.text.y = element_text(size = params$base_size - 5),
#     strip.text = element_text(size = params$base_size - 2, face = "bold"),
#     plot.margin = margin(10, 10, 10, 10)
#   )
# 
# p_multi_by
# save_plot_safe(p_multi_by, glue("multiple_chains_by_sample.{params$figure_format}"), width = 12, height = 8)
```

##Read and UMI Support
```{r}
#| label: read-umi-plots
if (isTRUE(params$show_reads_umis) && any(!is.na(contigs_all_post$reads))) {
  p_reads <- contigs_all_post %>%
    filter(!is.na(reads), reads > 0) %>%
    ggplot(aes(x = reads, fill = chain)) +
    geom_histogram(bins = 50, alpha = 0.8, position = "identity") +
    scale_x_log10(labels = label_number(scale_cut = cut_short_scale())) +
    labs(
      title = "Read Support of Passing Contigs",
      subtitle = "Histogram shown on log10 scale.",
      x = "Reads",
      y = "Contig count",
      fill = "Chain"
    ) +
    theme_scratch_pub(params$base_size)

  print(p_reads)
  save_plot_safe(p_reads, glue("reads_support_histogram.{params$figure_format}"))
}

if (isTRUE(params$show_reads_umis) && any(!is.na(contigs_all_post$umis))) {
  p_umis <- contigs_all_post %>%
    filter(!is.na(umis), umis > 0) %>%
    ggplot(aes(x = umis, fill = chain)) +
    geom_histogram(bins = 50, alpha = 0.8, position = "identity") +
    scale_x_log10(labels = label_number(scale_cut = cut_short_scale())) +
    labs(
      title = "UMI Support of Passing Contigs",
      subtitle = "Histogram shown on log10 scale.",
      x = "UMIs",
      y = "Contig count",
      fill = "Chain"
    ) +
    theme_scratch_pub(params$base_size)

  print(p_umis)
  save_plot_safe(p_umis, glue("umis_support_histogram.{params$figure_format}"))
}
```

## V and J gene usage
```{r}
#| label: v-gene-heatmap
if (!is.null(v_usage) && nrow(v_usage) > 0) {
  v_mat <- v_usage %>%
    select(sample, v_gene, freq) %>%
    pivot_wider(names_from = sample, values_from = freq, values_fill = 0) %>%
    as.data.frame()

  rownames(v_mat) <- v_mat$v_gene
  v_mat$v_gene <- NULL
  v_mat <- as.matrix(v_mat)

  max_v <- max(v_mat, na.rm = TRUE)
  if (!is.finite(max_v) || max_v <= 0) max_v <- 1

  ht_v <- Heatmap(
    v_mat,
    name = "Freq",
    col = colorRamp2(
      c(0, max_v / 2, max_v),
      c("white", "gold", "firebrick")
    ),
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    row_names_side = "left",
    row_names_gp = grid::gpar(fontsize = 8),
    column_names_gp = grid::gpar(fontsize = 7),
    column_names_rot = 45,
    column_title = "Top V gene usage across samples",
    column_title_gp = grid::gpar(fontsize = 12, fontface = "bold"),
    heatmap_legend_param = list(title = "Frequency")
  )

  draw(ht_v, padding = grid::unit(c(5, 5, 5, 12), "mm"))
}

# #| label: v-gene-heatmap
# if (!is.null(v_usage) && nrow(v_usage) > 0) {
#   v_mat <- v_usage %>%
#     select(sample, v_gene, freq) %>%
#     pivot_wider(names_from = sample, values_from = freq, values_fill = 0) %>%
#     as.data.frame()
# 
#   rownames(v_mat) <- v_mat$v_gene
#   v_mat$v_gene <- NULL
#   v_mat <- as.matrix(v_mat)
# 
#   ht_v <- Heatmap(
#     v_mat,
#     name = "Freq",
#     col = colorRamp2(
#       c(0, max(v_mat, na.rm = TRUE) / 2, max(v_mat, na.rm = TRUE)),
#       c("white", "gold", "firebrick")
#     ),
#     cluster_rows = TRUE,
#     cluster_columns = TRUE,
#     row_names_side = "left",
#     column_title = "Top V gene usage across samples",
#     heatmap_legend_param = list(title = "Frequency")
#   )
# 
#   draw(ht_v)
# }
```

## J gene heatmap
```{r}
#| label: j-gene-heatmap
if (!is.null(j_usage) && nrow(j_usage) > 0) {
  j_mat <- j_usage %>%
    select(sample, j_gene, freq) %>%
    pivot_wider(names_from = sample, values_from = freq, values_fill = 0) %>%
    as.data.frame()

  rownames(j_mat) <- j_mat$j_gene
  j_mat$j_gene <- NULL
  j_mat <- as.matrix(j_mat)

  max_j <- max(j_mat, na.rm = TRUE)
  if (!is.finite(max_j) || max_j <= 0) max_j <- 1

  ht_j <- Heatmap(
    j_mat,
    name = "Freq",
    col = colorRamp2(
      c(0, max_j / 2, max_j),
      c("white", "skyblue", "navy")
    ),
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    row_names_side = "left",
    row_names_gp = grid::gpar(fontsize = 8),
    column_names_gp = grid::gpar(fontsize = 7),
    column_names_rot = 45,
    column_title = "Top J gene usage across samples",
    column_title_gp = grid::gpar(fontsize = 12, fontface = "bold"),
    heatmap_legend_param = list(title = "Frequency")
  )

  draw(ht_j, padding = grid::unit(c(5, 5, 5, 12), "mm"))
}

# #| label: j-gene-heatmap
# if (!is.null(j_usage) && nrow(j_usage) > 0) {
#   j_mat <- j_usage %>%
#     select(sample, j_gene, freq) %>%
#     pivot_wider(names_from = sample, values_from = freq, values_fill = 0) %>%
#     as.data.frame()
# 
#   rownames(j_mat) <- j_mat$j_gene
#   j_mat$j_gene <- NULL
#   j_mat <- as.matrix(j_mat)
# 
#   ht_j <- Heatmap(
#     j_mat,
#     name = "Freq",
#     col = colorRamp2(
#       c(0, max(j_mat, na.rm = TRUE) / 2, max(j_mat, na.rm = TRUE)),
#       c("white", "skyblue", "navy")
#     ),
#     cluster_rows = TRUE,
#     cluster_columns = TRUE,
#     row_names_side = "left",
#     column_title = "Top J gene usage across samples",
#     heatmap_legend_param = list(title = "Frequency")
#   )
# 
#   draw(ht_j)
# }
```

## Clone rank-abundance
```{r}
#| label: clone-rank-plot
#| fig-width: 16
#| fig-height: !expr max(12, 7 + ceiling(dplyr::n_distinct(pairing_all$sample) / 6) * 0.8)

# 1. Create the plotting table
clone_rank_tbl_plot <- pairing_all %>%
  filter(!is.na(clone_id), clone_id != "") %>%
  count(sample, clone_id, name = "n_cells") %>%
  group_by(sample) %>%
  arrange(desc(n_cells), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  ungroup()

save_table_safe(clone_rank_tbl_plot, "clone_rank_abundance.tsv")

# 2. Build the plot
if (nrow(clone_rank_tbl_plot) > 0) {
  n_samples_clone <- dplyr::n_distinct(clone_rank_tbl_plot$sample)
  legend_rows <- ceiling(n_samples_clone / 6)
  
  # Adjusted height calculation for a more compact look
  plot_height <- max(12, 7 + legend_rows * 0.8)

  p_clone_rank <- ggplot(clone_rank_tbl_plot, aes(x = rank, y = n_cells, color = sample)) +
    geom_line(linewidth = 0.8, alpha = 0.8) + 
    geom_point(size = 1.5, alpha = 0.8) +    
    scale_x_log10() +
    scale_y_log10() +
    labs(
      title = "Clone Rank-Abundance Curve",
      subtitle = "All detected clonotypes are shown.",
      x = "Clone rank (log10)",
      y = "Cells per clone (log10)",
      color = "Sample"
    ) +
    theme_scratch_pub(params$base_size) +
    theme(
      # LEGEND POSITION: Change "bottom" to "right" if you want it on the side
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.box = "horizontal",
      
      # AXIS TEXT (Tick labels: 1, 10, 100) - Increased to +4
      axis.text = element_text(size = params$base_size + 4, color = "black"),
      
      # AXIS TITLES (Labels: Clone rank, Cells per clone) - Increased to +6
      axis.title = element_text(size = params$base_size + 6, face = "bold"),
      
      # LEGEND TEXT (Sample IDs) - Increased to +2
      legend.text = element_text(size = params$base_size + 2),
      legend.title = element_text(size = params$base_size + 4, face = "bold"),
      
      plot.title = element_text(size = params$base_size + 8, face = "bold"),
      plot.subtitle = element_text(size = params$base_size + 4),
      
      # Log scales look better without x-axis rotation
      axis.text.x = element_text(angle = 0, hjust = 0.5),
      panel.grid.major = element_line(linewidth = 0.5, color = "grey90")
    ) +
    guides(color = guide_legend(nrow = legend_rows, byrow = TRUE))

  print(p_clone_rank)
  save_plot_safe(
    p_clone_rank,
    glue("clone_rank_abundance.{params$figure_format}"),
    width = 16,
    height = plot_height
  )
} else {
  cat("No clonotypes were available to build the clone rank-abundance curve.")
}

# #| label: clone-rank
# 
# 
# clone_rank_tbl_plot <- pairing_all %>%
#   filter(!is.na(clone_id), clone_id != "") %>%
#   count(sample, clone_id, name = "n_cells") %>%
#   group_by(sample) %>%
#   arrange(desc(n_cells), .by_group = TRUE) %>%
#   mutate(rank = row_number()) %>%
#   ungroup()
# 
# save_table_safe(clone_rank_tbl_plot, "clone_rank_abundance.tsv")
# 
# if (nrow(clone_rank_tbl_plot) > 0) {
#   p_clone_rank <- ggplot(clone_rank_tbl_plot, aes(x = rank, y = n_cells, color = sample)) +
#     geom_line(linewidth = 0.9, alpha = 0.8) +
#     geom_point(size = 1.2, alpha = 0.9) +
#     scale_x_log10() +
#     scale_y_log10() +
#     labs(
#       title = "Clone Rank-Abundance Curve",
#       subtitle = "All detected clonotypes are shown.",
#       x = "Clone rank (log10)",
#       y = "Cells per clone (log10)",
#       color = "Sample"
#     ) +
#     theme_scratch_pub(params$base_size)
# 
#   print(p_clone_rank)
#   save_plot_safe(p_clone_rank, glue("clone_rank_abundance.{params$figure_format}"))
# } else {
#   cat("No clonotypes were available to build the clone rank-abundance curve.")
# }
# 
# # if (!is.null(clone_rank_tbl) && nrow(clone_rank_tbl) > 0) {
# #   p_clone_rank <- ggplot(clone_rank_tbl, aes(x = rank, y = n_cells, color = sample)) +
# #     geom_line(linewidth = 0.9, alpha = 0.8) +
# #     geom_point(size = 1.2, alpha = 0.9) +
# #     scale_x_log10() +
# #     scale_y_log10() +
# #     labs(
# #       title = "Clone Rank-Abundance Curve",
# #       subtitle = glue("Only clones with at least {params$min_clone_size_plot} cells are shown."),
# #       x = "Clone rank (log10)",
# #       y = "Cells per clone (log10)",
# #       color = "Sample"
# #     ) +
# #     theme_scratch_pub(params$base_size)
# # 
# #   p_clone_rank
# #   save_plot_safe(p_clone_rank, glue("clone_rank_abundance.{params$figure_format}"))
# # }
```

## Optional metadata summaries
```{r}
#| label: metadata-panels
show_patient <- isTRUE(params$show_patient_panels) && any(!is.na(sample_qc$patient_id))
show_condition <- isTRUE(params$show_condition_panels) && any(!is.na(sample_qc$condition))
# show_timepoint <- isTRUE(params$show_timepoint_panels) && any(!is.na(sample_qc$timepoint))
show_timepoint <- isTRUE(params$show_timepoint_panels) &&
  any(!is.na(sample_qc$timepoint)) &&
  dplyr::n_distinct(na.omit(sample_qc$timepoint)) >= 2
```

## patient-summary
```{r}
#| label: patient-summary
if (show_patient) {
  patient_source <- sample_qc %>%
    select(sample, patient_id, n_cells_total, n_unique_clones) %>%
    distinct() %>%
    left_join(
      qc_retention %>% select(sample, contigs_before, contigs_after, retained_prop, retained_pct),
      by = "sample"
    )

  patient_tbl <- patient_source %>%
    filter(!is.na(patient_id), patient_id != "") %>%
    group_by(patient_id) %>%
    summarise(
      n_samples = n_distinct(sample),
      contigs_before = sum(contigs_before, na.rm = TRUE),
      contigs_after = sum(contigs_after, na.rm = TRUE),
      retained_prop = contigs_after / pmax(1, contigs_before),
      retained_pct = 100 * retained_prop,
      n_cells_total = sum(n_cells_total, na.rm = TRUE),
      n_unique_clones = sum(n_unique_clones, na.rm = TRUE),
      .groups = "drop"
    )

  kable(patient_tbl, caption = "Patient-level VDJ QC summary.") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover","condensed"))

  p_patient <- patient_tbl %>%
    mutate(patient_id = fct_reorder(patient_id, retained_prop)) %>%
    ggplot(aes(x = patient_id, y = retained_pct)) +
    geom_col(fill = "darkgreen") +
    coord_flip() +
    labs(
      title = "QC Retention by Patient",
      x = NULL,
      y = "Retained (%)"
    ) +
    theme_scratch_pub(params$base_size)

  p_patient
  save_plot_safe(p_patient, glue("retention_by_patient.{params$figure_format}"))
}
```

## Condition Summary
```{r}
#| label: condition-summary
if (show_condition) {
  condition_source <- sample_qc %>%
    select(sample, condition, n_cells_total) %>%
    distinct() %>%
    left_join(
      qc_retention %>% select(sample, contigs_before, contigs_after, retained_prop, retained_pct),
      by = "sample"
    )

  condition_tbl <- condition_source %>%
    filter(!is.na(condition), condition != "") %>%
    group_by(condition) %>%
    summarise(
      n_samples = n_distinct(sample),
      contigs_before = sum(contigs_before, na.rm = TRUE),
      contigs_after = sum(contigs_after, na.rm = TRUE),
      retained_prop = contigs_after / pmax(1, contigs_before),
      retained_pct = 100 * retained_prop,
      n_cells_total = sum(n_cells_total, na.rm = TRUE),
      .groups = "drop"
    )

  kable(condition_tbl, caption = "Condition-level VDJ QC summary.") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover","condensed"))

  p_condition <- condition_tbl %>%
    mutate(condition = fct_reorder(condition, retained_prop)) %>%
    ggplot(aes(x = condition, y = retained_pct)) +
    geom_col(fill = "purple4") +
    coord_flip() +
    labs(
      title = "QC Retention by Condition",
      x = NULL,
      y = "Retained (%)"
    ) +
    theme_scratch_pub(params$base_size)

  p_condition
  save_plot_safe(p_condition, glue("retention_by_condition.{params$figure_format}"))
}
```

## timepoint summary
```{r}
#| label: timepoint-summary
if (show_timepoint) {
  time_source <- sample_qc %>%
    select(sample, timepoint) %>%
    distinct() %>%
    left_join(
      qc_retention %>% select(sample, contigs_before, contigs_after, retained_prop, retained_pct),
      by = "sample"
    )

  time_tbl <- time_source %>%
    filter(!is.na(timepoint), timepoint != "") %>%
    group_by(timepoint) %>%
    summarise(
      n_samples = n_distinct(sample),
      contigs_before = sum(contigs_before, na.rm = TRUE),
      contigs_after = sum(contigs_after, na.rm = TRUE),
      retained_prop = contigs_after / pmax(1, contigs_before),
      retained_pct = 100 * retained_prop,
      .groups = "drop"
    )

  kable(time_tbl, caption = "Timepoint-level VDJ QC summary.") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover","condensed"))

  p_time <- time_tbl %>%
    mutate(timepoint = fct_inorder(timepoint)) %>%
    ggplot(aes(x = timepoint, y = retained_pct, group = 1)) +
    geom_line(linewidth = 1, color = "steelblue4") +
    geom_point(size = 3, color = "steelblue4") +
    labs(
      title = "QC Retention Across Timepoints",
      x = "Timepoint",
      y = "Retained (%)"
    ) +
    theme_scratch_pub(params$base_size)

  p_time
  save_plot_safe(p_time, glue("retention_by_timepoint.{params$figure_format}"))
}
```

## Warnings
```{r}
#| label: warnings
warn_tbl <- tibble(
  warning = c(
    "Low contig retention",
    "High multi-chain burden",
    "Low paired fraction",
    "Missing reads column",
    "Missing UMIs column",
    "Missing patient metadata",
    "Missing condition metadata",
    "Missing timepoint metadata"
  ),
  triggered = c(
    n_contigs_after / max(1, n_contigs_before) < 0.5,
    mean(pairing_all$multi_alpha | pairing_all$multi_beta | pairing_all$multi_gamma | pairing_all$multi_delta, na.rm = TRUE) > 0.1,
    mean(pairing_all$pairing %in% c("Alpha+Beta","Gamma+Delta"), na.rm = TRUE) < 0.3,
    all(is.na(contigs_all_post$reads)),
    all(is.na(contigs_all_post$umis)),
    !show_patient,
    !show_condition,
    !show_timepoint
  ),
  interpretation = c(
    "Less than half of contigs were retained after QC. Inspect productive, high-confidence, and full-length filters as well as barcode/sample path resolution.",
    "More than 10% of cells are multi-chain under the current settings. This may indicate biological dual-chain usage, technical artifacts, or doublets.",
    "Paired-chain fraction is low after QC. Check assay recovery, allowed chain mode, and pairing requirements.",
    "Read-support plots were skipped because no reads column was available.",
    "UMI-support plots were skipped because no UMIs column was available.",
    "Patient-level summaries were skipped because patient metadata were unavailable.",
    "Condition-level summaries were skipped because condition metadata were unavailable.",
    "Timepoint-level summaries were skipped because timepoint metadata were unavailable."
  )
) %>%
  filter(triggered)

if (nrow(warn_tbl) == 0) {
  cat("No major automatic warnings were triggered under the current QC settings.")
} else {
  kable(warn_tbl %>% select(-triggered), caption = "Automatically generated QC warnings and notes.") %>%
    kable_styling(full_width = TRUE, bootstrap_options = c("striped","hover","condensed"))
}
```

## session info
```{r}
#| label: session-info
writeLines(capture.output(sessionInfo()), file.path(params$outdir, "sessionInfo.vdj_qc.txt"))
sessionInfo()
```


