---
title: "SCRATCH-TCR: Clonotype Clustering Consensus Report"
author: "Syed Shujaat Ali Zaidi"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    theme: cosmo
    df-print: paged
execute:
  echo: false
  warning: false
  message: false

params:
  # ======================================================
  # Inputs
  # ======================================================
  seurat_rds: "data/seurat_with_GIANA.rds"
  export_cells_file: "data/giana_export_cells.tsv"
  outdir: "Clonotype_Clustering_Consensus_Report"

  # Optional method-specific exports
  gliph_export_cells_file: ""
  tcrdist_export_cells_file: ""
  giana_export_cells_file: ""

  # Optional previous outputs
  metadata_file: ""
  previous_summary_file: ""

  # ======================================================
  # Output controls
  # ======================================================
  data_dir: "data"
  tables_dir: "tables"
  figures_dir: "figures"
  save_tables: true
  save_figures: true
  save_updated_seurat: true
  figure_format: "png"
  figure_width: 10
  figure_height: 7
  figure_dpi: 300
  base_size: 12

  # ======================================================
  # Metadata mapping in Seurat/export
  # ======================================================
  label_col: ""
  sample_col: "sample"
  patient_col: "patient"
  condition_col: "condition"
  timepoint_col: "timepoint"
  batch_col: "batch"
  clone_id_col: "clone_id"
  clone_size_col: "clone_size"
  paired_tcr_col: "paired_tcr"
  has_tcr_col: "has_tcr"

  label_candidates: "annot;Annotation;celltype;CellType;predicted_labels;predicted.celltype.l2;predicted.celltype.l1;celltypist;celltypist_label;azimuth_labels;seurat_clusters"
  sample_candidates: "sample;META_SAMPLE;orig.ident;sample_id;Sample;SampleID"
  patient_candidates: "patient;META_PATIENT;patient_id;Patient;subject;donor;case_id"
  condition_candidates: "condition;META_TIMECOND;Condition;group;Group;status"
  timepoint_candidates: "timepoint;META_TIMEPOINT;Timepoint;visit;Visit;day;Day"
  batch_candidates: "batch;META_BATCH;Batch;library;Library;run;Run;lane;Lane"
  clone_id_candidates: "clone_id;CTaa;clonotype;clone"
  clone_size_candidates: "clone_size;CloneSize;clone_n"
  paired_tcr_candidates: "paired_tcr;paired;is_paired"
  has_tcr_candidates: "has_tcr;hasTCR;tcr_positive"

  # ======================================================
  # Method-specific columns
  # ======================================================
  gliph_cluster_col: "gliph_cluster"
  gliph_pattern_col: "gliph_pattern"
  tcrdist_cluster_col: "tcrdist_cluster"
  tcrdist_group_col: "tcrdist_group"
  giana_cluster_col: "giana_cluster"
  giana_group_col: "giana_group"

  gliph_cluster_candidates: "gliph_cluster;cluster;gliph2_cluster"
  gliph_pattern_candidates: "gliph_pattern;pattern;motif"
  tcrdist_cluster_candidates: "tcrdist_cluster;cluster;tcrdist_cluster_id"
  tcrdist_group_candidates: "tcrdist_group;group;metaclonotype"
  giana_cluster_candidates: "giana_cluster;cluster;giana_cluster_id"
  giana_group_candidates: "giana_group;group;meta_group"

  # ======================================================
  # Consensus controls
  # ======================================================
  consensus_min_methods: 2
  prefer_method_order: "gliph;tcrdist;giana"
  use_majority_vote: true
  consensus_label_prefix: "CONS"
  assign_singleton_consensus: false
  min_cells_per_group: 10
  min_cluster_size_plot: 3
  top_n_clusters: 20
  top_n_states_heatmap: 30

  # ======================================================
  # Embedding controls
  # ======================================================
  reduction_use: "umap"
  make_umap_if_missing: false
  umap_dims_max: 30
  umap_nfeatures: 3000
  raster_large_umap: true
  label_clusters: true

  # ======================================================
  # Figure/report toggles
  # ======================================================
  show_consensus_umap: true
  show_consensus_size_barplot: true
  show_method_overlap_heatmap: true
  show_method_cluster_count_barplot: true
  show_consensus_annotation_heatmap: true
  show_consensus_sample_heatmap: true
  show_fraction_by_annotation: true
  show_fraction_by_sample: true
  show_fraction_by_condition: true
  show_fraction_by_patient: true
  show_method_presence_barplot: true
  show_summary_tables: true

  report_label: "Clonotype Clustering Consensus"
---

###############################################################################
# Integrates clonotype clustering results from GLIPH2, TCRdist3, and GIANA into a unified comparison framework within the TCR–GEX Seurat object.
# Quantifies overlap across methods and derives a consensus clonotype cluster assignment using configurable multi-method support rules.
# Supports flexible operation when all three methods are available or when only a subset of methods could be run.
# Generates publication-quality visualizations of method overlap, consensus cluster structure, and cluster distribution across cell states and samples.
# Exports an updated Seurat object and downstream-ready tables for final repertoire analysis and reporting.
###############################################################################

# setup
```{r}
#| label: setup
suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(forcats)
  library(scales)
  library(glue)
  library(knitr)
  library(kableExtra)
  library(ComplexHeatmap)
  library(circlize)
  library(patchwork)
})

options(stringsAsFactors = FALSE)

dir.create(params$outdir, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$data_dir), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$tables_dir), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$figures_dir), recursive = TRUE, showWarnings = FALSE)

`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0 && !all(is.na(a))) a else b

theme_scratch_pub <- function(base_size = 12) {
  theme_bw(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 2, hjust = 0),
      plot.subtitle = element_text(size = base_size, hjust = 0),
      axis.title = element_text(face = "bold"),
      axis.text = element_text(color = "black"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(linewidth = 0.2, color = "grey90"),
      strip.background = element_rect(fill = "grey95", color = "grey80"),
      strip.text = element_text(face = "bold"),
      legend.title = element_text(face = "bold"),
      legend.key = element_blank(),
      plot.caption = element_text(size = base_size - 2, color = "grey40")
    )
}

save_plot_safe <- function(plot_obj, filename,
                           width = params$figure_width,
                           height = params$figure_height,
                           dpi = params$figure_dpi) {
  if (!isTRUE(params$save_figures)) return(invisible(NULL))
  ggsave(
    filename = file.path(params$outdir, params$figures_dir, filename),
    plot = plot_obj,
    width = width,
    height = height,
    dpi = dpi,
    bg = "white",
    limitsize = FALSE
  )
}

save_table_safe <- function(df, filename) {
  if (!isTRUE(params$save_tables)) return(invisible(NULL))
  fwrite(df, file.path(params$outdir, params$tables_dir, filename), sep = "\t")
}

save_rds_safe <- function(obj, filename) {
  if (!isTRUE(params$save_updated_seurat)) return(invisible(NULL))
  saveRDS(obj, file.path(params$outdir, params$data_dir, filename))
}

safe_read_table <- function(path) {
  if (is.null(path) || is.na(path) || path == "" || !file.exists(path)) return(NULL)
  ext <- tolower(tools::file_ext(path))
  if (ext %in% c("tsv", "tab", "txt")) {
    fread(path, sep = "\t")
  } else {
    fread(path)
  }
}

require_file <- function(path, msg = NULL) {
  if (!file.exists(path)) stop(msg %||% paste("Missing required file:", path))
  path
}
# 
# first_existing_col <- function(df, preferred, candidates = character()) {
#   if (!is.null(preferred) && preferred != "" && preferred %in% colnames(df)) return(preferred)
#   hits <- intersect(candidates, colnames(df))
#   if (length(hits) > 0) return(hits[[1]])
#   NULL
# }


safe_make_umap <- function(seu, reduction_name = "umap", dims = 1:30, nfeatures = 3000, seed = 1234) {
  if ("RNA" %in% names(seu@assays)) DefaultAssay(seu) <- "RNA"
  npcs <- min(max(dims), max(2, ncol(seu) - 1), 50)
  nf <- min(nfeatures, nrow(seu))

  if (!"pca" %in% Reductions(seu)) {
    seu <- FindVariableFeatures(seu, nfeatures = nf, verbose = FALSE)
    seu <- ScaleData(seu, verbose = FALSE)
    seu <- RunPCA(seu, npcs = npcs, verbose = FALSE)
  }

  use_dims <- dims[dims <= npcs]
  if (length(use_dims) < 2) use_dims <- 1:min(10, npcs)

  seu <- FindNeighbors(seu, dims = use_dims, verbose = FALSE)
  seu <- RunUMAP(seu, dims = use_dims, reduction.name = reduction_name, verbose = FALSE)
  seu
}

make_method_signature <- function(gliph, tcrdist, giana) {
  paste0(
    "GLIPH:", ifelse(is.na(gliph) | gliph == "", "NA", gliph),
    "|TCRdist:", ifelse(is.na(tcrdist) | tcrdist == "", "NA", tcrdist),
    "|GIANA:", ifelse(is.na(giana) | giana == "", "NA", giana)
  )
}

safe_n_unique <- function(x) length(unique(na.omit(x)))

sanitize_param_string <- function(x) {
  if (is.null(x) || length(x) == 0) return(NULL)
  trimws(gsub("\\u00A0", " ", as.character(x)))
}

normalize_colnames <- function(df) {
  colnames(df) <- trimws(gsub("\\u00A0", " ", colnames(df)))
  df
}

split_candidate_string <- function(x) {
  if (is.null(x) || length(x) == 0 || is.na(x) || x == "") return(character())
  x <- gsub('"', "", as.character(x))
  x <- trimws(unlist(strsplit(x, ";", fixed = TRUE)))
  x[nzchar(x)]
}

first_existing_col <- function(df, preferred, candidates = character()) {
  df <- normalize_colnames(df)
  preferred <- sanitize_param_string(preferred)

  if (length(candidates) == 1 && is.character(candidates)) {
    candidates <- split_candidate_string(candidates)
  }
  candidates <- trimws(gsub("\\u00A0", " ", as.character(candidates)))

  if (!is.null(preferred) && preferred != "" && preferred %in% colnames(df)) return(preferred)

  hits <- intersect(candidates, colnames(df))
  if (length(hits) > 0) return(hits[[1]])
  NULL
}


```

# Input Loading
```{r}
#| label: read-inputs
require_file(params$seurat_rds, sprintf("Seurat RDS not found: %s", params$seurat_rds))

seu <- readRDS(params$seurat_rds)
export_cells_main <- safe_read_table(params$export_cells_file)
gliph_export_tbl <- safe_read_table(params$gliph_export_cells_file)
tcrdist_export_tbl <- safe_read_table(params$tcrdist_export_cells_file)
giana_export_tbl <- safe_read_table(params$giana_export_cells_file)
metadata_tbl <- safe_read_table(params$metadata_file)
previous_summary_tbl <- safe_read_table(params$previous_summary_file)

if (is.null(export_cells_main)) {
  export_cells_main <- seu@meta.data %>% tibble::rownames_to_column("cell_id")
}
```

# Resolve Columns
```{r}
#| label: resolve-columns
md <- normalize_colnames(seu@meta.data)
seu@meta.data <- md

label_col <- first_existing_col(md, params$label_col, params$label_candidates)
sample_col <- first_existing_col(md, params$sample_col, params$sample_candidates)
patient_col <- first_existing_col(md, params$patient_col, params$patient_candidates)
condition_col <- first_existing_col(md, params$condition_col, params$condition_candidates)
timepoint_col <- first_existing_col(md, params$timepoint_col, params$timepoint_candidates)
batch_col <- first_existing_col(md, params$batch_col, params$batch_candidates)
clone_id_col <- first_existing_col(md, params$clone_id_col, params$clone_id_candidates)
clone_size_col <- first_existing_col(md, params$clone_size_col, params$clone_size_candidates)
paired_tcr_col <- first_existing_col(md, params$paired_tcr_col, params$paired_tcr_candidates)
has_tcr_col <- first_existing_col(md, params$has_tcr_col, params$has_tcr_candidates)

gliph_cluster_col <- first_existing_col(md, params$gliph_cluster_col, params$gliph_cluster_candidates)
gliph_pattern_col <- first_existing_col(md, params$gliph_pattern_col, params$gliph_pattern_candidates)
tcrdist_cluster_col <- first_existing_col(md, params$tcrdist_cluster_col, params$tcrdist_cluster_candidates)
tcrdist_group_col <- first_existing_col(md, params$tcrdist_group_col, params$tcrdist_group_candidates)
giana_cluster_col <- first_existing_col(md, params$giana_cluster_col, params$giana_cluster_candidates)
giana_group_col <- first_existing_col(md, params$giana_group_col, params$giana_group_candidates)

if (is.null(label_col)) stop("Could not resolve an annotation label column in Seurat metadata.")
if (is.null(sample_col)) stop("Could not resolve sample column in Seurat metadata.")
if (is.null(patient_col)) stop("Could not resolve patient column in Seurat metadata.")
if (is.null(condition_col)) stop("Could not resolve condition column in Seurat metadata.")
if (is.null(timepoint_col)) stop("Could not resolve timepoint column in Seurat metadata.")
if (is.null(clone_id_col)) stop("Could not resolve a clone ID column in Seurat metadata.")
if (is.null(clone_size_col)) stop("Could not resolve a clone size column in Seurat metadata.")
if (is.null(has_tcr_col)) stop("Could not resolve a has_tcr column in Seurat metadata.")
if (is.null(paired_tcr_col)) stop("Could not resolve a paired_tcr column in Seurat metadata.")

if (!"META_SAMPLE" %in% colnames(seu@meta.data) && !is.null(sample_col)) {
  seu$META_SAMPLE <- as.character(seu@meta.data[[sample_col]])
}
if (!"META_PATIENT" %in% colnames(seu@meta.data) && !is.null(patient_col)) {
  seu$META_PATIENT <- as.character(seu@meta.data[[patient_col]])
}
if (!"META_TIMECOND" %in% colnames(seu@meta.data) && !is.null(condition_col)) {
  seu$META_TIMECOND <- as.character(seu@meta.data[[condition_col]])
}
if (!"META_TIMEPOINT" %in% colnames(seu@meta.data) && !is.null(timepoint_col)) {
  seu$META_TIMEPOINT <- as.character(seu@meta.data[[timepoint_col]])
}
if (!"META_BATCH" %in% colnames(seu@meta.data) && !is.null(batch_col)) {
  seu$META_BATCH <- as.character(seu@meta.data[[batch_col]])
}

print(list(
  resolved_label_col = label_col,
  resolved_sample_col = sample_col,
  resolved_patient_col = patient_col,
  resolved_condition_col = condition_col,
  resolved_timepoint_col = timepoint_col,
  resolved_batch_col = batch_col,
  resolved_clone_id_col = clone_id_col,
  resolved_clone_size_col = clone_size_col,
  resolved_paired_tcr_col = paired_tcr_col,
  resolved_has_tcr_col = has_tcr_col,
  resolved_gliph_cluster_col = gliph_cluster_col,
  resolved_gliph_pattern_col = gliph_pattern_col,
  resolved_tcrdist_cluster_col = tcrdist_cluster_col,
  resolved_tcrdist_group_col = tcrdist_group_col,
  resolved_giana_cluster_col = giana_cluster_col,
  resolved_giana_group_col = giana_group_col
))

```

# Ensure Embedding
```{r}
#| label: ensure-embedding
reduction_to_use <- params$reduction_use
if (!(reduction_to_use %in% names(seu@reductions)) && isTRUE(params$make_umap_if_missing)) {
  seu <- safe_make_umap(
    seu,
    reduction_name = params$reduction_use,
    dims = 1:params$umap_dims_max,
    nfeatures = params$umap_nfeatures
  )
}
if (!(reduction_to_use %in% names(seu@reductions))) {
  reduction_to_use <- if ("umap" %in% names(seu@reductions)) "umap" else if ("tsne" %in% names(seu@reductions)) "tsne" else if ("pca" %in% names(seu@reductions)) "pca" else stop("No usable embedding found.")
}
```

# Build Export table
```{r}
# Build Export table
```{r}
#| label: build-export
embed_df <- as.data.frame(Embeddings(seu, reduction_to_use)) %>%
  tibble::rownames_to_column("cell_id")

export_cells <- seu@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  transmute(
    cell_id,
    sample = if (!is.null(sample_col)) as.character(.data[[sample_col]]) else NA_character_,
    patient = if (!is.null(patient_col)) as.character(.data[[patient_col]]) else NA_character_,
    condition = if (!is.null(condition_col)) as.character(.data[[condition_col]]) else NA_character_,
    timepoint = if (!is.null(timepoint_col)) as.character(.data[[timepoint_col]]) else NA_character_,
    batch = if (!is.null(batch_col)) as.character(.data[[batch_col]]) else NA_character_,
    annot = as.character(.data[[label_col]]),
    clone_id = as.character(.data[[clone_id_col]]),
    clone_size = suppressWarnings(as.numeric(.data[[clone_size_col]])),
    has_tcr = as.logical(.data[[has_tcr_col]]),
    paired_tcr = as.logical(.data[[paired_tcr_col]]),
    gliph_cluster = if (!is.null(gliph_cluster_col)) as.character(.data[[gliph_cluster_col]]) else NA_character_,
    gliph_pattern = if (!is.null(gliph_pattern_col)) as.character(.data[[gliph_pattern_col]]) else NA_character_,
    tcrdist_cluster = if (!is.null(tcrdist_cluster_col)) as.character(.data[[tcrdist_cluster_col]]) else NA_character_,
    tcrdist_group = if (!is.null(tcrdist_group_col)) as.character(.data[[tcrdist_group_col]]) else NA_character_,
    giana_cluster = if (!is.null(giana_cluster_col)) as.character(.data[[giana_cluster_col]]) else NA_character_,
    giana_group = if (!is.null(giana_group_col)) as.character(.data[[giana_group_col]]) else NA_character_
  ) %>%
  left_join(embed_df, by = "cell_id")

if (!is.null(export_cells_main) && "cell_id" %in% colnames(export_cells_main)) {
  export_cells_main2 <- export_cells_main %>%
    mutate(cell_id = as.character(cell_id)) %>%
    transmute(
      cell_id = cell_id,
      sample.main = if ("sample" %in% colnames(.)) as.character(sample) else NA_character_,
      patient.main = if ("patient" %in% colnames(.)) as.character(patient) else NA_character_,
      condition.main = if ("condition" %in% colnames(.)) as.character(condition) else NA_character_,
      timepoint.main = if ("timepoint" %in% colnames(.)) as.character(timepoint) else NA_character_,
      batch.main = if ("batch" %in% colnames(.)) as.character(batch) else NA_character_,
      annot.main = if ("annot" %in% colnames(.)) as.character(annot) else NA_character_,
      clone_id.main = if ("clone_id" %in% colnames(.)) as.character(clone_id) else NA_character_,
      clone_size.main = if ("clone_size" %in% colnames(.)) suppressWarnings(as.numeric(clone_size)) else NA_real_,
      has_tcr.main = if ("has_tcr" %in% colnames(.)) as.logical(has_tcr) else NA,
      paired_tcr.main = if ("paired_tcr" %in% colnames(.)) as.logical(paired_tcr) else NA
    )

  export_cells <- export_cells %>%
    left_join(export_cells_main2, by = "cell_id") %>%
    mutate(
      sample = coalesce(sample, sample.main),
      patient = coalesce(patient, patient.main),
      condition = coalesce(condition, condition.main),
      timepoint = coalesce(timepoint, timepoint.main),
      batch = coalesce(batch, batch.main),
      annot = coalesce(annot, annot.main),
      clone_id = coalesce(clone_id, clone_id.main),
      clone_size = coalesce(clone_size, clone_size.main),
      has_tcr = coalesce(has_tcr, has_tcr.main),
      paired_tcr = coalesce(paired_tcr, paired_tcr.main)
    ) %>%
    select(-ends_with(".main"))
}

if (!is.null(gliph_export_tbl) && "cell_id" %in% colnames(gliph_export_tbl)) {
  gliph_tbl2 <- gliph_export_tbl %>%
    mutate(cell_id = as.character(cell_id)) %>%
    transmute(
      cell_id = cell_id,
      gliph_cluster.gliph = if ("gliph_cluster" %in% colnames(.)) as.character(gliph_cluster) else NA_character_,
      gliph_pattern.gliph = if ("gliph_pattern" %in% colnames(.)) as.character(gliph_pattern) else NA_character_,
      gliph_group.gliph = if ("gliph_group" %in% colnames(.)) as.character(gliph_group) else NA_character_,
      gliph_motif.gliph = if ("gliph_motif" %in% colnames(.)) as.character(gliph_motif) else NA_character_
    )

  export_cells <- export_cells %>%
    left_join(gliph_tbl2, by = "cell_id") %>%
    mutate(
      gliph_cluster = coalesce(gliph_cluster, gliph_cluster.gliph),
      gliph_pattern = coalesce(gliph_pattern, gliph_pattern.gliph, gliph_group.gliph, gliph_motif.gliph)
    ) %>%
    select(-gliph_cluster.gliph, -gliph_pattern.gliph, -gliph_group.gliph, -gliph_motif.gliph)
}

if (!is.null(tcrdist_export_tbl) && "cell_id" %in% colnames(tcrdist_export_tbl)) {
  tcrdist_tbl2 <- tcrdist_export_tbl %>%
    mutate(cell_id = as.character(cell_id)) %>%
    transmute(
      cell_id = cell_id,
      tcrdist_cluster.tcrdist = if ("tcrdist_cluster" %in% colnames(.)) as.character(tcrdist_cluster) else NA_character_,
      tcrdist_group.tcrdist = if ("tcrdist_group" %in% colnames(.)) as.character(tcrdist_group) else NA_character_,
      metaclonotype.tcrdist = if ("metaclonotype" %in% colnames(.)) as.character(metaclonotype) else NA_character_
    )

  export_cells <- export_cells %>%
    left_join(tcrdist_tbl2, by = "cell_id") %>%
    mutate(
      tcrdist_cluster = coalesce(tcrdist_cluster, tcrdist_cluster.tcrdist),
      tcrdist_group = coalesce(tcrdist_group, tcrdist_group.tcrdist, metaclonotype.tcrdist)
    ) %>%
    select(-tcrdist_cluster.tcrdist, -tcrdist_group.tcrdist, -metaclonotype.tcrdist)
}

if (!is.null(giana_export_tbl) && "cell_id" %in% colnames(giana_export_tbl)) {
  giana_tbl2 <- giana_export_tbl %>%
    mutate(cell_id = as.character(cell_id)) %>%
    transmute(
      cell_id = cell_id,
      giana_cluster.giana = if ("giana_cluster" %in% colnames(.)) as.character(giana_cluster) else NA_character_,
      giana_group.giana = if ("giana_group" %in% colnames(.)) as.character(giana_group) else NA_character_
    )

  export_cells <- export_cells %>%
    left_join(giana_tbl2, by = "cell_id") %>%
    mutate(
      giana_cluster = coalesce(giana_cluster, giana_cluster.giana),
      giana_group = coalesce(giana_group, giana_group.giana)
    ) %>%
    select(-giana_cluster.giana, -giana_group.giana)
}

save_table_safe(export_cells, "consensus_export_cells.tsv")

```

# Method presence and Overlap
```{r}
#| label: method-presence
export_cells <- export_cells %>%
  mutate(
    has_gliph = !is.na(gliph_cluster) | !is.na(gliph_pattern),
    has_tcrdist = !is.na(tcrdist_cluster) | !is.na(tcrdist_group),
    has_giana = !is.na(giana_cluster) | !is.na(giana_group),
    n_methods = as.integer(has_gliph) + as.integer(has_tcrdist) + as.integer(has_giana),
    method_signature = make_method_signature(gliph_cluster, tcrdist_cluster, giana_cluster)
  )

method_presence_summary <- tibble(
  method = c("GLIPH2", "TCRdist3", "GIANA"),
  n_cells = c(
    sum(export_cells$has_gliph, na.rm = TRUE),
    sum(export_cells$has_tcrdist, na.rm = TRUE),
    sum(export_cells$has_giana, na.rm = TRUE)
  ),
  frac_cells = c(
    mean(export_cells$has_gliph, na.rm = TRUE),
    mean(export_cells$has_tcrdist, na.rm = TRUE),
    mean(export_cells$has_giana, na.rm = TRUE)
  )
)
save_table_safe(method_presence_summary, "method_presence_summary.tsv")

pairwise_overlap_tbl <- tibble(
  comparison = c("GLIPH2_vs_TCRdist3", "GLIPH2_vs_GIANA", "TCRdist3_vs_GIANA"),
  n_overlap = c(
    sum(export_cells$has_gliph & export_cells$has_tcrdist, na.rm = TRUE),
    sum(export_cells$has_gliph & export_cells$has_giana, na.rm = TRUE),
    sum(export_cells$has_tcrdist & export_cells$has_giana, na.rm = TRUE)
  )
)
save_table_safe(pairwise_overlap_tbl, "pairwise_method_overlap.tsv")
```

# Build consensus labels
```{r}
#| label: build-consensus
export_cells <- export_cells %>%
  mutate(
    gliph_use = ifelse(!is.na(gliph_cluster) & gliph_cluster != "", gliph_cluster,
                       ifelse(!is.na(gliph_pattern) & gliph_pattern != "", paste0("PAT:", gliph_pattern), NA_character_)),
    tcrdist_use = ifelse(!is.na(tcrdist_cluster) & tcrdist_cluster != "", tcrdist_cluster,
                         ifelse(!is.na(tcrdist_group) & tcrdist_group != "", tcrdist_group, NA_character_)),
    giana_use = ifelse(!is.na(giana_cluster) & giana_cluster != "", giana_cluster,
                       ifelse(!is.na(giana_group) & giana_group != "", giana_group, NA_character_))
  )

consensus_base <- export_cells %>%
  mutate(
    n_methods_nonmissing = as.integer(!is.na(gliph_use)) + as.integer(!is.na(tcrdist_use)) + as.integer(!is.na(giana_use))
  )

if (isTRUE(params$use_majority_vote)) {
  consensus_base <- consensus_base %>%
    rowwise() %>%
    mutate(
      consensus_raw = {
        vals <- c(gliph_use, tcrdist_use, giana_use)
        vals <- vals[!is.na(vals) & vals != ""]
        if (length(vals) < params$consensus_min_methods) {
          if (isTRUE(params$assign_singleton_consensus) && length(vals) == 1) vals[1] else NA_character_
        } else {
          tab <- sort(table(vals), decreasing = TRUE)
          names(tab)[1]
        }
      }
    ) %>%
    ungroup()
} else {
  consensus_base <- consensus_base %>%
    mutate(
      consensus_raw = ifelse(
        n_methods_nonmissing >= params$consensus_min_methods,
        method_signature,
        ifelse(isTRUE(params$assign_singleton_consensus), coalesce(gliph_use, tcrdist_use, giana_use), NA_character_)
      )
    )
}

consensus_map <- consensus_base %>%
  filter(!is.na(consensus_raw), consensus_raw != "") %>%
  count(consensus_raw, name = "n_cells") %>%
  arrange(desc(n_cells)) %>%
  mutate(consensus_cluster = paste0(params$consensus_label_prefix, "_", row_number()))

export_cells <- consensus_base %>%
  left_join(consensus_map, by = "consensus_raw")

save_table_safe(consensus_map, "consensus_cluster_map.tsv")
```

# Merge consensus back into Seurat
```{r}
#| label: merge-consensus-into-seurat
seu_md <- seu@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  left_join(
    export_cells %>%
      select(cell_id, consensus_raw, consensus_cluster, n_methods_nonmissing, method_signature),
    by = "cell_id"
  )

rownames(seu_md) <- seu_md$cell_id
seu_md$cell_id <- NULL
seu@meta.data <- as.data.frame(seu_md)

save_rds_safe(seu, "seurat_with_consensus_clonotype_clusters.rds")
```

# Summary Tables
```{r}
#| label: summary-tables
summary_rollup <- tibble(
  metric = c(
    "Report label",
    "Cells in Seurat",
    "Cells with GLIPH2 annotation",
    "Cells with TCRdist3 annotation",
    "Cells with GIANA annotation",
    "Cells with >=2 methods",
    "Cells with consensus cluster",
    "Unique consensus clusters",
    "Unique annotations",
    "Unique samples",
    "Selected embedding"
  ),
  value = c(
    params$report_label,
    nrow(export_cells),
    sum(export_cells$has_gliph, na.rm = TRUE),
    sum(export_cells$has_tcrdist, na.rm = TRUE),
    sum(export_cells$has_giana, na.rm = TRUE),
    sum(export_cells$n_methods_nonmissing >= 2, na.rm = TRUE),
    sum(!is.na(export_cells$consensus_cluster)),
    safe_n_unique(export_cells$consensus_cluster),
    safe_n_unique(export_cells$annot),
    safe_n_unique(export_cells$sample),
    reduction_to_use
  )
)
save_table_safe(summary_rollup, "consensus_summary_rollup.tsv")

consensus_cluster_summary <- export_cells %>%
  filter(!is.na(consensus_cluster)) %>%
  count(consensus_cluster, consensus_raw, n_methods_nonmissing, name = "n_cells") %>%
  mutate(frac = n_cells / sum(n_cells)) %>%
  arrange(desc(n_cells))
save_table_safe(consensus_cluster_summary, "consensus_cluster_summary.tsv")

annotation_consensus_summary <- export_cells %>%
  filter(!is.na(annot)) %>%
  group_by(annot) %>%
  summarise(
    n_cells = n(),
    n_consensus = sum(!is.na(consensus_cluster)),
    frac_consensus = n_consensus / n_cells,
    .groups = "drop"
  ) %>%
  arrange(desc(frac_consensus))
save_table_safe(annotation_consensus_summary, "annotation_consensus_summary.tsv")

sample_consensus_summary <- export_cells %>%
  filter(!is.na(sample)) %>%
  group_by(sample) %>%
  summarise(
    n_cells = n(),
    n_consensus = sum(!is.na(consensus_cluster)),
    frac_consensus = n_consensus / n_cells,
    .groups = "drop"
  ) %>%
  arrange(desc(frac_consensus))
save_table_safe(sample_consensus_summary, "sample_consensus_summary.tsv")

if (any(!is.na(export_cells$condition))) {
  condition_consensus_summary <- export_cells %>%
    filter(!is.na(condition)) %>%
    group_by(condition) %>%
    summarise(
      n_cells = n(),
      n_consensus = sum(!is.na(consensus_cluster)),
      frac_consensus = n_consensus / n_cells,
      .groups = "drop"
    )
  save_table_safe(condition_consensus_summary, "condition_consensus_summary.tsv")
}

if (any(!is.na(export_cells$patient))) {
  patient_consensus_summary <- export_cells %>%
    filter(!is.na(patient)) %>%
    group_by(patient) %>%
    summarise(
      n_cells = n(),
      n_consensus = sum(!is.na(consensus_cluster)),
      frac_consensus = n_consensus / n_cells,
      .groups = "drop"
    )
  save_table_safe(patient_consensus_summary, "patient_consensus_summary.tsv")
}

consensus_annotation_tbl <- export_cells %>%
  filter(!is.na(consensus_cluster), !is.na(annot)) %>%
  count(consensus_cluster, annot, name = "n_cells")
save_table_safe(consensus_annotation_tbl, "consensus_cluster_annotation_composition.tsv")

consensus_sample_tbl <- export_cells %>%
  filter(!is.na(consensus_cluster), !is.na(sample)) %>%
  count(consensus_cluster, sample, name = "n_cells")
save_table_safe(consensus_sample_tbl, "consensus_cluster_sample_composition.tsv")
```

# Method Overlap matrix
```{r}
#| label: overlap-matrix

print(table(export_cells$has_gliph, useNA = "ifany"))
print(table(export_cells$has_tcrdist, useNA = "ifany"))
print(table(export_cells$has_giana, useNA = "ifany"))

print(colnames(export_cells))
print(sum(!is.na(export_cells$giana_cluster)))
print(sum(!is.na(export_cells$giana_group)))

overlap_matrix <- matrix(
  c(
    sum(export_cells$has_gliph, na.rm = TRUE),
    sum(export_cells$has_gliph & export_cells$has_tcrdist, na.rm = TRUE),
    sum(export_cells$has_gliph & export_cells$has_giana, na.rm = TRUE),
    sum(export_cells$has_tcrdist & export_cells$has_gliph, na.rm = TRUE),
    sum(export_cells$has_tcrdist, na.rm = TRUE),
    sum(export_cells$has_tcrdist & export_cells$has_giana, na.rm = TRUE),
    sum(export_cells$has_giana & export_cells$has_gliph, na.rm = TRUE),
    sum(export_cells$has_giana & export_cells$has_tcrdist, na.rm = TRUE),
    sum(export_cells$has_giana, na.rm = TRUE)
  ),
  nrow = 3,
  byrow = TRUE
)
rownames(overlap_matrix) <- c("GLIPH2", "TCRdist3", "GIANA")
colnames(overlap_matrix) <- c("GLIPH2", "TCRdist3", "GIANA")
save_table_safe(as.data.frame(overlap_matrix) %>% tibble::rownames_to_column("method"), "method_overlap_matrix.tsv")
```

# Report Overview
```{r}
#| label: overview
kable(summary_rollup, caption = "High-level overview of consensus clonotype clustering.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))
```

# Method Presence Plots
```{r}
#| label: method-presence-plot
if (isTRUE(params$show_method_presence_barplot)) {
  p_presence <- method_presence_summary %>%
    mutate(method = fct_reorder(method, frac_cells)) %>%
    ggplot(aes(x = method, y = frac_cells)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      title = "Method coverage across cells",
      x = NULL,
      y = "Fraction of cells annotated"
    ) +
    theme_scratch_pub(params$base_size)

  p_presence
  save_plot_safe(p_presence, glue("method_presence_barplot.{params$figure_format}"))
}
```

# method-overlap-heatmap
```{r}
#| label: method-overlap-heatmap
#| fig-width: 7
#| fig-height: 6
if (isTRUE(params$show_method_overlap_heatmap)) {

  finite_vals <- as.numeric(overlap_matrix)
  finite_vals <- finite_vals[is.finite(finite_vals)]

  if (length(finite_vals) == 0) {
    cat("No method-overlap heatmap was produced because the overlap matrix contained no finite values.\n")
  } else {
    min_val <- min(finite_vals, na.rm = TRUE)
    max_val <- max(finite_vals, na.rm = TRUE)

    if (max_val <= min_val) {
      cat("Method-overlap heatmap was skipped because all overlap values were identical.\n\n")

      print(
        kableExtra::kbl(
          as.data.frame(overlap_matrix) %>% tibble::rownames_to_column("method"),
          caption = "Method overlap matrix."
        ) %>%
          kableExtra::kable_styling(
            full_width = FALSE,
            bootstrap_options = c("striped", "hover", "condensed")
          )
      )
    } else {
      mid_val <- (min_val + max_val) / 2
      ht <- Heatmap(
        overlap_matrix,
        name = "Cells",
        col = colorRamp2(
          c(min_val, mid_val, max_val),
          c("white", "gold", "firebrick")
        ),
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        row_names_side = "left",
        row_names_gp = grid::gpar(fontsize = 10),
        column_names_gp = grid::gpar(fontsize = 10),
        rect_gp = grid::gpar(col = "grey85"),
        column_title = "Pairwise overlap between clustering methods",
        heatmap_legend_param = list(title = "Cells")
      )
            

      ComplexHeatmap::draw(ht)
    }
  }
}
```

# Consensus UMAP and cluster sizes
```{r}
#| label: consensus-cluster-plots
#| fig-width: 12
#| fig-height: 9

if (isTRUE(params$show_consensus_umap) && any(!is.na(export_cells$consensus_cluster))) {
  p_consensus <- DimPlot(
    seu,
    reduction = reduction_to_use,
    group.by = "consensus_cluster",
    label = FALSE,
    raster = isTRUE(params$raster_large_umap)
  ) +
    ggtitle("Embedding overlay: consensus clonotype clusters") +
    theme_scratch_pub(params$base_size) +
    theme(
      legend.position = "none"
    )

  print(p_consensus)
  save_plot_safe(
    p_consensus,
    glue("consensus_cluster_umap.{params$figure_format}"),
    width = 10,
    height = 8
  )
} else {
  cat("No consensus UMAP was produced because no cells received a consensus cluster assignment.\n")
}

if (isTRUE(params$show_consensus_size_barplot) && nrow(consensus_cluster_summary) > 0) {
  top_clusters_display <- params$top_n_clusters
  if (is.null(top_clusters_display) || length(top_clusters_display) == 0 || is.na(top_clusters_display)) {
    top_clusters_display <- 20
  }
  top_clusters_display <- as.integer(top_clusters_display)

  top_clusters <- consensus_cluster_summary %>%
    slice_head(n = top_clusters_display)

  p_bar <- top_clusters %>%
    mutate(consensus_cluster = fct_reorder(consensus_cluster, n_cells)) %>%
    ggplot(aes(x = consensus_cluster, y = n_cells)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    scale_y_continuous(labels = scales::comma) +
    labs(
      title = "Top consensus clonotype clusters",
      x = NULL,
      y = "Cells"
    ) +
    theme_scratch_pub(params$base_size)

  print(p_bar)
  save_plot_safe(
    p_bar,
    glue("consensus_cluster_size_barplot.{params$figure_format}")
  )
} else {
  cat("No consensus cluster-size barplot was produced because no consensus clusters were available.\n")
}

```

## Method cluster count comparison
```{r}
#| label: method-cluster-count-plot
if (isTRUE(params$show_method_cluster_count_barplot)) {
  method_cluster_counts <- tibble(
    method = c("GLIPH2", "TCRdist3", "GIANA", "Consensus"),
    n_clusters = c(
      safe_n_unique(export_cells$gliph_cluster),
      safe_n_unique(export_cells$tcrdist_cluster),
      safe_n_unique(export_cells$giana_cluster),
      safe_n_unique(export_cells$consensus_cluster)
    )
  )
  save_table_safe(method_cluster_counts, "method_cluster_counts.tsv")

  p_counts <- method_cluster_counts %>%
    mutate(method = fct_reorder(method, n_clusters)) %>%
    ggplot(aes(x = method, y = n_clusters)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    scale_y_continuous(labels = comma) +
    labs(
      title = "Number of clusters detected by method",
      x = NULL,
      y = "Number of clusters"
    ) +
    theme_scratch_pub(params$base_size)

  p_counts
  save_plot_safe(p_counts, glue("method_cluster_count_barplot.{params$figure_format}"))
}
```

# Consensus heatmaps
```{r}
#| label: consensus-annotation-heatmap
if (isTRUE(params$show_consensus_annotation_heatmap) && nrow(consensus_annotation_tbl) > 0) {
  top_clusters <- consensus_cluster_summary %>%
    filter(n_cells >= params$min_cluster_size_plot) %>%
    slice_head(n = params$top_n_clusters) %>%
    pull(consensus_cluster)

  mat_df <- consensus_annotation_tbl %>%
    filter(consensus_cluster %in% top_clusters) %>%
    pivot_wider(names_from = annot, values_from = n_cells, values_fill = 0) %>%
    as.data.frame()

  rownames(mat_df) <- mat_df$consensus_cluster
  mat_df$consensus_cluster <- NULL
  mat <- as.matrix(mat_df)

  max_mat <- max(mat, na.rm = TRUE)
  if (!is.finite(max_mat) || max_mat <= 0) max_mat <- 1
  
  ht <- Heatmap(
    mat,
    name = "Cells",
    col = circlize::colorRamp2(
      c(0, max_mat / 2, max_mat),
      c("white", "gold", "firebrick")
    ),
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    row_names_side = "left",
    row_names_gp = grid::gpar(fontsize = 8),
    column_names_gp = grid::gpar(fontsize = 9),
    column_names_rot = 45,
    rect_gp = grid::gpar(col = "grey85"),
    column_title = "Annotation composition of consensus clusters",
    heatmap_legend_param = list(title = "Cells")
  )
    
  ComplexHeatmap::draw(ht)
}

```

# consensus-sample-heatmap
```{r}
#| label: consensus-sample-heatmap
if (isTRUE(params$show_consensus_sample_heatmap) && nrow(consensus_sample_tbl) > 0) {
  top_clusters <- consensus_cluster_summary %>%
    filter(n_cells >= params$min_cluster_size_plot) %>%
    slice_head(n = params$top_n_clusters) %>%
    pull(consensus_cluster)

  mat_df <- consensus_sample_tbl %>%
    filter(consensus_cluster %in% top_clusters) %>%
    pivot_wider(names_from = sample, values_from = n_cells, values_fill = 0) %>%
    as.data.frame()

  rownames(mat_df) <- mat_df$consensus_cluster
  mat_df$consensus_cluster <- NULL
  mat <- as.matrix(mat_df)

  max_mat <- max(mat, na.rm = TRUE)
  if (!is.finite(max_mat) || max_mat <= 0) max_mat <- 1
  
  ht <- Heatmap(
    mat,
    name = "Cells",
    col = circlize::colorRamp2(
      c(0, max_mat / 2, max_mat),
      c("white", "skyblue", "navy")
    ),
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    row_names_side = "left",
    row_names_gp = grid::gpar(fontsize = 8),
    column_names_gp = grid::gpar(fontsize = 8),
    column_names_rot = 45,
    rect_gp = grid::gpar(col = "grey85"),
    column_title = "Sample composition of consensus clusters",
    heatmap_legend_param = list(title = "Cells")
  )
    

  ComplexHeatmap::draw(ht)
}

```

```{r}
#| label: fraction-plots
plot_frac_by_group <- function(df, group_col, title_txt, filename) {
  tmp <- df %>%
    filter(!is.na(.data[[group_col]])) %>%
    group_by(.data[[group_col]]) %>%
    summarise(
      n_cells = n(),
      n_consensus = sum(!is.na(consensus_cluster)),
      frac_consensus = n_consensus / n_cells,
      .groups = "drop"
    ) %>%
    filter(n_cells >= params$min_cells_per_group)

  if (nrow(tmp) == 0) return(NULL)

  p <- tmp %>%
    mutate(.group = fct_reorder(.data[[group_col]], frac_consensus)) %>%
    ggplot(aes(x = .group, y = frac_consensus)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      title = title_txt,
      x = NULL,
      y = "Cells with consensus cluster"
    ) +
    theme_scratch_pub(params$base_size)

  print(p)
  save_plot_safe(p, filename)
  invisible(p)
}

if (isTRUE(params$show_fraction_by_annotation)) {
  plot_frac_by_group(export_cells %>% rename(group_annot = annot), "group_annot",
                     "Fraction of cells with consensus cluster by cell state",
                     glue("consensus_fraction_by_annotation.{params$figure_format}"))
}
if (isTRUE(params$show_fraction_by_sample) && any(!is.na(export_cells$sample))) {
  plot_frac_by_group(export_cells, "sample",
                     "Fraction of cells with consensus cluster by sample",
                     glue("consensus_fraction_by_sample.{params$figure_format}"))
}
if (isTRUE(params$show_fraction_by_condition) && any(!is.na(export_cells$condition))) {
  plot_frac_by_group(export_cells, "condition",
                     "Fraction of cells with consensus cluster by condition",
                     glue("consensus_fraction_by_condition.{params$figure_format}"))
}
if (isTRUE(params$show_fraction_by_patient) && any(!is.na(export_cells$patient))) {
  plot_frac_by_group(export_cells, "patient",
                     "Fraction of cells with consensus cluster by patient",
                     glue("consensus_fraction_by_patient.{params$figure_format}"))
}
```

# Summary tables display
```{r}
#| label: summary-tables-display
if (isTRUE(params$show_summary_tables)) {
  kable(consensus_cluster_summary %>% slice_head(n = params$top_n_clusters),
        caption = "Top consensus clonotype clusters by cell count.") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))

  kable(annotation_consensus_summary %>% slice_head(n = params$top_n_states_heatmap),
        caption = "Top annotations ranked by consensus-clustered fraction.") %>%
    kable_styling(full_width = TRUE, bootstrap_options = c("striped", "hover", "condensed", "responsive"))

  kable(method_presence_summary,
        caption = "Method-level cell coverage summary.") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))
}
```

# Warnings
```{r}
#| label: warnings
warn_tbl <- tibble(
  warning = c(
    "Only one method available",
    "Few cells with consensus cluster",
    "No condition metadata",
    "No patient metadata",
    "Consensus requires >=2 methods",
    "Singleton consensus disabled"
  ),
  triggered = c(
    sum(c(any(export_cells$has_gliph, na.rm = TRUE),
          any(export_cells$has_tcrdist, na.rm = TRUE),
          any(export_cells$has_giana, na.rm = TRUE))) < 2,
    mean(!is.na(export_cells$consensus_cluster)) < 0.3,
    all(is.na(export_cells$condition)),
    all(is.na(export_cells$patient)),
    params$consensus_min_methods > 1,
    !isTRUE(params$assign_singleton_consensus)
  ),
  interpretation = c(
    "Fewer than two clustering methods are available, so this module is acting as a method summary rather than a true consensus comparison.",
    "Less than 30% of cells received a consensus cluster annotation under the current rules.",
    "Condition-level summaries were skipped because condition metadata were unavailable.",
    "Patient-level summaries were skipped because patient metadata were unavailable.",
    "Consensus assignment requires support from at least the specified number of methods.",
    "Cells supported by only one method are excluded from consensus assignment under current settings."
  )
) %>%
  filter(triggered)

if (nrow(warn_tbl) == 0) {
  cat("No major automatic warnings were triggered under the current consensus clustering settings.")
} else {
  kable(warn_tbl %>% select(-triggered), caption = "Automatically generated consensus clustering warnings and notes.") %>%
    kable_styling(full_width = TRUE, bootstrap_options = c("striped", "hover", "condensed"))
}
```

# session info
```{r}
#| label: session-info
writeLines(capture.output(sessionInfo()), file.path(params$outdir, "sessionInfo.consensus_clustering.txt"))
sessionInfo()
```
