---
title: "SCRATCH-TCR: TCRi Report"
author: "Syed Shujaat Ali Zaidi"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    theme: cosmo
    df-print: paged
execute:
  echo: false
  warning: false
  message: false
params:
  # ======================================================
  # Inputs
  # ======================================================
  seurat_rds: "data/seurat_tcells_with_TCR.rds"
  tcri_scores_file: "data/tcri_scores.tsv"
  tcr_export_cells_file: "data/tcr_export_cells_with_embedding.tsv"
  outdir: "TCRi_Report"
  
  python_bin: ""

  # Optional extra inputs
  metadata_file: ""
  previous_summary_file: ""

  # ======================================================
  # Output controls
  # ======================================================
  data_dir: "data"
  tables_dir: "tables"
  figures_dir: "figures"
  save_tables: true
  save_figures: true
  save_updated_seurat: true
  figure_format: "png"
  figure_width: 10
  figure_height: 7
  figure_dpi: 300
  base_size: 12

  # ======================================================
  # Core column mapping
  # ======================================================
  cell_id_col: "cell_id"
  tcri_score_col: "tcri_score"
  tcri_group_col: "tcri_group"
  tcri_rank_col: ""
  tcri_pvalue_col: ""
  tcri_qvalue_col: ""

  cell_id_candidates: !expr c("cell_id","barcode","cell","Cell","cellid")
  tcri_score_candidates: !expr c("tcri_score","score","TCRi","tcri","invertibility_score")
  tcri_group_candidates: !expr c("tcri_group","group","class","label","TCRi_group")
  tcri_rank_candidates: !expr c("rank","tcri_rank","score_rank")
  tcri_pvalue_candidates: !expr c("pvalue","p_value","pval","tcri_pvalue")
  tcri_qvalue_candidates: !expr c("qvalue","q_value","fdr","padj","adj_p","tcri_qvalue")

  # ======================================================
  # Metadata in Seurat / export table
  # ======================================================
  assay_use: "RNA"
  label_col: "predicted_labels"
  sample_col: "META_SAMPLE"
  patient_col: "META_PATIENT"
  condition_col: "META_TIMECOND"
  timepoint_col: "META_TIMEPOINT"
  batch_col: "META_BATCH"
  clone_id_col: "clone_id"
  clone_size_col: "clone_size"
  paired_tcr_col: "paired_tcr"
  has_tcr_col: "has_tcr"

  label_candidates: "predicted_labels;predicted.celltype.l2;predicted.celltype.l1;azimuth_labels;celltypist;celltypist_label;majority_voting;scType;sctype;sctype_label;celltype;CellType;annot;Annotation;seurat_clusters"
  sample_candidates: "META_SAMPLE;Sample;SampleID;sample_id;orig.ident;sample;library;donor;patient"
  patient_candidates: "META_PATIENT;patient_id;patient;Patient;subject;donor;case_id"
  condition_candidates: "META_TIMECOND;condition;Condition;group;Group;status;timepoint;Timepoint"
  timepoint_candidates: "META_TIMEPOINT;timepoint;Timepoint;visit;Visit;day;Day;META_TIMECOND"
  batch_candidates: "META_BATCH;batch;Batch;library;Library;run;Run;lane;Lane"
  clone_id_candidates: "clone_id;CTaa;clonotype;clone"
  clone_size_candidates: "clone_size;CloneSize;clone_n"
  paired_tcr_candidates: "paired_tcr;paired;is_paired"
  has_tcr_candidates: "has_tcr;hasTCR;tcr_positive"
  
  
  # ======================================================
  # TCRi modeling
  # ======================================================
  subset_to_tcr_positive: true
  subset_to_paired_tcr: true
  min_cells_per_phenotype: 15
  min_cells_total: 100
  
  phenotype_obs_name: "phenotype"
  clonotype_obs_name: "clone_id"
  covariate_obs_name: "condition"
  

  
  global_scale: 10.0
  local_scale: 5.0
  prior_temperature: 1.0
  guide_temperature: 1.0
  use_enumeration: false

  margin_scale: 0.0
  margin_value: 2.0
  adaptive_margin: false

  # ======================================================
  # Embedding controls
  # ======================================================
  reduction_use: "umap"
  make_umap_if_missing: false
  umap_dims_max: 30
  umap_nfeatures: 3000
  raster_large_umap: true
  label_clusters: true

  # ======================================================
  # Analysis knobs
  # Score grouping
  # ======================================================
  auto_create_tcri_group_if_missing: true
  tcri_high_cutoff: 0.80
  tcri_mid_cutoff: 0.50
  use_quantile_cutoffs_if_score_not_bounded: true
  high_quantile: 0.90
  mid_quantile: 0.50

  min_cells_per_group: 10
  min_clone_size_for_assoc: 2
  top_n_high_tcri_clones: 30
  top_n_groups_for_ridge: 20
  top_n_states_heatmap: 30
  top_n_states_table: 30

  # ======================================================
  # Figure/report toggles
  # ======================================================
  show_feature_plot: true
  show_group_umap: true
  show_density_plot: true
  show_violin_by_annotation: true
  show_boxplots_by_sample: true
  show_boxplots_by_condition: true
  show_boxplots_by_patient: true
  show_boxplots_by_timepoint: true
  show_score_vs_clone_size: true
  show_high_tcri_clone_heatmap: true
  show_high_tcri_annotation_enrichment: true
  show_summary_tables: true

  report_label: "TCRi"
  # ======================================================
  # TCRi modeling (Optimized)
  # ======================================================
  n_latent: 16            # Increased for better resolution
  n_hidden: 128           # Increased for more "brain power"
  max_epochs: 100         # Essential for convergence
  batch_size: 128         # Better for smaller/medium datasets
  learning_rate: 0.001
  reconstruction_loss_scale: 0.01
  n_steps_kl_warmup: 1000
---

# =================================================================
# * Integrates precomputed TCRi results into the TCR–GEX Seurat object and aligns scores to individual cells through robust cell ID matching.
# * Generates standardized per-cell, per-annotation, and per-sample summaries of TCRi scores, groups, and clone-associated patterns.
# * Supports flexible operation across datasets with optional auto-generation of TCRi groups when categorical labels are not provided.
# * Produces publication-quality visualizations, including embedding overlays, score distributions, annotation-level comparisons, and clone size associations.
# * Identifies high-TCRi clonotypes and summarizes their phenotypic occupancy and enrichment across annotated T-cell states.
# * Exports an updated Seurat object plus downstream-ready tables and figures for subsequent modules such as CoNGA, clonotype clustering, and repertoire analysis.
# =================================================================


## setup
```{r}
#| label: setup
suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(forcats)
  library(scales)
  library(glue)
  library(knitr)
  library(kableExtra)
  library(ComplexHeatmap)
  library(circlize)
  library(patchwork)
  library(Matrix)
  library(reticulate)
})

options(stringsAsFactors = FALSE)

dir.create(params$outdir, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$data_dir), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$tables_dir), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$figures_dir), recursive = TRUE, showWarnings = FALSE)

`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0 && !all(is.na(a))) a else b

# theme_scratch_pub <- function(base_size = 12) {
#   theme_bw(base_size = base_size) +
#     theme(
#       plot.title = element_text(face = "bold", size = base_size + 2, hjust = 0),
#       plot.subtitle = element_text(size = base_size, hjust = 0),
#       axis.title = element_text(face = "bold"),
#       axis.text = element_text(color = "black"),
#       panel.grid.minor = element_blank(),
#       panel.grid.major = element_line(linewidth = 0.2, color = "grey90"),
#       strip.background = element_rect(fill = "grey95", color = "grey80"),
#       strip.text = element_text(face = "bold"),
#       legend.title = element_text(face = "bold"),
#       legend.key = element_blank(),
#       plot.caption = element_text(size = base_size - 2, color = "grey40")
#     )
# }

# theme_scratch_pub <- function(base_size = 12) {
#   theme_bw(base_size = base_size) +
#     theme(
#       plot.title = element_text(face = "bold", size = base_size + 2, hjust = 0),
#       plot.subtitle = element_text(size = base_size, hjust = 0),
#       axis.title = element_text(face = "bold", size = base_size - 1),
#       # Shrink Y-axis labels slightly
#       axis.text.y = element_text(color = "black", size = base_size - 3), 
#       # Rotate X-axis labels 45 degrees and shrink
#       axis.text.x = element_text(color = "black", size = base_size - 3, angle = 45, hjust = 1), 
#       panel.grid.minor = element_blank(),
#       panel.grid.major = element_line(linewidth = 0.1, color = "grey90"),
#       strip.background = element_rect(fill = "grey95", color = "grey80"),
#       strip.text = element_text(face = "bold", size = base_size - 2),
#       legend.title = element_text(face = "bold", size = base_size - 2),
#       legend.text = element_text(size = base_size - 3),
#       plot.caption = element_text(size = base_size - 4, color = "grey40")
#     )
# }


theme_scratch_pub <- function(base_size = 12) {
  theme_bw(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 2, hjust = 0),
      plot.subtitle = element_text(size = base_size, hjust = 0),
      axis.title = element_text(face = "bold", size = base_size - 1),
      axis.text.y = element_text(color = "black", size = base_size - 3),
      # Fix for Q2 & Q3: Smaller text and 45-degree rotation for X-axis
      axis.text.x = element_text(color = "black", size = base_size - 3, angle = 45, hjust = 1),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(linewidth = 0.1, color = "grey90"),
      strip.background = element_rect(fill = "grey95", color = "grey80"),
      strip.text = element_text(face = "bold", size = base_size - 2),
      legend.title = element_text(face = "bold", size = base_size - 2),
      legend.text = element_text(size = base_size - 3),
      legend.key = element_blank(),
      plot.caption = element_text(size = base_size - 4, color = "grey40")
    )
}

save_plot_safe <- function(plot_obj, filename,
                           width = params$figure_width,
                           height = params$figure_height,
                           dpi = params$figure_dpi) {
  if (!isTRUE(params$save_figures)) return(invisible(NULL))
  ggsave(
    filename = file.path(params$outdir, params$figures_dir, filename),
    plot = plot_obj,
    width = width,
    height = height,
    dpi = dpi,
    bg = "white",
    limitsize = FALSE
  )
}

save_table_safe <- function(df, filename) {
  if (!isTRUE(params$save_tables)) return(invisible(NULL))
  fwrite(df, file.path(params$outdir, params$tables_dir, filename), sep = "\t")
}

save_rds_safe <- function(obj, filename) {
  if (!isTRUE(params$save_updated_seurat)) return(invisible(NULL))
  saveRDS(obj, file.path(params$outdir, params$data_dir, filename))
}

first_existing_col <- function(df, preferred, candidates = character()) {
  if (!is.null(preferred) && preferred != "" && preferred %in% colnames(df)) return(preferred)
  hits <- intersect(candidates, colnames(df))
  if (length(hits) > 0) return(hits[[1]])
  NULL
}

safe_percent <- function(x, denom) ifelse(denom > 0, x / denom, NA_real_)

safe_make_umap <- function(seu, reduction_name = "umap", dims = 1:30, nfeatures = 3000) {
  if ("RNA" %in% names(seu@assays)) DefaultAssay(seu) <- "RNA"
  npcs <- min(max(dims), max(2, ncol(seu) - 1), 50)
  nf <- min(nfeatures, nrow(seu))

  if (!"pca" %in% Reductions(seu)) {
    seu <- FindVariableFeatures(seu, nfeatures = nf, verbose = FALSE)
    seu <- ScaleData(seu, verbose = FALSE)
    seu <- RunPCA(seu, npcs = npcs, verbose = FALSE)
  }

  use_dims <- dims[dims <= npcs]
  if (length(use_dims) < 2) use_dims <- 1:min(10, npcs)

  seu <- FindNeighbors(seu, dims = use_dims, verbose = FALSE)
  seu <- RunUMAP(seu, dims = use_dims, reduction.name = reduction_name, verbose = FALSE)
  seu
}

get_expr_mat <- function(seu, assay = "RNA", layer = "data") {
  DefaultAssay(seu) <- assay
  out <- tryCatch(
    SeuratObject::LayerData(seu, assay = assay, layer = layer),
    error = function(e) NULL
  )
  if (is.null(out)) {
    out <- tryCatch(
      Seurat::GetAssayData(seu, assay = assay, slot = "data"),
      error = function(e) NULL
    )
  }
  if (is.null(out)) stop("Could not retrieve expression matrix from assay/layer.")
  out
}

make_tcri_group <- function(score_vec,
                            high_cutoff = 0.80,
                            mid_cutoff = 0.50,
                            use_quantile = TRUE,
                            high_quantile = 0.90,
                            mid_quantile = 0.50) {
  x <- suppressWarnings(as.numeric(score_vec))
  out <- rep(NA_character_, length(x))

  finite_x <- x[is.finite(x)]
  if (length(finite_x) == 0) return(out)

  bounded_01 <- min(finite_x, na.rm = TRUE) >= 0 && max(finite_x, na.rm = TRUE) <= 1

  if (bounded_01 || !use_quantile) {
    hi <- high_cutoff
    mid <- mid_cutoff
  } else {
    hi <- as.numeric(quantile(finite_x, probs = high_quantile, na.rm = TRUE))
    mid <- as.numeric(quantile(finite_x, probs = mid_quantile, na.rm = TRUE))
  }

  dplyr::case_when(
    is.na(x) ~ NA_character_,
    x >= hi ~ "High",
    x >= mid ~ "Intermediate",
    TRUE ~ "Low"
  )
}

compute_group_enrichment <- function(df, group_col, label_col, min_cells = 5) {
  df %>%
    filter(!is.na(.data[[group_col]]), !is.na(.data[[label_col]])) %>%
    count(.data[[group_col]], .data[[label_col]], name = "n") %>%
    group_by(.data[[group_col]]) %>%
    mutate(group_total = sum(n), frac_in_group = n / group_total) %>%
    ungroup() %>%
    group_by(.data[[label_col]]) %>%
    mutate(label_total = sum(n), frac_in_label = n / label_total) %>%
    ungroup() %>%
    filter(n >= min_cells)
}

sanitize_param_string <- function(x) {
  if (is.null(x) || length(x) == 0) return(NULL)
  trimws(gsub("\\u00A0", " ", as.character(x)))
}

normalize_colnames <- function(df) {
  colnames(df) <- trimws(gsub("\\u00A0", " ", colnames(df)))
  df
}

split_candidate_string <- function(x) {
  if (is.null(x) || length(x) == 0 || is.na(x) || x == "") return(character())
  x <- gsub('"', "", as.character(x))
  x <- trimws(unlist(strsplit(x, ";", fixed = TRUE)))
  x[nzchar(x)]
}

first_existing_col <- function(df, preferred, candidates = character()) {
  df <- normalize_colnames(df)
  preferred <- sanitize_param_string(preferred)

  if (length(candidates) == 1 && is.character(candidates)) {
    candidates <- split_candidate_string(candidates)
  }
  candidates <- trimws(gsub("\\u00A0", " ", as.character(candidates)))

  if (!is.null(preferred) && preferred != "" && preferred %in% colnames(df)) return(preferred)

  hits <- intersect(candidates, colnames(df))
  if (length(hits) > 0) return(hits[[1]])

  NULL
}

choose_col <- function(df, primary = NULL, fallback = NULL) {
  nm <- colnames(df)
  if (!is.null(primary) && primary != "" && primary %in% nm) return(primary)
  if (!is.null(fallback) && fallback != "" && fallback %in% nm) return(fallback)
  NULL
}

safe_percent <- function(x, denom) ifelse(denom > 0, x / denom, NA_real_)

# sanitize_param_string <- function(x) {
#   if (is.null(x) || length(x) == 0) return(NULL)
#   trimws(gsub("\\u00A0", " ", as.character(x)))
# }
# 
# normalize_colnames <- function(df) {
#   colnames(df) <- trimws(gsub("\\u00A0", " ", colnames(df)))
#   df
# }
# 
# split_candidate_string <- function(x) {
#   if (is.null(x) || length(x) == 0 || is.na(x) || x == "") return(character())
#   x <- gsub('"', "", as.character(x))
#   x <- trimws(unlist(strsplit(x, ";", fixed = TRUE)))
#   x[nzchar(x)]
# }
# 
# first_existing_col <- function(df, preferred, candidates = character()) {
#   df <- normalize_colnames(df)
#   preferred <- sanitize_param_string(preferred)
# 
#   if (length(candidates) == 1 && is.character(candidates)) {
#     candidates <- split_candidate_string(candidates)
#   }
#   candidates <- trimws(gsub("\\u00A0", " ", as.character(candidates)))
# 
#   if (!is.null(preferred) && preferred != "" && preferred %in% colnames(df)) return(preferred)
# 
#   hits <- intersect(candidates, colnames(df))
#   if (length(hits) > 0) return(hits[[1]])
# 
#   NULL
# }
# 
# choose_col <- function(df, primary = NULL, fallback = NULL) {
#   nm <- colnames(df)
#   if (!is.null(primary) && primary != "" && primary %in% nm) return(primary)
#   if (!is.null(fallback) && fallback != "" && fallback %in% nm) return(fallback)
#   NULL
# }

```

## load-and-resolve
```{r}
#| label: load-and-resolve
stopifnot(file.exists(params$seurat_rds))
seu <- readRDS(params$seurat_rds)
md <- normalize_colnames(seu@meta.data)
seu@meta.data <- md

label_col <- first_existing_col(md, params$label_col, params$label_candidates)
clone_id_col <- first_existing_col(md, params$clone_id_col, params$clone_id_candidates)
clone_size_col <- first_existing_col(md, params$clone_size_col, params$clone_size_candidates)
paired_tcr_col <- first_existing_col(md, params$paired_tcr_col, params$paired_tcr_candidates)
has_tcr_col <- first_existing_col(md, params$has_tcr_col, params$has_tcr_candidates)
sample_col <- first_existing_col(md, params$sample_col, params$sample_candidates)
patient_col <- first_existing_col(md, params$patient_col, params$patient_candidates)
condition_col <- first_existing_col(md, params$condition_col, params$condition_candidates)
timepoint_col <- first_existing_col(md, params$timepoint_col, params$timepoint_candidates)
batch_col <- first_existing_col(md, params$batch_col, params$batch_candidates)

if (is.null(label_col)) stop("Could not resolve annotation label column.")
if (is.null(clone_id_col)) stop("Could not resolve clone_id column.")
if (is.null(clone_size_col)) stop("Could not resolve clone_size column.")
if (is.null(has_tcr_col)) stop("Could not resolve has_tcr column.")
if (is.null(paired_tcr_col)) stop("Could not resolve paired_tcr column.")
if (is.null(sample_col)) stop("Could not resolve sample column.")
if (is.null(patient_col)) stop("Could not resolve patient column.")
if (is.null(condition_col)) stop("Could not resolve condition column.")
if (is.null(timepoint_col)) stop("Could not resolve timepoint column.")

if (!"META_SAMPLE" %in% colnames(seu@meta.data) && !is.null(sample_col)) {
  seu$META_SAMPLE <- as.character(seu@meta.data[[sample_col]])
}
if (!"META_PATIENT" %in% colnames(seu@meta.data) && !is.null(patient_col)) {
  seu$META_PATIENT <- as.character(seu@meta.data[[patient_col]])
}
if (!"META_TIMECOND" %in% colnames(seu@meta.data) && !is.null(condition_col)) {
  seu$META_TIMECOND <- as.character(seu@meta.data[[condition_col]])
}
if (!"META_TIMEPOINT" %in% colnames(seu@meta.data) && !is.null(timepoint_col)) {
  seu$META_TIMEPOINT <- as.character(seu@meta.data[[timepoint_col]])
}
if (!"META_BATCH" %in% colnames(seu@meta.data) && !is.null(batch_col)) {
  seu$META_BATCH <- as.character(seu@meta.data[[batch_col]])
}

print(list(
  resolved_label_col = label_col,
  resolved_clone_id_col = clone_id_col,
  resolved_clone_size_col = clone_size_col,
  resolved_sample_col = sample_col,
  resolved_patient_col = patient_col,
  resolved_condition_col = condition_col,
  resolved_timepoint_col = timepoint_col,
  resolved_batch_col = batch_col
))
# #| label: load-and-resolve
# stopifnot(file.exists(params$seurat_rds))
# seu <- readRDS(params$seurat_rds)
# md <- normalize_colnames(seu@meta.data)
# seu@meta.data <- md
# 
# label_col <- first_existing_col(md, params$label_col, params$label_candidates)
# clone_id_col <- first_existing_col(md, params$clone_id_col, params$clone_id_candidates)
# clone_size_col <- first_existing_col(md, params$clone_size_col, params$clone_size_candidates)
# paired_tcr_col <- first_existing_col(md, params$paired_tcr_col, params$paired_tcr_candidates)
# has_tcr_col <- first_existing_col(md, params$has_tcr_col, params$has_tcr_candidates)
# sample_col <- first_existing_col(md, params$sample_col, params$sample_candidates)
# patient_col <- first_existing_col(md, params$patient_col, params$patient_candidates)
# condition_col <- first_existing_col(md, params$condition_col, params$condition_candidates)
# timepoint_col <- first_existing_col(md, params$timepoint_col, params$timepoint_candidates)
# batch_col <- first_existing_col(md, params$batch_col, params$batch_candidates)
# 
# if (is.null(label_col)) stop("Could not resolve annotation label column.")
# if (is.null(clone_id_col)) stop("Could not resolve clone_id column.")
# if (is.null(has_tcr_col)) stop("Could not resolve has_tcr column.")
# if (is.null(paired_tcr_col)) stop("Could not resolve paired_tcr column.")
# if (is.null(sample_col)) stop("Could not resolve sample column.")
# if (is.null(patient_col)) stop("Could not resolve patient column.")
# if (is.null(condition_col)) stop("Could not resolve condition column.")
# if (is.null(timepoint_col)) stop("Could not resolve timepoint column.")
# 
# if (!"META_SAMPLE" %in% colnames(seu@meta.data) && !is.null(sample_col)) {
#   seu$META_SAMPLE <- as.character(seu@meta.data[[sample_col]])
# }
# if (!"META_PATIENT" %in% colnames(seu@meta.data) && !is.null(patient_col)) {
#   seu$META_PATIENT <- as.character(seu@meta.data[[patient_col]])
# }
# if (!"META_TIMECOND" %in% colnames(seu@meta.data) && !is.null(condition_col)) {
#   seu$META_TIMECOND <- as.character(seu@meta.data[[condition_col]])
# }
# if (!"META_TIMEPOINT" %in% colnames(seu@meta.data) && !is.null(timepoint_col)) {
#   seu$META_TIMEPOINT <- as.character(seu@meta.data[[timepoint_col]])
# }
# if (!"META_BATCH" %in% colnames(seu@meta.data) && !is.null(batch_col)) {
#   seu$META_BATCH <- as.character(seu@meta.data[[batch_col]])
# }
# 
# print(list(
#   resolved_label_col = label_col,
#   resolved_clone_id_col = clone_id_col,
#   resolved_sample_col = sample_col,
#   resolved_patient_col = patient_col,
#   resolved_condition_col = condition_col,
#   resolved_timepoint_col = timepoint_col,
#   resolved_batch_col = batch_col
# ))
```

## subset-and-embeddin
```{r}
#| label: subset-and-embedding
if (isTRUE(params$subset_to_tcr_positive)) {
  keep <- !is.na(seu@meta.data[[has_tcr_col]]) & as.logical(seu@meta.data[[has_tcr_col]])
  seu <- subset(seu, cells = rownames(seu@meta.data)[keep])
}

if (isTRUE(params$subset_to_paired_tcr)) {
  keep <- !is.na(seu@meta.data[[paired_tcr_col]]) & as.logical(seu@meta.data[[paired_tcr_col]])
  seu <- subset(seu, cells = rownames(seu@meta.data)[keep])
}

if (ncol(seu) < params$min_cells_total) {
  stop("Too few cells after subsetting for TCRi. Found: ", ncol(seu))
}

reduction_to_use <- params$reduction_use
if (!(reduction_to_use %in% names(seu@reductions)) && isTRUE(params$make_umap_if_missing)) {
  seu <- safe_make_umap(
    seu,
    reduction_name = params$reduction_use,
    dims = 1:params$umap_dims_max,
    nfeatures = params$umap_nfeatures
  )
}
if (!(reduction_to_use %in% names(seu@reductions))) {
  reduction_to_use <- if ("umap" %in% names(seu@reductions)) "umap" else if ("tsne" %in% names(seu@reductions)) "tsne" else if ("pca" %in% names(seu@reductions)) "pca" else stop("No usable embedding found.")
}
```

## prepare-tcri-inputs
```{r}
#| label: prepare-tcri-inputs
assay_use <- sanitize_param_string(params$assay_use)
layer_use <- sanitize_param_string(params$layer_use)

if (is.null(assay_use) || assay_use == "") {
  stop("params$assay_use is missing or empty.")
}
if (is.null(layer_use) || layer_use == "") {
  layer_use <- "data"
}

message(sprintf("1. Extracting '%s' layer from '%s' assay...", layer_use, assay_use))

available_assays <- Seurat::Assays(seu)
if (!assay_use %in% available_assays) {
  stop(sprintf(
    "Assay '%s' not found in Seurat object. Available assays: %s",
    assay_use, paste(available_assays, collapse = ", ")
  ))
}

if (inherits(seu[[assay_use]], "Assay5")) {
  message("Detected Seurat v5 Assay. Joining layers...")
  seu <- tryCatch({
    SeuratObject::JoinLayers(seu, assay = assay_use)
  }, error = function(e) {
    message("JoinLayers skipped: ", conditionMessage(e))
    seu
  })
}

avail_layers <- tryCatch(
  SeuratObject::Layers(seu[[assay_use]]),
  error = function(e) character(0)
)

avail_layers <- as.character(avail_layers)
avail_layers <- avail_layers[!is.na(avail_layers) & nzchar(avail_layers)]

message("Available layers detected: ", if (length(avail_layers) > 0) paste(avail_layers, collapse = ", ") else "<none detected>")

if (length(avail_layers) > 0 && !(layer_use %in% avail_layers) && ("counts" %in% avail_layers)) {
  message(sprintf(
    "Layer '%s' is missing! Found 'counts' layer. Running LogNormalize on the fly...",
    layer_use
  ))
  seu <- Seurat::NormalizeData(
    seu,
    assay = assay_use,
    normalization.method = "LogNormalize",
    scale.factor = 10000
  )

  avail_layers <- tryCatch(
    SeuratObject::Layers(seu[[assay_use]]),
    error = function(e) character(0)
  )
  avail_layers <- as.character(avail_layers)
  avail_layers <- avail_layers[!is.na(avail_layers) & nzchar(avail_layers)]
}

expr <- tryCatch({
  Seurat::GetAssayData(seu, assay = assay_use, layer = layer_use)
}, error = function(e) {
  tryCatch({
    Seurat::GetAssayData(seu, assay = assay_use, slot = layer_use)
  }, error = function(e2) NULL)
})

if (is.null(expr) || ncol(expr) == 0) {
  avail_layers_txt <- if (length(avail_layers) > 0) paste(avail_layers, collapse = ", ") else "<none detected>"
  stop(sprintf(
    "Matrix extraction failed! The layer/slot '%s' is empty or missing in assay '%s'. Available layers are: %s",
    layer_use, assay_use, avail_layers_txt
  ))
}

if (is.null(colnames(expr)) && ncol(expr) == ncol(seu)) colnames(expr) <- colnames(seu)
if (is.null(rownames(expr)) && nrow(expr) == nrow(seu)) rownames(expr) <- rownames(seu)

expr <- suppressWarnings(as(expr, "dgCMatrix"))

message(sprintf("Extraction successful: %d genes across %d cells.", nrow(expr), ncol(expr)))

message("2. Building Metadata with Python-Safe Hardening...")
md <- normalize_colnames(seu@meta.data)

sample_col_export    <- choose_col(md, "META_SAMPLE", sample_col)
patient_col_export   <- choose_col(md, "META_PATIENT", patient_col)
condition_col_export <- choose_col(md, "META_TIMECOND", condition_col)
timepoint_col_export <- choose_col(md, "META_TIMEPOINT", timepoint_col)
batch_col_export     <- choose_col(md, "META_BATCH", batch_col)

harden_meta <- function(col_name, default_val) {
  if (!is.null(col_name) && col_name %in% colnames(md)) {
    val <- as.character(md[[col_name]])
    val[is.na(val) | trimws(val) == "" | val == "NA" | val == "<NA>"] <- "Unknown"
    return(val)
  } else {
    return(rep(default_val, nrow(md)))
  }
}

obs_df <- data.frame(cell_id = rownames(md), stringsAsFactors = FALSE)

obs_df$phenotype  <- harden_meta(label_col, "Unknown")
obs_df$clone_id   <- harden_meta(clone_id_col, "None")
obs_df$clone_size <- if (!is.null(clone_size_col) && clone_size_col %in% colnames(md)) {
  suppressWarnings(as.numeric(md[[clone_size_col]]))
} else {
  rep(0, nrow(md))
}

obs_df$sample     <- harden_meta(sample_col_export, "Sample1")
obs_df$patient    <- harden_meta(patient_col_export, "Patient1")
obs_df$condition  <- harden_meta(condition_col_export, "Condition1")
obs_df$timepoint  <- harden_meta(timepoint_col_export, "T1")
obs_df$batch      <- harden_meta(batch_col_export, "Batch1")

safe_extract_log <- function(col) {
  if (!is.null(col) && col %in% colnames(md)) {
    as.logical(md[[col]])
  } else {
    rep(FALSE, nrow(md))
  }
}

obs_df$paired_tcr <- safe_extract_log(paired_tcr_col)
obs_df$has_tcr    <- safe_extract_log(has_tcr_col)

obs_df <- obs_df[obs_df$phenotype != "Unknown" & obs_df$clone_id != "None", , drop = FALSE]

message("3. Synchronizing Matrix and Metadata...")
valid_cells <- as.character(intersect(colnames(expr), obs_df$cell_id))

if (length(valid_cells) == 0) {
  stop("Zero cells matched between the expression matrix and the TCR metadata.")
}

expr <- expr[, valid_cells, drop = FALSE]
obs_df <- obs_df[obs_df$cell_id %in% valid_cells, , drop = FALSE]

pheno_counts <- table(obs_df$phenotype)
keep_pheno <- names(pheno_counts)[pheno_counts >= params$min_cells_per_phenotype]
obs_df <- obs_df[obs_df$phenotype %in% keep_pheno, , drop = FALSE]

final_cells <- as.character(obs_df$cell_id)
expr <- expr[, final_cells, drop = FALSE]

if (ncol(expr) < params$min_cells_total) {
  stop("Too few cells remain after phenotype/clonotype filtering for TCRi.")
}

message("4. Exporting files...")
var_df <- data.frame(gene = rownames(expr), row.names = rownames(expr), stringsAsFactors = FALSE)

mm_file   <- file.path(params$outdir, params$data_dir, "tcri_expr.mtx")
obs_file  <- file.path(params$outdir, params$data_dir, "tcri_obs.tsv")
var_file  <- file.path(params$outdir, params$data_dir, "tcri_var.tsv")

Matrix::writeMM(expr, mm_file)
write.table(obs_df, obs_file, sep = "\t", quote = FALSE, row.names = FALSE)
write.table(var_df, var_file, sep = "\t", quote = FALSE, row.names = FALSE)

message("TCRi Inputs successfully prepared. Missing metadata columns were filled with placeholders.")
print(colSums(!is.na(obs_df[, c("sample", "patient", "condition", "timepoint", "batch")])))

# #| label: prepare-tcri-inputs
# assay_use <- params$assay_use
# layer_use <- params$layer_use
# 
# message(sprintf("1. Extracting '%s' layer from '%s' assay...", layer_use, assay_use))
# 
# # Safely check if assay exists
# if (!assay_use %in% Seurat::Assays(seu)) {
#   stop(sprintf("Assay '%s' not found in Seurat object. Available assays: %s", 
#                assay_use, paste(Seurat::Assays(seu), collapse = ", ")))
# }
# 
# # If this is a Seurat v5 object with split layers, join them first
# if (inherits(seu[[assay_use]], "Assay5")) {
#   message("Detected Seurat v5 Assay. Joining layers...")
#   seu <- tryCatch({
#     SeuratObject::JoinLayers(seu, assay = assay_use)
#   }, error = function(e) seu)
# }
# 
# # --- AUTO-NORMALIZATION FAILSAFE ---
# avail_layers <- SeuratObject::Layers(seu[[assay_use]])
# if (!layer_use %in% avail_layers && "counts" %in% avail_layers) {
#   message(sprintf("Layer '%s' is missing! Found 'counts' layer. Running LogNormalize on the fly...", layer_use))
#   seu <- Seurat::NormalizeData(seu, assay = assay_use, normalization.method = "LogNormalize", scale.factor = 10000)
# }
# 
# # Robust Extraction
# expr <- tryCatch({
#   Seurat::GetAssayData(seu, assay = assay_use, layer = layer_use) # v5 argument
# }, error = function(e) {
#   tryCatch({
#     Seurat::GetAssayData(seu, assay = assay_use, slot = layer_use) # v4 argument
#   }, error = function(e2) NULL)
# })
# 
# # Safety Net
# if (is.null(expr) || ncol(expr) == 0) {
#   avail_layers <- paste(SeuratObject::Layers(seu[[assay_use]]), collapse = ", ")
#   stop(sprintf("Matrix extraction failed! The layer/slot '%s' is empty or missing in assay '%s'. Available layers are: %s", 
#                layer_use, assay_use, avail_layers))
# }
# 
# # Attach dimnames only if sizes match perfectly
# if (is.null(colnames(expr)) && ncol(expr) == ncol(seu)) colnames(expr) <- colnames(seu)
# if (is.null(rownames(expr)) && nrow(expr) == nrow(seu)) rownames(expr) <- rownames(seu)
# 
# # Force to pure sparse matrix
# expr <- suppressWarnings(as(expr, "dgCMatrix"))
# 
# message(sprintf("Extraction successful: %d genes across %d cells.", nrow(expr), ncol(expr)))
# 
# message("2. Building Metadata with Python-Safe Hardening...")
# md <- seu@meta.data
# 
# # HELPER FUNCTION: Prevents Python 'ValueError' by ensuring columns are not all NA
# # If column is missing, uses default_val. If column has NAs, replaces with "Unknown".
# harden_meta <- function(col_name, default_val) {
#   if (!is.null(col_name) && col_name %in% colnames(md)) {
#     val <- as.character(md[[col_name]])
#     val[is.na(val) | val == "" | val == "NA"] <- "Unknown"
#     return(val)
#   } else {
#     return(rep(default_val, nrow(md)))
#   }
# }
# 
# obs_df <- data.frame(cell_id = rownames(md), stringsAsFactors = FALSE)
# 
# # Core identifiers (Must be valid)
# obs_df$phenotype  <- harden_meta(label_col, "Unknown")
# obs_df$clone_id   <- harden_meta(clone_id_col, "None")
# obs_df$clone_size <- if(!is.null(clone_size_col) && clone_size_col %in% colnames(md)) suppressWarnings(as.numeric(md[[clone_size_col]])) else 0
# 
# # Covariates and Batch Keys (The focus of the fix)
# obs_df$sample     <- harden_meta(sample_col, "Sample1")
# obs_df$patient    <- harden_meta(patient_col, "Patient1")
# obs_df$condition  <- harden_meta(condition_col, "Condition1")
# obs_df$timepoint  <- harden_meta(timepoint_col, "T1")
# obs_df$batch      <- harden_meta(batch_col, "Batch1")
# 
# # Boolean flags
# safe_extract_log <- function(col) if (!is.null(col) && col %in% colnames(md)) as.logical(md[[col]]) else FALSE
# obs_df$paired_tcr <- safe_extract_log(paired_tcr_col)
# obs_df$has_tcr    <- safe_extract_log(has_tcr_col)
# 
# # Filter for valid cells (Must have phenotype and clone_id)
# obs_df <- obs_df[obs_df$phenotype != "Unknown" & obs_df$clone_id != "None", ]
# 
# message("3. Synchronizing Matrix and Metadata...")
# valid_cells <- as.character(intersect(colnames(expr), obs_df$cell_id))
# 
# if (length(valid_cells) == 0) {
#   stop("Zero cells matched between the expression matrix and the TCR metadata.")
# }
# 
# expr <- expr[, valid_cells, drop = FALSE]
# obs_df <- obs_df[obs_df$cell_id %in% valid_cells, ]
# 
# # Filter rare phenotypes
# pheno_counts <- table(obs_df$phenotype)
# keep_pheno <- names(pheno_counts)[pheno_counts >= params$min_cells_per_phenotype]
# obs_df <- obs_df[obs_df$phenotype %in% keep_pheno, ]
# 
# # Final strict subset
# final_cells <- as.character(obs_df$cell_id)
# expr <- expr[, final_cells, drop = FALSE]
# 
# if (ncol(expr) < params$min_cells_total) {
#   stop("Too few cells remain after phenotype/clonotype filtering for TCRi.")
# }
# 
# message("4. Exporting files...")
# var_df <- data.frame(gene = rownames(expr), row.names = rownames(expr), stringsAsFactors = FALSE)
# 
# mm_file   <- file.path(params$outdir, params$data_dir, "tcri_expr.mtx")
# obs_file  <- file.path(params$outdir, params$data_dir, "tcri_obs.tsv")
# var_file  <- file.path(params$outdir, params$data_dir, "tcri_var.tsv")
# 
# Matrix::writeMM(expr, mm_file)
# write.table(obs_df, obs_file, sep = "\t", quote = FALSE, row.names = FALSE)
# write.table(var_df, var_file, sep = "\t", quote = FALSE, row.names = FALSE)
# 
# message("TCRi Inputs successfully prepared. Missing metadata columns were filled with placeholders.")

```

# run-tcri
```{r}
#| label: run-tcri

python_bin <- params$python_bin
if (is.null(python_bin) || python_bin == "") {
  stop("params$python_bin must point to the working Python executable for TCRi.")
}

scores_file <- file.path(params$outdir, params$tables_dir, "tcri_scores.tsv")
adata_file  <- file.path(params$outdir, params$data_dir, "tcri_model_output.h5ad")
py_script   <- file.path(params$outdir, params$data_dir, "run_tcri.py")
log_file    <- file.path(params$outdir, params$tables_dir, "tcri_python_stdout_stderr.log")

cat("Using python_bin:", python_bin, "\n")

py_lines <- c(
  "import os",
  "os.environ['TQDM_DISABLE'] = '1'",
  "os.environ['PYTHONWARNINGS'] = 'ignore'",
  "import scipy.io",
  "import pandas as pd",
  "import anndata as ad",
  "import numpy as np",
  "import inspect",
  "import tcri",
  "from tcri.model import TCRIModel",
  "import tcri.preprocessing as pp",
  "print('tcri module path:', getattr(tcri, '__file__', 'NA'))",
  "print('tcri attrs sample:', [x for x in dir(tcri) if 'TCRI' in x or x in ['pp','pl','tl']])",
  "",
  sprintf("mm_file = r'''%s'''", mm_file),
  sprintf("obs_file = r'''%s'''", obs_file),
  sprintf("var_file = r'''%s'''", var_file),
  sprintf("scores_file = r'''%s'''", scores_file),
  sprintf("adata_file = r'''%s'''", adata_file),
  "",
  "X = scipy.io.mmread(mm_file).T.tocsr()",
  "obs = pd.read_csv(obs_file, sep='\\t')",
  "var = pd.read_csv(var_file, sep='\\t')",
  "",
  "obs = obs.set_index('cell_id')",
  "var = var.set_index('gene')",
  "",
  "adata = ad.AnnData(X=X, obs=obs, var=var)",
  "adata.obs_names = obs.index.astype(str)",
  "adata.var_names = var.index.astype(str)",
  "",
  "print('adata.obs non-null counts:')",
  "print(adata.obs.notna().sum())",
  "print('timepoint unique values:')",
  "print(adata.obs['timepoint'].dropna().astype(str).unique() if 'timepoint' in adata.obs.columns else 'MISSING')",
  "print('batch unique values:')",
  "print(adata.obs['batch'].dropna().astype(str).unique() if 'batch' in adata.obs.columns else 'MISSING')",
  "",
  "print('TCRIModel.setup_anndata signature:', inspect.signature(TCRIModel.setup_anndata))",
  "setup_sig = inspect.signature(TCRIModel.setup_anndata)",
  "setup_kwargs = {}",
  "",
  "def has_real_values(series):",
  "    s = series.dropna().astype(str).str.strip()",
  "    s = s[~s.isin(['', 'nan', 'NA', '<NA>', 'None'])]",
  "    return len(s) > 0",
  "",
  "if 'phenotype_key' in setup_sig.parameters and 'phenotype' in adata.obs.columns:",
  "    if has_real_values(adata.obs['phenotype']):",
  "        setup_kwargs['phenotype_key'] = 'phenotype'",
  "",
  "if 'labels_key' in setup_sig.parameters and 'phenotype' in adata.obs.columns:",
  "    if has_real_values(adata.obs['phenotype']):",
  "        setup_kwargs['labels_key'] = 'phenotype'",
  "",
  "if 'clonotype_key' in setup_sig.parameters and 'clone_id' in adata.obs.columns:",
  "    if has_real_values(adata.obs['clone_id']):",
  "        setup_kwargs['clonotype_key'] = 'clone_id'",
  "",
  "if 'clone_key' in setup_sig.parameters and 'clone_id' in adata.obs.columns:",
  "    if has_real_values(adata.obs['clone_id']):",
  "        setup_kwargs['clone_key'] = 'clone_id'",
  "",
  "if 'tcr_key' in setup_sig.parameters and 'clone_id' in adata.obs.columns:",
  "    if has_real_values(adata.obs['clone_id']):",
  "        setup_kwargs['tcr_key'] = 'clone_id'",
  "",
  "if 'batch_key' in setup_sig.parameters and 'batch' in adata.obs.columns:",
  "    if has_real_values(adata.obs['batch']):",
  "        setup_kwargs['batch_key'] = 'batch'",
  "",
  "if 'covariate_key' in setup_sig.parameters and 'timepoint' in adata.obs.columns:",
  "    if has_real_values(adata.obs['timepoint']):",
  "        setup_kwargs['covariate_key'] = 'timepoint'",
  "",
  "# use adata.X directly for this installed TCRI version",
  "if 'layer' in setup_sig.parameters:",
  "    setup_kwargs['layer'] = None",
  "",
  "print('setup_kwargs:', setup_kwargs)",
  "TCRIModel.setup_anndata(adata, **setup_kwargs)",
  "print('setup_anndata completed')",
  "",
  sprintf(
    "model = TCRIModel(adata, n_latent=%d, n_hidden=%d, global_scale=%s, local_scale=%s, prior_temperature=%s, guide_temperature=%s, use_enumeration=%s, device=None)",
    as.integer(params$n_latent),
    as.integer(params$n_hidden),
    as.character(params$global_scale),
    as.character(params$local_scale),
    as.character(params$prior_temperature),
    as.character(params$guide_temperature),
    ifelse(isTRUE(params$use_enumeration), "True", "False")
  ),
  "",
  "print('Preparing model.train() arguments...')",
  sprintf("max_epochs = %d", as.integer(params$max_epochs)),
  sprintf("batch_size = %d", as.integer(params$batch_size)),
  sprintf("lr = %s", as.character(params$learning_rate)),
  sprintf("margin_scale = %s", as.character(params$margin_scale)),
  sprintf("margin_value = %s", as.character(params$margin_value)),
  sprintf("adaptive_margin = %s", ifelse(isTRUE(params$adaptive_margin), "True", "False")),
  sprintf("reconstruction_loss_scale = %s", as.character(params$reconstruction_loss_scale)),
  sprintf("n_steps_kl_warmup = %d", as.integer(params$n_steps_kl_warmup)),
  "",
  "train_sig = inspect.signature(model.train)",
  "train_kwargs = {'max_epochs': max_epochs, 'batch_size': batch_size}",
  "",
  "plan_vals = {'lr': lr, 'margin_scale': margin_scale, 'margin_value': margin_value,",
  "             'adaptive_margin': adaptive_margin, 'reconstruction_loss_scale': reconstruction_loss_scale,",
  "             'n_steps_kl_warmup': n_steps_kl_warmup}",
  "",
  "if 'plan_kwargs' in train_sig.parameters:",
  "    train_kwargs['plan_kwargs'] = plan_vals",
  "else:",
  "    for p in plan_vals:",
  "        if p in train_sig.parameters:",
  "            train_kwargs[p] = plan_vals[p]",
  "",
  "print('TCRIModel.train signature:', train_sig)",
  "print('Final train_kwargs:', train_kwargs)",
  "model.train(**train_kwargs)",
  "print('model.train completed')",
  "",
  "# compatibility for installed TCRI preprocessing helpers",
  "if 'clone_id' in adata.obs.columns and 'trb_unique' not in adata.obs.columns:",
  "    adata.obs['trb_unique'] = adata.obs['clone_id'].astype(str)",
  "",
  sprintf(
    "pp.register_model(adata, model, phenotype_prob_slot='X_tcri_phenotypes', phenotype_assignment_obs='tcri_predicted_phenotype', latent_slot='X_tcri', batch_size=%d)",
    as.integer(params$batch_size)
  ),
  "",
  "scores = pd.DataFrame(index=adata.obs.index)",
  "scores['cell_id'] = scores.index.astype(str)",
  "",
  "if 'X_tcri_phenotypes' in adata.obsm.keys():",
  "    probs = np.asarray(adata.obsm['X_tcri_phenotypes'])",
  "    scores['tcri_score'] = probs.max(axis=1)",
  "else:",
  "    scores['tcri_score'] = np.nan",
  "",
  "if 'tcri_predicted_phenotype' in adata.obs.columns:",
  "    scores['tcri_predicted_phenotype'] = adata.obs['tcri_predicted_phenotype'].astype(str)",
  "else:",
  "    scores['tcri_predicted_phenotype'] = np.nan",
  "",
  "if 'X_tcri' in adata.obsm.keys():",
  "    latent = np.asarray(adata.obsm['X_tcri'])",
  "    scores['tcri_latent_norm'] = np.sqrt((latent ** 2).sum(axis=1))",
  "else:",
  "    scores['tcri_latent_norm'] = np.nan",
  "",
  "scores.to_csv(scores_file, sep='\\t', index=False)",
  "",
  sprintf("save_h5ad = %s", ifelse(isTRUE(params$save_h5ad), "True", "False")),
  "if save_h5ad:",
  "    adata.write_h5ad(adata_file)"
)

writeLines(py_lines, py_script)

res <- system2(
  python_bin,
  args = c(py_script),
  stdout = TRUE,
  stderr = TRUE
)

status <- attr(res, "status")
if (is.null(status)) status <- 0L

writeLines(res, log_file)

if (status != 0L) {
  stop(
    paste0(
      "TCRi Python script failed with exit status ", status, ".\n",
      "See log: ", log_file, "\n\n",
      paste(res, collapse = "\n")
    )
  )
} else {
  cat("TCRi Python run completed successfully. Detailed logs were saved to:\n")
  cat(log_file, "\n")
}

# res <- system2(
#   python_bin,
#   args = c(py_script),
#   stdout = TRUE,
#   stderr = TRUE
# )
# 
# status <- attr(res, "status")
# if (is.null(status)) status <- 0L
# 
# writeLines(res, log_file)
# cat(paste(res, collapse = "\n"), "\n")
# 
# if (status != 0L) {
#   stop(
#     paste0(
#       "TCRi Python script failed with exit status ", status, ".\n",
#       "See log: ", log_file, "\n\n",
#       paste(res, collapse = "\n")
#     )
#   )
# }

if (!file.exists(scores_file)) {
  stop(
    paste0(
      "TCRi Python script exited without creating scores_file.\n",
      "Expected: ", scores_file, "\n",
      "See log: ", log_file, "\n\n",
      paste(res, collapse = "\n")
    )
  )
}
```

# merge-back-into-seurat
```{r}
#| label: merge-back-into-seurat
tcri_scores <- fread(scores_file) %>% as.data.frame()

tcri_scores <- tcri_scores %>%
  mutate(
    cell_id = as.character(cell_id),
    tcri_score = suppressWarnings(as.numeric(tcri_score)),
    tcri_predicted_phenotype = as.character(tcri_predicted_phenotype),
    tcri_latent_norm = suppressWarnings(as.numeric(tcri_latent_norm)),
    tcri_group = make_tcri_group(
      tcri_score,
      high_cutoff = params$tcri_high_cutoff,
      mid_cutoff = params$tcri_mid_cutoff,
      use_quantile = params$use_quantile_cutoffs_if_score_not_bounded,
      high_quantile = params$high_quantile,
      mid_quantile = params$mid_quantile
    )
  )

md_dt <- as.data.table(seu@meta.data, keep.rownames = "cell_id")
setkey(md_dt, cell_id)

tcri_dt <- as.data.table(tcri_scores)
setkey(tcri_dt, cell_id)

md_dt <- tcri_dt[md_dt]

rn <- md_dt$cell_id
md_dt$cell_id <- NULL
seu@meta.data <- as.data.frame(md_dt)
rownames(seu@meta.data) <- rn

save_rds_safe(seu, "seurat_tcells_with_TCRi.rds")
save_table_safe(tcri_scores, "tcri_scores.tsv")
```


## build-export
```{r}
#| label: build-export
embed_df <- as.data.frame(Embeddings(seu, reduction_to_use)) %>%
  tibble::rownames_to_column("cell_id")

sample_col_export    <- choose_col(seu@meta.data, "META_SAMPLE", sample_col)
patient_col_export   <- choose_col(seu@meta.data, "META_PATIENT", patient_col)
condition_col_export <- choose_col(seu@meta.data, "META_TIMECOND", condition_col)
timepoint_col_export <- choose_col(seu@meta.data, "META_TIMEPOINT", timepoint_col)
batch_col_export     <- choose_col(seu@meta.data, "META_BATCH", batch_col)

export_cells <- seu@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  transmute(
    cell_id,
    sample = if (!is.null(sample_col_export)) as.character(.data[[sample_col_export]]) else NA_character_,
    patient = if (!is.null(patient_col_export)) as.character(.data[[patient_col_export]]) else NA_character_,
    condition = if (!is.null(condition_col_export)) as.character(.data[[condition_col_export]]) else NA_character_,
    timepoint = if (!is.null(timepoint_col_export)) as.character(.data[[timepoint_col_export]]) else NA_character_,
    batch = if (!is.null(batch_col_export)) as.character(.data[[batch_col_export]]) else NA_character_,
    annot = as.character(.data[[label_col]]),
    clone_id = as.character(.data[[clone_id_col]]),
    clone_size = suppressWarnings(as.numeric(.data[[clone_size_col]])),
    has_tcr = as.logical(.data[[has_tcr_col]]),
    paired_tcr = as.logical(.data[[paired_tcr_col]]),
    tcri_score = suppressWarnings(as.numeric(.data[["tcri_score"]])),
    tcri_group = as.character(.data[["tcri_group"]]),
    tcri_predicted_phenotype = as.character(.data[["tcri_predicted_phenotype"]]),
    tcri_latent_norm = suppressWarnings(as.numeric(.data[["tcri_latent_norm"]]))
  ) %>%
  left_join(embed_df, by = "cell_id")

print(colSums(!is.na(export_cells[, c("sample", "patient", "condition", "timepoint", "batch")])))
save_table_safe(export_cells, "tcri_export_cells.tsv")


# #| label: build-export
# embed_df <- as.data.frame(Embeddings(seu, reduction_to_use)) %>%
#   tibble::rownames_to_column("cell_id")
# 
# sample_col_export    <- choose_col(seu@meta.data, "META_SAMPLE", sample_col)
# patient_col_export   <- choose_col(seu@meta.data, "META_PATIENT", patient_col)
# condition_col_export <- choose_col(seu@meta.data, "META_TIMECOND", condition_col)
# timepoint_col_export <- choose_col(seu@meta.data, "META_TIMEPOINT", timepoint_col)
# batch_col_export     <- choose_col(seu@meta.data, "META_BATCH", batch_col)
# 
# export_cells <- seu@meta.data %>%
#   tibble::rownames_to_column("cell_id") %>%
#   transmute(
#     cell_id,
#     sample = if (!is.null(sample_col_export)) as.character(.data[[sample_col_export]]) else NA_character_,
#     patient = if (!is.null(patient_col_export)) as.character(.data[[patient_col_export]]) else NA_character_,
#     condition = if (!is.null(condition_col_export)) as.character(.data[[condition_col_export]]) else NA_character_,
#     timepoint = if (!is.null(timepoint_col_export)) as.character(.data[[timepoint_col_export]]) else NA_character_,
#     batch = if (!is.null(batch_col_export)) as.character(.data[[batch_col_export]]) else NA_character_,
#     annot = as.character(.data[[label_col]]),
#     clone_id = as.character(.data[[clone_id_col]]),
#     clone_size = suppressWarnings(as.numeric(.data[[clone_size_col]])),
#     has_tcr = as.logical(.data[[has_tcr_col]]),
#     paired_tcr = as.logical(.data[[paired_tcr_col]]),
#     tcri_score = suppressWarnings(as.numeric(.data[["tcri_score"]])),
#     tcri_group = as.character(.data[["tcri_group"]]),
#     tcri_predicted_phenotype = as.character(.data[["tcri_predicted_phenotype"]]),
#     tcri_latent_norm = suppressWarnings(as.numeric(.data[["tcri_latent_norm"]]))
#   ) %>%
#   left_join(embed_df, by = "cell_id")
# 
# print(colSums(!is.na(export_cells[, c("sample", "patient", "condition", "timepoint")])))
# save_table_safe(export_cells, "tcri_export_cells.tsv")

```


## summary tables
```{r}
#| label: summary-tables
summary_rollup <- tibble(
  metric = c(
    "Report label",
    "Cells modeled",
    "Cells with TCRi score",
    "Fraction with TCRi score",
    "TCR-positive cells",
    "Paired TCR cells",
    "Unique annotations",
    "Unique clones",
    "Selected embedding"
  ),
  value = c(
    params$report_label,
    nrow(export_cells),
    sum(!is.na(export_cells$tcri_score)),
    sprintf("%.2f%%", 100 * mean(!is.na(export_cells$tcri_score))),
    sum(export_cells$has_tcr, na.rm = TRUE),
    sum(export_cells$paired_tcr, na.rm = TRUE),
    length(unique(na.omit(export_cells$annot))),
    length(unique(na.omit(export_cells$clone_id))),
    reduction_to_use
  )
)
save_table_safe(summary_rollup, "tcri_summary_rollup.tsv")

tcri_group_summary <- export_cells %>%
  filter(!is.na(tcri_group)) %>%
  count(tcri_group, name = "n_cells") %>%
  mutate(frac = n_cells / sum(n_cells))
save_table_safe(tcri_group_summary, "tcri_group_summary.tsv")

annotation_tcri_summary <- export_cells %>%
  filter(!is.na(tcri_score), !is.na(annot)) %>%
  group_by(annot) %>%
  summarise(
    n_cells = n(),
    mean_tcri = mean(tcri_score, na.rm = TRUE),
    median_tcri = median(tcri_score, na.rm = TRUE),
    sd_tcri = sd(tcri_score, na.rm = TRUE),
    n_high = sum(tcri_group == "High", na.rm = TRUE),
    frac_high = n_high / n_cells,
    .groups = "drop"
  ) %>%
  arrange(desc(mean_tcri))
save_table_safe(annotation_tcri_summary, "annotation_tcri_summary.tsv")

sample_tcri_summary <- export_cells %>%
  filter(!is.na(tcri_score), !is.na(sample)) %>%
  group_by(sample) %>%
  summarise(
    n_cells = n(),
    mean_tcri = mean(tcri_score, na.rm = TRUE),
    median_tcri = median(tcri_score, na.rm = TRUE),
    n_high = sum(tcri_group == "High", na.rm = TRUE),
    frac_high = n_high / n_cells,
    .groups = "drop"
  ) %>%
  arrange(desc(mean_tcri))
save_table_safe(sample_tcri_summary, "sample_tcri_summary.tsv")

if (any(!is.na(export_cells$condition))) {
  condition_tcri_summary <- export_cells %>%
    filter(!is.na(tcri_score), !is.na(condition)) %>%
    group_by(condition) %>%
    summarise(
      n_cells = n(),
      mean_tcri = mean(tcri_score, na.rm = TRUE),
      median_tcri = median(tcri_score, na.rm = TRUE),
      n_high = sum(tcri_group == "High", na.rm = TRUE),
      frac_high = n_high / n_cells,
      .groups = "drop"
    )
  save_table_safe(condition_tcri_summary, "condition_tcri_summary.tsv")
}

if (any(!is.na(export_cells$patient))) {
  patient_tcri_summary <- export_cells %>%
    filter(!is.na(tcri_score), !is.na(patient)) %>%
    group_by(patient) %>%
    summarise(
      n_cells = n(),
      mean_tcri = mean(tcri_score, na.rm = TRUE),
      median_tcri = median(tcri_score, na.rm = TRUE),
      n_high = sum(tcri_group == "High", na.rm = TRUE),
      frac_high = n_high / n_cells,
      .groups = "drop"
    )
  save_table_safe(patient_tcri_summary, "patient_tcri_summary.tsv")
}

if (any(!is.na(export_cells$timepoint))) {
  timepoint_tcri_summary <- export_cells %>%
    filter(!is.na(tcri_score), !is.na(timepoint)) %>%
    group_by(timepoint) %>%
    summarise(
      n_cells = n(),
      mean_tcri = mean(tcri_score, na.rm = TRUE),
      median_tcri = median(tcri_score, na.rm = TRUE),
      n_high = sum(tcri_group == "High", na.rm = TRUE),
      frac_high = n_high / n_cells,
      .groups = "drop"
    )
  save_table_safe(timepoint_tcri_summary, "timepoint_tcri_summary.tsv")
}

high_tcri_clone_summary <- export_cells %>%
  filter(!is.na(clone_id), clone_id != "", !is.na(tcri_group), clone_size >= params$min_clone_size_for_assoc) %>%
  group_by(clone_id, clone_size) %>%
  summarise(
    n_cells = n(),
    n_high = sum(tcri_group == "High", na.rm = TRUE),
    frac_high = n_high / n_cells,
    mean_tcri = mean(tcri_score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(frac_high), desc(mean_tcri), desc(clone_size)) %>%
  slice_head(n = params$top_n_high_tcri_clones)
save_table_safe(high_tcri_clone_summary, "high_tcri_clone_summary.tsv")

group_annotation_enrichment <- compute_group_enrichment(
  export_cells,
  group_col = "tcri_group",
  label_col = "annot",
  min_cells = params$min_cells_per_group
)
save_table_safe(group_annotation_enrichment, "tcri_group_annotation_enrichment.tsv")
```

## overview
```{r}
#| label: overview
kable(summary_rollup, caption = "High-level overview of TCRi analysis.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))
```

# TCRi score distributions
```{r}
#| label: density-plot
if (isTRUE(params$show_density_plot)) {
  p_density <- export_cells %>%
    filter(!is.na(tcri_score)) %>%
    ggplot(aes(x = tcri_score)) +
    geom_density(fill = "steelblue", alpha = 0.5, linewidth = 0.8) +
    labs(
      title = "Distribution of TCRi scores",
      subtitle = "Per-cell TCRi score defined as max predicted phenotype probability.",
      x = "TCRi score",
      y = "Density"
    ) +
    theme_scratch_pub(params$base_size)

  p_density
  save_plot_safe(p_density, glue("tcri_score_density.{params$figure_format}"))
}
```

## tcri-feature-plot
```{r}
#| label: feature-plot
if (isTRUE(params$show_feature_plot)) {
  p_feature <- FeaturePlot(
    seu,
    reduction = reduction_to_use,
    features = "tcri_score",
    raster = isTRUE(params$raster_large_umap)
  ) +
    ggtitle("Embedding overlay: TCRi score") +
    theme_scratch_pub(params$base_size)

  p_feature
  save_plot_safe(p_feature, glue("tcri_feature_plot.{params$figure_format}"), width = 8, height = 6)
}
```

# tcri-group-umap
```{r}
#| label: group-umap
if (isTRUE(params$show_group_umap)) {
  p_group <- DimPlot(
    seu,
    reduction = reduction_to_use,
    group.by = "tcri_group",
    label = FALSE,
    raster = isTRUE(params$raster_large_umap)
  ) +
    ggtitle("Embedding overlay: TCRi group") +
    theme_scratch_pub(params$base_size)

  p_group
  save_plot_safe(p_group, glue("tcri_group_umap.{params$figure_format}"), width = 8, height = 6)
}
```

# TCRi by annotation
```{r}
#| label: violin-by-annotation
if (isTRUE(params$show_violin_by_annotation)) {
  annot_order <- annotation_tcri_summary %>%
    arrange(desc(mean_tcri)) %>%
    pull(annot)

  p_violin_annot <- export_cells %>%
    filter(!is.na(tcri_score), !is.na(annot)) %>%
    mutate(annot = factor(annot, levels = annot_order)) %>%
    ggplot(aes(x = annot, y = tcri_score, fill = annot)) +
    geom_violin(scale = "width", trim = TRUE, alpha = 0.8) +
    geom_boxplot(width = 0.12, outlier.size = 0.2, fill = "white") +
    coord_flip() +
    guides(fill = "none") +
    labs(
      title = "TCRi score by annotation",
      subtitle = "Violin + boxplot summary across annotated T-cell states.",
      x = NULL,
      y = "TCRi score"
    ) +
    theme_scratch_pub(params$base_size)

  p_violin_annot
  save_plot_safe(p_violin_annot, glue("tcri_violin_by_annotation.{params$figure_format}"), width = 10, height = 8)
}

```

# boxplots-by-group TCRi by sample / condition / patient / timepoint
```{r}
#| label: boxplots-by-group
plot_box_by_group <- function(df, group_col, title_txt, filename) {
  min_cells_use <- if (!is.null(params$min_cells_per_group) && length(params$min_cells_per_group) == 1) {
    params$min_cells_per_group
  } else {
    10
  }

  tmp <- df %>%
    filter(!is.na(tcri_score), !is.na(.data[[group_col]])) %>%
    group_by(.data[[group_col]]) %>%
    mutate(group_n = n()) %>%
    ungroup() %>%
    filter(group_n >= min_cells_use)

  if (nrow(tmp) == 0) return(NULL)

  group_summary <- tmp %>%
    group_by(.data[[group_col]]) %>%
    summarise(mean_tcri = mean(tcri_score, na.rm = TRUE), .groups = "drop")

  group_levels <- group_summary %>%
    arrange(desc(mean_tcri)) %>%
    pull(.data[[group_col]])

  p <- tmp %>%
    mutate(.group = factor(.data[[group_col]], levels = group_levels)) %>%
    ggplot(aes(x = .group, y = tcri_score, fill = .group)) +
    geom_boxplot(outlier.size = 0.3, alpha = 0.85) +
    coord_flip() +
    guides(fill = "none") +
    labs(
      title = title_txt,
      x = NULL,
      y = "TCRi score"
    ) +
    theme_scratch_pub(params$base_size)

  print(p)
  save_plot_safe(p, filename)
  invisible(p)
}


if (isTRUE(params$show_boxplots_by_sample) && any(!is.na(export_cells$sample))) {
  plot_box_by_group(export_cells, "sample", "TCRi score by sample", glue("tcri_boxplot_by_sample.{params$figure_format}"))
}
if (isTRUE(params$show_boxplots_by_condition) && any(!is.na(export_cells$condition))) {
  plot_box_by_group(export_cells, "condition", "TCRi score by condition", glue("tcri_boxplot_by_condition.{params$figure_format}"))
}
if (isTRUE(params$show_boxplots_by_patient) && any(!is.na(export_cells$patient))) {
  plot_box_by_group(export_cells, "patient", "TCRi score by patient", glue("tcri_boxplot_by_patient.{params$figure_format}"))
}
if (isTRUE(params$show_boxplots_by_timepoint) && any(!is.na(export_cells$timepoint))) {
  plot_box_by_group(export_cells, "timepoint", "TCRi score by timepoint", glue("tcri_boxplot_by_timepoint.{params$figure_format}"))
}
```

## TCRi and clone size
```{r}
#| label: score-vs-clone-size
if (isTRUE(params$show_score_vs_clone_size)) {
  p_clone_assoc <- export_cells %>%
    filter(!is.na(tcri_score), !is.na(clone_size), clone_size > 0) %>%
    ggplot(aes(x = clone_size, y = tcri_score)) +
    geom_point(alpha = 0.35, size = 1.1) +
    scale_x_log10() +
    geom_smooth(method = "lm", se = TRUE, color = "firebrick") +
    labs(
      title = "Association between clone size and TCRi score",
      subtitle = "Clone size shown on log10 scale.",
      x = "Clone size (log10 scale)",
      y = "TCRi score"
    ) +
    theme_scratch_pub(params$base_size)

  p_clone_assoc
  save_plot_safe(p_clone_assoc, glue("tcri_vs_clone_size.{params$figure_format}"))
}
```

# High-TCRi clone summaries
```{r}
#| label: high-tcri-clone-heatmap

if (isTRUE(params$show_high_tcri_clone_heatmap) && nrow(high_tcri_clone_summary) > 0) {

  top_clone_ids <- high_tcri_clone_summary$clone_id

  mat_df <- export_cells %>%
    filter(clone_id %in% top_clone_ids, !is.na(annot), annot != "") %>%
    count(clone_id, annot, name = "n_cells") %>%
    tidyr::pivot_wider(names_from = annot, values_from = n_cells, values_fill = 0) %>%
    as.data.frame()

  if (nrow(mat_df) > 0 && ncol(mat_df) > 1) {
    rownames(mat_df) <- mat_df$clone_id
    mat_df$clone_id <- NULL
    mat <- as.matrix(mat_df)

    # keep row order consistent with high_tcri_clone_summary where possible
    row_order_keep <- intersect(top_clone_ids, rownames(mat))
    mat <- mat[row_order_keep, , drop = FALSE]

    max_mat <- max(mat, na.rm = TRUE)
    if (!is.finite(max_mat) || max_mat <= 0) max_mat <- 1

    ht <- ComplexHeatmap::Heatmap(
      mat,
      name = "Cells",
      col = circlize::colorRamp2(
        c(0, max_mat / 2, max_mat),
        c("white", "gold", "firebrick")
      ),
      cluster_rows = TRUE,
      cluster_columns = TRUE,
      row_names_side = "left",
      row_names_gp = grid::gpar(fontsize = 6),
      column_names_gp = grid::gpar(fontsize = 9),
      column_names_rot = 45,
      column_title = "Annotation occupancy of top high-TCRi clonotypes",
      heatmap_legend_param = list(title = "Cells"),
      width = grid::unit(max(5, ncol(mat) * 0.8), "in"),
      height = grid::unit(max(5, nrow(mat) * 0.18), "in")
    )

    ComplexHeatmap::draw(
      ht,
      heatmap_legend_side = "right"
    )
  } else {
    cat("High-TCRi clone heatmap skipped: insufficient clone-by-annotation matrix after filtering.")
  }
}

# if (isTRUE(params$show_high_tcri_clone_heatmap) && nrow(high_tcri_clone_summary) > 0) {
#   top_clone_ids <- high_tcri_clone_summary$clone_id
# 
#   mat_df <- export_cells %>%
#     filter(clone_id %in% top_clone_ids, !is.na(annot)) %>%
#     count(clone_id, annot, name = "n_cells") %>%
#     tidyr::pivot_wider(names_from = annot, values_from = n_cells, values_fill = 0) %>%
#     as.data.frame()
# 
#   rownames(mat_df) <- mat_df$clone_id
#   mat_df$clone_id <- NULL
#   mat <- as.matrix(mat_df)
# 
#   ht <- Heatmap(
#     mat,
#     name = "Cells",
#     col = colorRamp2(
#       c(0, max(mat, na.rm = TRUE) / 2, max(mat, na.rm = TRUE)),
#       c("white", "gold", "firebrick")
#     ),
#     cluster_rows = TRUE,
#     cluster_columns = TRUE,
#     row_names_side = "left",
#     column_title = "Annotation occupancy of top high-TCRi clonotypes",
#     heatmap_legend_param = list(title = "Cells")
#   )
# 
#   draw(ht)
# }
```

## TCRi group enrichment across annotations
```{r}
#| label: annotation-enrichment
if (isTRUE(params$show_annotation_enrichment) && nrow(group_annotation_enrichment) > 0) {
  enrich_plot_df <- group_annotation_enrichment %>%
    filter(!is.na(tcri_group), !is.na(annot)) %>%
    mutate(annot = fct_reorder(annot, frac_in_group, .fun = max))

  p_enrich <- ggplot(enrich_plot_df, aes(x = annot, y = frac_in_group, fill = tcri_group)) +
    geom_col(position = "dodge") +
    coord_flip() +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      title = "Annotation enrichment across TCRi groups",
      subtitle = "Fraction of each TCRi group represented by each annotation.",
      x = NULL,
      y = "Fraction within TCRi group",
      fill = "TCRi group"
    ) +
    theme_scratch_pub(params$base_size)

  p_enrich
  save_plot_safe(p_enrich, glue("tcri_group_annotation_enrichment.{params$figure_format}"), width = 10, height = 8)
}
```

# Summary tables
```{r}
#| label: summary-tables-display
kable(tcri_group_summary, caption = "TCRi group composition.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))

kable(annotation_tcri_summary %>% slice_head(n = params$top_n_states_table),
      caption = "Top annotations ranked by mean TCRi score.") %>%
  kable_styling(full_width = TRUE, bootstrap_options = c("striped", "hover", "condensed", "responsive"))

if (nrow(high_tcri_clone_summary) > 0) {
  kable(high_tcri_clone_summary,
        caption = "Top clonotypes enriched for high TCRi scores.") %>%
    kable_styling(full_width = TRUE, bootstrap_options = c("striped", "hover", "condensed", "responsive"))
}
```

# warnings
```{r}
#| label: warnings
condition_missing <- is.null(condition_col_export) || all(is.na(export_cells$condition))
patient_missing   <- is.null(patient_col_export)   || all(is.na(export_cells$patient))
timepoint_missing <- is.null(timepoint_col_export) || all(is.na(export_cells$timepoint))

warn_tbl <- tibble(
  warning = c(
    "Few cells with TCRi scores",
    "Low paired TCR fraction among modeled cells",
    "Few phenotype classes after filtering",
    "Condition metadata unresolved or empty after export",
    "Patient metadata unresolved or empty after export",
    "Timepoint metadata unresolved or empty after export"
  ),
  triggered = c(
    mean(!is.na(export_cells$tcri_score)) < 0.5,
    mean(export_cells$paired_tcr, na.rm = TRUE) < 0.3,
    length(unique(na.omit(export_cells$annot))) < 3,
    condition_missing,
    patient_missing,
    timepoint_missing
  ),
  interpretation = c(
    "Less than half the modeled cells received a TCRi score. Verify Python TCRi execution and cell alignment.",
    "Among modeled cells, paired TCR fraction is low. Interpretation should consider incomplete receptor pairing.",
    "Very few phenotype classes remained after filtering. Lower min_cells_per_phenotype if biologically appropriate.",
    "Condition-level summaries were skipped because condition metadata were not resolved or retained correctly in the exported TCRi table.",
    "Patient-level summaries were skipped because patient metadata were not resolved or retained correctly in the exported TCRi table.",
    "Timepoint-level summaries were skipped because timepoint metadata were not resolved or retained correctly in the exported TCRi table."
  )
) %>%
  filter(triggered)

if (nrow(warn_tbl) == 0) {
  cat("No major automatic warnings were triggered under the current TCRi settings.")
} else {
  print(
    kable(
      warn_tbl %>% select(-triggered),
      caption = "Automatically generated TCRi warnings and notes."
    ) %>%
      kable_styling(
        full_width = TRUE,
        bootstrap_options = c("striped", "hover", "condensed")
      )
  )
}

```

# session-info
```{r}
#| label: session-info

dir.create(file.path(params$outdir, params$data_dir), recursive = TRUE, showWarnings = FALSE)
saveRDS(seu, file.path(params$outdir, params$data_dir, "seurat_with_TCRi.rds"))


writeLines(capture.output(sessionInfo()), file.path(params$outdir, "sessionInfo.tcri.txt"))
sessionInfo()
```
