---
title: "SCRATCH-TCR: TCRdist3 Analysis"
author: "Syed Shujaat Ali Zaidi"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    theme: cosmo
    df-print: paged
execute:
  echo: false
  warning: false
  message: false

params:
  # ======================================================
  # Inputs
  # ======================================================
  seurat_rds: "data/seurat_tcells_with_TCR.rds"
  tcr_export_cells_file: "tables/tcr_export_cells_with_embedding.tsv"
  outdir: "TCRdist3_Report"

  # Optional precomputed inputs
  tcrdist_clusters_file: ""
  tcrdist_neighbors_file: ""
  tcrdist_distance_file: ""
  organism: "human"
  db_file: "alphabeta_gammadelta_db.tsv"
  tcrdist_radius: 50
  python_bin: "/opt/conda/envs/tcrenv/bin/python"

  # ======================================================
  # Output controls
  # ======================================================
  data_dir: "data"
  tables_dir: "tables"
  figures_dir: "figures"
  save_tables: true
  save_figures: true
  save_updated_seurat: true
  figure_format: "png"
  figure_width: 10
  figure_height: 7
  figure_dpi: 300
  base_size: 12

  # ======================================================
  # Metadata mapping
  # ======================================================
  label_col: ""
  sample_col: "META_SAMPLE"
  patient_col: "META_PATIENT"
  condition_col: "META_TIMECOND"
  timepoint_col: "META_TIMEPOINT"
  batch_col: "META_BATCH"
  clone_id_col: "clone_id"
  clone_size_col: "clone_size"
  paired_tcr_col: "paired_tcr"
  has_tcr_col: "has_tcr"

  clone_match_mode: "cdr3b_from_CTaa"
  derive_clone_key_from_CTaa: true

  label_candidates: "predicted_labels;predicted.celltype.l2;predicted.celltype.l1;azimuth_labels;celltypist;celltypist_label;majority_voting;scType;sctype;sctype_label;celltype;CellType;annot;Annotation;seurat_clusters"
  sample_candidates: "META_SAMPLE;orig.ident;sample;sample_id;Sample;SampleID"
  patient_candidates: "META_PATIENT;patient_id;patient;Patient;subject;donor;case_id"
  condition_candidates: "META_TIMECOND;condition;Condition;group;Group;status"
  timepoint_candidates: "META_TIMEPOINT;timepoint;Timepoint;visit;Visit;day;Day;META_TIMECOND"
  batch_candidates: "META_BATCH;batch;Batch;library;Library;run;Run;lane;Lane"
  clone_id_candidates: "clone_id;CTaa;clonotype;clone"
  clone_size_candidates: "clone_size;CloneSize;clone_n"
  paired_tcr_candidates: "paired_tcr;paired;is_paired"
  has_tcr_candidates: "has_tcr;hasTCR;tcr_positive"

  # ======================================================
  # Embedding controls
  # ======================================================
  reduction_use: "umap"
  make_umap_if_missing: true
  umap_dims_max: 30
  umap_nfeatures: 3000
  raster_large_umap: true

  # ======================================================
  # Analysis controls
  # ======================================================
  subset_to_tcr_positive: true
  subset_to_paired_tcr: true
  min_cells_total: 100
  
  min_cells_per_group: 10
  min_cluster_size_plot: 3
  top_n_clusters: 20
  top_n_neighbors: 50
  top_n_states_heatmap: 20

  show_cluster_umap: true
  show_neighbor_graph: true
  show_cluster_size_barplot: true
  show_annotation_heatmap: true
  show_sample_heatmap: true
  show_summary_tables: true
  
  


  report_label: "TCRdist3 Analysis"
---


# setup
```{r}
#| label: setup
suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(forcats)
  library(scales)
  library(glue)
  library(knitr)
  library(kableExtra)
  library(ComplexHeatmap)
  library(circlize)
  library(patchwork)
  library(reticulate)
})

reticulate::use_python("/opt/conda/envs/tcrenv/bin/python", required = TRUE)

options(stringsAsFactors = FALSE)

dir.create(params$outdir, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$data_dir), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$tables_dir), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$figures_dir), recursive = TRUE, showWarnings = FALSE)

`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0 && !all(is.na(a))) a else b

Sys.setenv(RETICULATE_PYTHON = "/opt/conda/envs/tcrenv/bin/python")
Sys.setenv(LD_LIBRARY_PATH = paste(
  "/opt/conda/envs/tcrenv/lib",
  Sys.getenv("LD_LIBRARY_PATH"),
  sep = ":"
))


theme_scratch_pub <- function(base_size = 12) {
  theme_bw(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 2, hjust = 0),
      plot.subtitle = element_text(size = base_size, hjust = 0),
      axis.title = element_text(face = "bold"),
      axis.text = element_text(color = "black"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(linewidth = 0.2, color = "grey90"),
      strip.background = element_rect(fill = "grey95", color = "grey80"),
      strip.text = element_text(face = "bold"),
      legend.title = element_text(face = "bold"),
      legend.key = element_blank(),
      plot.caption = element_text(size = base_size - 2, color = "grey40")
    )
}

save_plot_safe <- function(plot_obj, filename,
                           width = params$figure_width,
                           height = params$figure_height,
                           dpi = params$figure_dpi) {
  if (!isTRUE(params$save_figures)) return(invisible(NULL))
  ggsave(
    filename = file.path(params$outdir, params$figures_dir, filename),
    plot = plot_obj,
    width = width,
    height = height,
    dpi = dpi,
    bg = "white",
    limitsize = FALSE
  )
}

save_table_safe <- function(df, filename) {
  if (!isTRUE(params$save_tables)) return(invisible(NULL))
  fwrite(df, file.path(params$outdir, params$tables_dir, filename), sep = "\t")
}

save_rds_safe <- function(obj, filename) {
  if (!isTRUE(params$save_updated_seurat)) return(invisible(NULL))
  saveRDS(obj, file.path(params$outdir, params$data_dir, filename))
}

require_file <- function(path, msg = NULL) {
  if (!file.exists(path)) stop(msg %||% paste("Missing required file:", path))
  path
}

first_existing_col <- function(df, preferred, candidates = character()) {
  if (!is.null(preferred) && preferred != "" && preferred %in% colnames(df)) return(preferred)
  hits <- intersect(candidates, colnames(df))
  if (length(hits) > 0) return(hits[[1]])
  NULL
}

safe_make_umap <- function(seu, reduction_name = "umap", dims = 1:30, nfeatures = 3000, seed = 1234) {
  if ("RNA" %in% names(seu@assays)) DefaultAssay(seu) <- "RNA"
  npcs <- min(max(dims), max(2, ncol(seu) - 1), 50)
  nf <- min(nfeatures, nrow(seu))

  if (!"pca" %in% Reductions(seu)) {
    seu <- FindVariableFeatures(seu, nfeatures = nf, verbose = FALSE)
    seu <- ScaleData(seu, verbose = FALSE)
    seu <- RunPCA(seu, npcs = npcs, verbose = FALSE)
  }

  use_dims <- dims[dims <= npcs]
  if (length(use_dims) < 2) use_dims <- 1:min(10, npcs)

  seu <- FindNeighbors(seu, dims = use_dims, verbose = FALSE)
  seu <- RunUMAP(seu, dims = use_dims, reduction.name = reduction_name, verbose = FALSE)
  seu
}

extract_cdr3b_from_CTaa <- function(x) {
  x <- as.character(x)
  out <- rep(NA_character_, length(x))
  has_b <- grepl("\\|B:", x)
  out[has_b] <- sub("^.*\\|B:", "", x[has_b])
  only_b <- grepl("^B:", x)
  out[only_b] <- sub("^B:", "", x[only_b])
  out
}

split_candidate_string <- function(x) {
  if (is.null(x) || length(x) == 0 || is.na(x) || x == "") return(character())
  x <- gsub('"', "", as.character(x))
  x <- trimws(unlist(strsplit(x, ";", fixed = TRUE)))
  x[nzchar(x)]
}

safe_percent <- function(x, denom) {
  ifelse(denom > 0, x / denom, 0)
}

# MD Anderson Publication Theme
theme_scratch_pub <- function(base_size = 12) {
  theme_bw(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 2),
      panel.grid.major = element_line(linewidth = 0.2, color = "grey90"),
      strip.background = element_rect(fill = "grey95")
    )
}

# sanitize_param_string <- function(x) {
#   if (is.null(x) || length(x) == 0) return(NULL)
#   trimws(gsub("\\u00A0", " ", as.character(x)))
# }

# first_existing_col <- function(df, preferred, candidates = character()) {
#   preferred <- sanitize_param_string(preferred)
#   if (length(candidates) == 1 && is.character(candidates)) {
#     candidates <- split_candidate_string(candidates)
#   }
#   candidates <- trimws(gsub("\\u00A0", " ", as.character(candidates)))
# 
#   if (!is.null(preferred) && preferred != "" && preferred %in% colnames(df)) return(preferred)
#   hits <- intersect(candidates, colnames(df))
#   if (length(hits) > 0) return(hits[[1]])
#   NULL
# }

sanitize_param_string <- function(x) {
  if (is.null(x) || length(x) == 0) return(NULL)
  trimws(gsub("\\u00A0", " ", as.character(x)))
}

normalize_colnames <- function(df) {
  colnames(df) <- trimws(gsub("\\u00A0", " ", colnames(df)))
  df
}

split_candidate_string <- function(x) {
  if (is.null(x) || length(x) == 0 || is.na(x) || x == "") return(character())
  x <- gsub('"', "", as.character(x))
  x <- trimws(unlist(strsplit(x, ";", fixed = TRUE)))
  x[nzchar(x)]
}

first_existing_col <- function(df, preferred, candidates = character()) {
  df <- normalize_colnames(df)
  preferred <- sanitize_param_string(preferred)

  if (length(candidates) == 1 && is.character(candidates)) {
    candidates <- split_candidate_string(candidates)
  }
  candidates <- trimws(gsub("\\u00A0", " ", as.character(candidates)))

  if (!is.null(preferred) && preferred != "" && preferred %in% colnames(df)) return(preferred)

  hits <- intersect(candidates, colnames(df))
  if (length(hits) > 0) return(hits[[1]])

  NULL
}

```

# read-inputs
```{r}
#| label: read-inputs
require_file(params$seurat_rds, sprintf("Seurat RDS not found: %s", params$seurat_rds))
seu <- readRDS(params$seurat_rds)

```

# resolved columns
```{r}
#| label: resolve-columns


md <- normalize_colnames(seu@meta.data)
seu@meta.data <- md

label_col <- first_existing_col(md, params$label_col, params$label_candidates)
sample_col <- first_existing_col(md, params$sample_col, params$sample_candidates)
patient_col <- first_existing_col(md, params$patient_col, params$patient_candidates)
condition_col <- first_existing_col(md, params$condition_col, params$condition_candidates)
timepoint_col <- first_existing_col(md, params$timepoint_col, params$timepoint_candidates)
batch_col <- first_existing_col(md, params$batch_col, params$batch_candidates)

clone_id_col <- first_existing_col(md, params$clone_id_col, params$clone_id_candidates)
clone_size_col <- first_existing_col(md, params$clone_size_col, params$clone_size_candidates)
paired_tcr_col <- first_existing_col(md, params$paired_tcr_col, params$paired_tcr_candidates)
has_tcr_col <- first_existing_col(md, params$has_tcr_col, params$has_tcr_candidates)

tcri_score_col_md <- first_existing_col(md, params$tcri_score_col, params$tcri_score_candidates)
tcri_group_col_md <- first_existing_col(md, params$tcri_group_col, params$tcri_group_candidates)
conga_cluster_col_md <- first_existing_col(md, params$conga_cluster_col, params$conga_cluster_candidates_md)
conga_score_col_md <- first_existing_col(md, params$conga_score_col, params$conga_score_candidates_md)

if (is.null(label_col)) stop("Could not resolve an annotation label column in Seurat metadata.")
if (is.null(sample_col)) stop("Could not resolve sample column in Seurat metadata.")
if (is.null(patient_col)) stop("Could not resolve patient column in Seurat metadata.")
if (is.null(condition_col)) stop("Could not resolve condition column in Seurat metadata.")
if (is.null(timepoint_col)) stop("Could not resolve timepoint column in Seurat metadata.")
if (is.null(clone_id_col)) stop("Could not resolve a clone ID column in Seurat metadata.")
if (is.null(clone_size_col)) stop("Could not resolve a clone size column in Seurat metadata.")
if (is.null(has_tcr_col)) stop("Could not resolve a has_tcr column in Seurat metadata.")
if (is.null(paired_tcr_col)) stop("Could not resolve a paired_tcr column in Seurat metadata.")

if (!"META_SAMPLE" %in% colnames(seu@meta.data) && !is.null(sample_col)) {
  seu$META_SAMPLE <- as.character(seu@meta.data[[sample_col]])
}
if (!"META_PATIENT" %in% colnames(seu@meta.data) && !is.null(patient_col)) {
  seu$META_PATIENT <- as.character(seu@meta.data[[patient_col]])
}
if (!"META_TIMECOND" %in% colnames(seu@meta.data) && !is.null(condition_col)) {
  seu$META_TIMECOND <- as.character(seu@meta.data[[condition_col]])
}
if (!"META_TIMEPOINT" %in% colnames(seu@meta.data) && !is.null(timepoint_col)) {
  seu$META_TIMEPOINT <- as.character(seu@meta.data[[timepoint_col]])
}
if (!"META_BATCH" %in% colnames(seu@meta.data) && !is.null(batch_col)) {
  seu$META_BATCH <- as.character(seu@meta.data[[batch_col]])
}

print(list(
  resolved_label_col = label_col,
  resolved_sample_col = sample_col,
  resolved_patient_col = patient_col,
  resolved_condition_col = condition_col,
  resolved_timepoint_col = timepoint_col,
  resolved_batch_col = batch_col,
  resolved_clone_id_col = clone_id_col,
  resolved_clone_size_col = clone_size_col,
  resolved_paired_tcr_col = paired_tcr_col,
  resolved_has_tcr_col = has_tcr_col,
  resolved_tcri_score_col = tcri_score_col_md,
  resolved_conga_score_col = conga_score_col_md
))


```

# subset-for-tcrdist3
```{r}
#| label: subset-for-tcrdist3
seu_t <- seu

if (isTRUE(params$subset_to_tcr_positive)) {
  keep <- !is.na(seu_t@meta.data[[has_tcr_col]]) & as.logical(seu_t@meta.data[[has_tcr_col]])
  seu_t <- subset(seu_t, cells = rownames(seu_t@meta.data)[keep])
}

if (isTRUE(params$subset_to_paired_tcr)) {
  keep <- !is.na(seu_t@meta.data[[paired_tcr_col]]) & as.logical(seu_t@meta.data[[paired_tcr_col]])
  seu_t <- subset(seu_t, cells = rownames(seu_t@meta.data)[keep])
}

if (ncol(seu_t) < params$min_cells_total) {
  stop("Too few cells remain for tcrdist3 analysis after filtering: ", ncol(seu_t))
}
```

#  build-tcrdist3-input
```{r}
#| label: build-tcrdist3-input
seu_md <- seu_t@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  mutate(
    clone_id_md = as.character(.data[[clone_id_col]]),
    cdr3_b_aa = if (isTRUE(params$derive_clone_key_from_CTaa)) {
      extract_cdr3b_from_CTaa(clone_id_md)
    } else {
      clone_id_md
    }
  ) %>%
  filter(!is.na(cdr3_b_aa), cdr3_b_aa != "")

# Try to recover TRBV / TRBJ from metadata if present, otherwise leave NA.
v_col <- intersect(c("trbv", "v_b_gene", "vb_gene", "TRBV"), colnames(seu_md))
j_col <- intersect(c("trbj", "j_b_gene", "jb_gene", "TRBJ"), colnames(seu_md))

seu_md$v_b_gene <- if (length(v_col) > 0) as.character(seu_md[[v_col[1]]]) else NA_character_
seu_md$j_b_gene <- if (length(j_col) > 0) as.character(seu_md[[j_col[1]]]) else NA_character_

# Collapse to one row per clone for tcrdist3
tcrdist_input <- seu_md %>%
  group_by(clone_id_md, cdr3_b_aa) %>%
  summarise(
    v_b_gene = first(na.omit(v_b_gene)) %||% NA_character_,
    j_b_gene = first(na.omit(j_b_gene)) %||% NA_character_,
    count = n(),
    .groups = "drop"
  ) %>%
  mutate(
    clone_id = clone_id_md
  ) %>%
  select(clone_id, cdr3_b_aa, v_b_gene, j_b_gene, count)

# tcrdist3 expects canonical column names for beta chain
tcrdist_input <- tcrdist_input %>%
  transmute(
    clone_id = clone_id,
    subject = "SCRATCH",
    cell_type = "T",
    v_b_gene = v_b_gene,
    j_b_gene = j_b_gene,
    cdr3_b_aa = cdr3_b_aa,
    count = count
  ) %>%
  filter(!is.na(cdr3_b_aa), cdr3_b_aa != "")

min_unique_clones <- params$min_unique_clones
if (is.null(min_unique_clones) || length(min_unique_clones) == 0 || is.na(min_unique_clones)) {
  min_unique_clones <- 25
}
min_unique_clones <- as.integer(min_unique_clones)

if (nrow(tcrdist_input) < min_unique_clones) {
  stop("Too few unique beta-chain clonotypes for tcrdist3: ", nrow(tcrdist_input))
}

save_table_safe(tcrdist_input, "tcrdist3_input_clones.tsv")


```

# run-tcrdist3
```{r}
#| label: run-tcrdist3
python_bin <- "/opt/conda/envs/tcrenv/bin/python"
if (!is.null(params$python_bin) && params$python_bin != "") {
  python_bin <- params$python_bin
}

Sys.setenv(RETICULATE_PYTHON = python_bin)
Sys.setenv(
  LD_LIBRARY_PATH = paste(
    "/opt/conda/envs/tcrenv/lib",
    Sys.getenv("LD_LIBRARY_PATH"),
    sep = ":"
  )
)

organism <- params$organism
if (is.null(organism) || length(organism) == 0 || is.na(organism) || organism == "") {
  organism <- "human"
}

db_file <- params$db_file
if (is.null(db_file) || length(db_file) == 0 || is.na(db_file) || db_file == "") {
  db_file <- "alphabeta_gammadelta_db.tsv"
}

tcrdist_radius <- params$tcrdist_radius
if (is.null(tcrdist_radius) || length(tcrdist_radius) == 0 || is.na(tcrdist_radius)) {
  tcrdist_radius <- 50
}
tcrdist_radius <- as.integer(tcrdist_radius)

tcrdist_input_path <- file.path(params$outdir, params$tables_dir, "tcrdist3_input_clones.tsv")
tcrdist_results_path <- file.path(params$outdir, params$tables_dir, "tcrdist3_clusters.tsv")
tcrdist_neighbors_path <- file.path(params$outdir, params$tables_dir, "tcrdist3_neighbors.tsv")
tcrdist_dist_path <- file.path(params$outdir, params$tables_dir, "tcrdist3_distance_summary.tsv")

py_script <- file.path(params$outdir, params$data_dir, "run_tcrdist3.py")
test_py <- file.path(params$outdir, params$data_dir, "test_tcrdist3_env.py")
py_log <- file.path(params$outdir, params$data_dir, "tcrdist3_python.log")

py_lines <- c(
  "import os",
  "import pandas as pd",
  "import numpy as np",
  "from scipy.sparse.csgraph import connected_components",
  "from tcrdist.repertoire import TCRrep",
  "",
  sprintf("tcrdist_input_path = r'%s'", tcrdist_input_path),
  sprintf("tcrdist_results_path = r'%s'", tcrdist_results_path),
  sprintf("tcrdist_neighbors_path = r'%s'", tcrdist_neighbors_path),
  sprintf("tcrdist_dist_path = r'%s'", tcrdist_dist_path),
  sprintf("organism = r'%s'", organism),
  sprintf("db_file = r'%s'", db_file),
  sprintf("radius = %d", tcrdist_radius),
  "",
  "print('Running TCRdist3 python step')",
  "print('Input table:', tcrdist_input_path)",
  "print('Organism:', organism)",
  "print('DB file:', db_file)",
  "print('Radius:', radius)",
  "",
  "df = pd.read_csv(tcrdist_input_path, sep='\\t')",
  "print('Input shape:', df.shape)",
  "",
  "if df.shape[0] == 0:",
  "    raise ValueError('tcrdist_input is empty.')",
  "",
  "print('Starting input cleanup and gene normalization')",
  "",
  "# Basic cleanup",
  "for col in ['clone_id', 'cdr3_b_aa', 'v_b_gene', 'j_b_gene']:",
  "    if col in df.columns:",
  "        df[col] = df[col].astype(str).str.strip()",
  "",
  "df = df.replace({'nan': np.nan, 'None': np.nan, '': np.nan})",
  "",
  "# Keep only rows with essential beta-chain fields",
  "df = df.dropna(subset=['clone_id', 'cdr3_b_aa', 'v_b_gene', 'j_b_gene']).copy()",
  "print('After dropping missing clone/V/J/CDR3 fields:', df.shape)",
  "",
  "# Normalize gene names",
  "df['v_b_gene'] = df['v_b_gene'].str.replace(r'\\s+', '', regex=True)",
  "df['j_b_gene'] = df['j_b_gene'].str.replace(r'\\s+', '', regex=True)",
  "",
  "# If allele is missing, add *01",
  "df['v_b_gene'] = df['v_b_gene'].apply(lambda x: x if pd.isna(x) else (x if '*' in x else x + '*01'))",
  "df['j_b_gene'] = df['j_b_gene'].apply(lambda x: x if pd.isna(x) else (x if '*' in x else x + '*01'))",
  "",
  "# Keep only beta-chain V/J genes",
  "df = df[",
  "    df['v_b_gene'].str.startswith('TRBV', na=False) &",
  "    df['j_b_gene'].str.startswith('TRBJ', na=False)",
  "].copy()",
  "print('After TRBV/TRBJ prefix filter:', df.shape)",
  "",
  "if df.shape[0] == 0:",
  "    raise ValueError('No valid beta-chain rows remain after V/J filtering.')",
  "",
  "if df.shape[0] < 10:",
  "    raise ValueError(f'Too few valid clones remain for tcrdist3 after gene filtering: {df.shape[0]}')",
  "",
  "print('Example normalized genes:')",
  "print(df[['v_b_gene','j_b_gene']].drop_duplicates().head(20).to_string(index=False))",
  "",
  "tr = TCRrep(",
  "    cell_df=df,",
  "    organism=organism,",
  "    chains=['beta'],",
  "    db_file=db_file,",
  "    infer_all_genes=True,",
  "    infer_cdrs=True,",
  "    deduplicate=False,",
  "    compute_distances=False",
  ")",
  "",
  "# Compatibility across tcrdist3 versions",
  "if hasattr(tr, 'compute_distances'):",
  "    tr.compute_distances()",
  "",
  "pw = None",
  "for attr in ['pw_beta', 'pw_tcrdist', 'pw_cdr3_b_aa', 'pw_beta_tcrdist']:",
  "    if hasattr(tr, attr):",
  "        candidate = getattr(tr, attr)",
  "        if candidate is not None:",
  "            pw = np.asarray(candidate)",
  "            print(f'Using distance matrix attribute: {attr}')",
  "            break",
  "",
  "if pw is None:",
  "    for attr in dir(tr):",
  "        if attr.startswith('pw'):",
  "            candidate = getattr(tr, attr)",
  "            try:",
  "                arr = np.asarray(candidate)",
  "                if arr.ndim == 2 and arr.shape[0] == arr.shape[1] and arr.shape[0] == df.shape[0]:",
  "                    pw = arr",
  "                    print(f'Using fallback distance matrix attribute: {attr}')",
  "                    break",
  "            except Exception:",
  "                pass",
  "",
  "if pw is None:",
  "    raise AttributeError('Could not find a usable pairwise distance matrix on the TCRrep object.')",
  "",
  "adj = (pw <= radius).astype(int)",
  "np.fill_diagonal(adj, 0)",
  "",
  "n_components, labels = connected_components(adj, directed=False, connection='weak')",
  "cluster_sizes = pd.Series(labels).value_counts().to_dict()",
  "",
  "clone_df = df.copy()",
  "clone_df['tcrdist_cluster'] = labels.astype(str)",
  "clone_df['tcrdist_group'] = labels.astype(str)",
  "clone_df['tcrdist_score'] = adj.sum(axis=1).astype(float)",
  "clone_df['tcrdist_radius'] = float(radius)",
  "clone_df['tcrdist_pvalue'] = np.nan",
  "clone_df['tcrdist_qvalue'] = np.nan",
  "clone_df['tcrdist_cluster_size'] = [cluster_sizes[x] for x in labels]",
  "clone_df['metaclonotype'] = ['MC_' + str(x) for x in labels]",
  "",
  "clone_df[['clone_id','cdr3_b_aa','tcrdist_cluster','tcrdist_group','tcrdist_score','tcrdist_radius','tcrdist_pvalue','tcrdist_qvalue','tcrdist_cluster_size','metaclonotype']].to_csv(",
  "    tcrdist_results_path, sep='\\t', index=False",
  ")",
  "",
  "rows, cols = np.where(np.triu(adj, k=1) > 0)",
  "edges = pd.DataFrame({",
  "    'source': df.iloc[rows]['clone_id'].astype(str).values,",
  "    'target': df.iloc[cols]['clone_id'].astype(str).values,",
  "    'distance': pw[rows, cols].astype(float),",
  "    'radius': float(radius)",
  "})",
  "edges.to_csv(tcrdist_neighbors_path, sep='\\t', index=False)",
  "",
  "dist_summary = pd.DataFrame({",
  "    'metric': ['n_clones','radius','mean_degree','median_degree','max_degree'],",
  "    'value': [",
  "        int(df.shape[0]),",
  "        float(radius),",
  "        float(adj.sum(axis=1).mean()),",
  "        float(np.median(adj.sum(axis=1))),",
  "        float(adj.sum(axis=1).max())",
  "    ]",
  "})",
  "dist_summary.to_csv(tcrdist_dist_path, sep='\\t', index=False)",
  "",
  "print('TCRdist3 python step completed successfully')"
)
writeLines(py_lines, py_script)


writeLines(
  c(
    "import sys",
    "print(sys.executable)",
    "import pandas",
    "import numpy",
    "import scipy",
    "from tcrdist.repertoire import TCRrep",
    "print('PYTHON IMPORT SUCCESS!')"
  ),
  test_py
)

test_res <- system2(
  python_bin,
  args = test_py,
  stdout = TRUE,
  stderr = TRUE
)

test_status <- attr(test_res, "status")
if (is.null(test_status)) test_status <- 0L

if (test_status != 0L) {
  writeLines(test_res, py_log)
  stop(
    paste0(
      "Python environment test failed before running tcrdist3.\n",
      "See log: ", py_log, "\n\n",
      paste(test_res, collapse = "\n")
    )
  )
}

res <- system2(
  python_bin,
  args = py_script,
  stdout = TRUE,
  stderr = TRUE
)

status <- attr(res, "status")
if (is.null(status)) status <- 0L

writeLines(
  c(
    "=== Python environment test ===",
    test_res,
    "",
    "=== TCRdist3 run ===",
    res
  ),
  py_log
)

if (status != 0L) {
  stop(
    paste0(
      "TCRdist3 Python script failed with exit status ", status, ".\n",
      "See log: ", py_log, "\n\n",
      paste(res, collapse = "\n")
    )
  )
}

if (!file.exists(tcrdist_results_path)) {
  stop("TCRdist3 did not produce the expected cluster output file: ", tcrdist_results_path)
}
if (!file.exists(tcrdist_neighbors_path)) {
  stop("TCRdist3 did not produce the expected neighbors output file: ", tcrdist_neighbors_path)
}
if (!file.exists(tcrdist_dist_path)) {
  stop("TCRdist3 did not produce the expected distance summary file: ", tcrdist_dist_path)
}

cat("TCRdist3 Python step completed successfully.\n")
cat("Detailed Python log saved to: ", py_log, "\n", sep = "")
```

# read-tcrdist3-results
```{r}
#| label: read-tcrdist3-results
tcrdist_clusters_path <- file.path(params$outdir, params$tables_dir, "tcrdist3_clusters.tsv")
tcrdist_neighbors_path <- file.path(params$outdir, params$tables_dir, "tcrdist3_neighbors.tsv")
tcrdist_dist_path <- file.path(params$outdir, params$tables_dir, "tcrdist3_distance_summary.tsv")

require_file(tcrdist_clusters_path, "Missing tcrdist3 cluster results file.")
tcrdist_clusters <- fread(tcrdist_clusters_path)

tcrdist_neighbors <- if (file.exists(tcrdist_neighbors_path)) {
  fread(tcrdist_neighbors_path)
} else {
  data.table::data.table()
}

tcrdist_distance_summary <- if (file.exists(tcrdist_dist_path)) {
  fread(tcrdist_dist_path)
} else {
  data.table::data.table()
}

if (!"clone_id" %in% colnames(tcrdist_clusters)) {
  stop("tcrdist_clusters.tsv is missing required column: clone_id")
}
if (!"cdr3_b_aa" %in% colnames(tcrdist_clusters)) {
  stop("tcrdist_clusters.tsv is missing required column: cdr3_b_aa")
}
if (!"tcrdist_cluster" %in% colnames(tcrdist_clusters)) {
  stop("tcrdist_clusters.tsv is missing required column: tcrdist_cluster")
}

```

# Merge TCRdist3 annotations back into Seurat
```{r}
#| label: merge-tcrdist-into-seurat
seu_md_full <- seu@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  mutate(
    clone_id_md = as.character(.data[[clone_id_col]]),
    clone_key = if (isTRUE(params$derive_clone_key_from_CTaa)) {
      extract_cdr3b_from_CTaa(clone_id_md)
    } else {
      clone_id_md
    }
  )

tcrdist_map <- tcrdist_clusters %>%
  mutate(
    clone_id = as.character(clone_id),
    clone_key = as.character(cdr3_b_aa)
  ) %>%
  select(
    clone_id,
    clone_key,
    tcrdist_cluster,
    tcrdist_group,
    tcrdist_score,
    tcrdist_radius,
    tcrdist_pvalue,
    tcrdist_qvalue,
    tcrdist_cluster_size,
    metaclonotype
  ) %>%
  distinct(clone_key, .keep_all = TRUE)

seu_md_full <- seu_md_full %>%
  left_join(
    tcrdist_map %>%
      select(
        clone_key,
        tcrdist_cluster,
        tcrdist_group,
        tcrdist_score,
        tcrdist_radius,
        tcrdist_pvalue,
        tcrdist_qvalue,
        tcrdist_cluster_size,
        metaclonotype
      ),
    by = "clone_key"
  )

rownames(seu_md_full) <- seu_md_full$cell_id
seu@meta.data <- seu_md_full %>% select(-cell_id)

save_rds_safe(seu, "seurat_with_TCRdist3.rds")
```

# ensure-embedding
```{r}
#| label: ensure-embedding
reduction_to_use <- params$reduction_use
if (!(reduction_to_use %in% names(seu@reductions)) && isTRUE(params$make_umap_if_missing)) {
  seu <- safe_make_umap(
    seu,
    reduction_name = params$reduction_use,
    dims = 1:params$umap_dims_max,
    nfeatures = params$umap_nfeatures
  )
}
if (!(reduction_to_use %in% names(seu@reductions))) {
  reduction_to_use <- if ("umap" %in% names(seu@reductions)) "umap" else if ("tsne" %in% names(seu@reductions)) "tsne" else if ("pca" %in% names(seu@reductions)) "pca" else stop("No usable embedding found.")
}
```

# build-export
```{r}
#| label: build-export
# Explicitly mapping resolved names into the export table
export_cells <- seu@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  transmute(
    cell_id = cell_id,
    sample    = as.character(.data[[sample_col]]),
    patient   = as.character(.data[[patient_col]]),
    condition = as.character(.data[[condition_col]]),
    annot     = as.character(.data[[label_col]]),
    clone_id  = as.character(.data[[clone_id_col]]),
    # TCRdist3 results
    tcrdist_cluster = as.character(tcrdist_cluster),
    tcrdist_score   = suppressWarnings(as.numeric(tcrdist_score)),
    metaclonotype   = as.character(metaclonotype)
  ) %>%
  left_join(as.data.frame(Embeddings(seu, reduction_to_use)) %>% tibble::rownames_to_column("cell_id"), by = "cell_id")

# Create the display labels (DIST_01...) for the plots
cluster_summary <- export_cells %>%
  filter(!is.na(tcrdist_cluster)) %>%
  count(tcrdist_cluster, name = "n_cells") %>%
  mutate(tcrdist_display = paste0("DIST_", sprintf("%02d", seq_len(n()))))

export_cells <- export_cells %>% 
  left_join(cluster_summary %>% select(tcrdist_cluster, tcrdist_display), by = "tcrdist_cluster")

save_table_safe(export_cells, "tcrdist3_export_cells.tsv")


# #| label: build-export
# embed_df <- as.data.frame(Embeddings(seu, reduction = reduction_to_use)) %>%
#   tibble::rownames_to_column("cell_id")
# 
# export_cells <- seu@meta.data %>%
#   tibble::rownames_to_column("cell_id") %>%
#   transmute(
#     cell_id = cell_id,
#     sample = if (!is.null(sample_col) && sample_col %in% colnames(seu@meta.data)) as.character(.data[[sample_col]]) else NA_character_,
#     patient = if (!is.null(patient_col) && patient_col %in% colnames(seu@meta.data)) as.character(.data[[patient_col]]) else NA_character_,
#     condition = if (!is.null(condition_col) && condition_col %in% colnames(seu@meta.data)) as.character(.data[[condition_col]]) else NA_character_,
#     timepoint = if (!is.null(timepoint_col) && timepoint_col %in% colnames(seu@meta.data)) as.character(.data[[timepoint_col]]) else NA_character_,
#     batch = if (!is.null(batch_col) && batch_col %in% colnames(seu@meta.data)) as.character(.data[[batch_col]]) else NA_character_,
#     annot = if (!is.null(label_col) && label_col %in% colnames(seu@meta.data)) as.character(.data[[label_col]]) else NA_character_,
#     clone_id = as.character(.data[[clone_id_col]]),
#     clone_size = if (!is.null(clone_size_col) && clone_size_col %in% colnames(seu@meta.data)) suppressWarnings(as.numeric(.data[[clone_size_col]])) else NA_real_,
#     paired_tcr = if (!is.null(paired_tcr_col) && paired_tcr_col %in% colnames(seu@meta.data)) as.logical(.data[[paired_tcr_col]]) else NA,
#     has_tcr = if (!is.null(has_tcr_col) && has_tcr_col %in% colnames(seu@meta.data)) as.logical(.data[[has_tcr_col]]) else NA,
#     tcrdist_cluster = if ("tcrdist_cluster" %in% colnames(seu@meta.data)) as.character(tcrdist_cluster) else NA_character_,
#     tcrdist_group = if ("tcrdist_group" %in% colnames(seu@meta.data)) as.character(tcrdist_group) else NA_character_,
#     tcrdist_score = if ("tcrdist_score" %in% colnames(seu@meta.data)) suppressWarnings(as.numeric(tcrdist_score)) else NA_real_,
#     tcrdist_radius = if ("tcrdist_radius" %in% colnames(seu@meta.data)) suppressWarnings(as.numeric(tcrdist_radius)) else NA_real_,
#     tcrdist_pvalue = if ("tcrdist_pvalue" %in% colnames(seu@meta.data)) suppressWarnings(as.numeric(tcrdist_pvalue)) else NA_real_,
#     tcrdist_qvalue = if ("tcrdist_qvalue" %in% colnames(seu@meta.data)) suppressWarnings(as.numeric(tcrdist_qvalue)) else NA_real_,
#     tcrdist_cluster_size = if ("tcrdist_cluster_size" %in% colnames(seu@meta.data)) suppressWarnings(as.numeric(tcrdist_cluster_size)) else NA_real_,
#     metaclonotype = if ("metaclonotype" %in% colnames(seu@meta.data)) as.character(metaclonotype) else NA_character_
#   ) %>%
#   left_join(embed_df, by = "cell_id")
# 
# save_table_safe(export_cells, "tcrdist3_export_cells.tsv")

```

# Merge diagnostics
```{r}
#| label: merge-diagnostics
merge_diag <- tibble::tibble(
  metric = c(
    "n_cells_in_seurat",
    "n_cells_in_export",
    "n_unique_clones_input",
    "n_unique_clones_clustered",
    "n_neighbor_edges",
    "n_distance_summary_rows"
  ),
  value = c(
    ncol(seu),
    nrow(export_cells),
    dplyr::n_distinct(export_cells$clone_id[!is.na(export_cells$clone_id)]),
    if (exists("tcrdist_clusters") && nrow(tcrdist_clusters) > 0) {
      dplyr::n_distinct(tcrdist_clusters$clone_id[!is.na(tcrdist_clusters$clone_id)])
    } else {
      0
    },
    if (exists("tcrdist_neighbors") && nrow(tcrdist_neighbors) > 0) {
      nrow(tcrdist_neighbors)
    } else {
      0
    },
    if (exists("tcrdist_distance_summary") && nrow(tcrdist_distance_summary) > 0) {
      nrow(tcrdist_distance_summary)
    } else {
      0
    }
  )
)

save_table_safe(merge_diag, "tcrdist3_merge_diagnostics.tsv")
```

# summary-tables
```{r}
#| label: summary-tables
#| results: hide

summary_rollup <- tibble(
  metric = c(
    "Report label",
    "Cells in Seurat",
    "Cells with TCRdist3 clusters",
    "Cells with TCRdist3 scores",
    "Cells with metaclonotype",
    "TCR-positive cells",
    "Paired TCR cells",
    "Unique TCRdist3 clusters",
    "Unique annotations",
    "Unique samples",
    "Selected embedding"
  ),
  value = c(
    params$report_label,
    nrow(export_cells),
    sum(!is.na(export_cells$tcrdist_cluster)),
    sum(!is.na(export_cells$tcrdist_score)),
    sum(!is.na(export_cells$metaclonotype)),
    if ("has_tcr" %in% colnames(seu@meta.data)) sum(seu@meta.data[[has_tcr_col]], na.rm = TRUE) else NA,
    if ("paired_tcr" %in% colnames(seu@meta.data)) sum(seu@meta.data[[paired_tcr_col]], na.rm = TRUE) else NA,
    length(unique(na.omit(export_cells$tcrdist_cluster))),
    length(unique(na.omit(export_cells$annot))),
    length(unique(na.omit(export_cells$sample))),
    reduction_to_use
  )
)
save_table_safe(summary_rollup, "tcrdist3_summary_rollup.tsv")

cluster_summary <- export_cells %>%
  filter(!is.na(tcrdist_cluster)) %>%
  count(tcrdist_cluster, name = "n_cells") %>%
  mutate(frac = n_cells / sum(n_cells)) %>%
  arrange(desc(n_cells)) %>%
  mutate(tcrdist_display = paste0("DIST_", sprintf("%02d", seq_len(n()))))

cluster_label_map <- cluster_summary %>%
  select(tcrdist_cluster, tcrdist_display)

export_cells <- export_cells %>%
  left_join(cluster_label_map, by = "tcrdist_cluster")

seu@meta.data$tcrdist_display <- export_cells$tcrdist_display[
  match(rownames(seu@meta.data), export_cells$cell_id)
]

cluster_annotation_tbl <- export_cells %>%
  filter(!is.na(tcrdist_cluster), !is.na(annot)) %>%
  count(tcrdist_cluster, annot, name = "n_cells")

cluster_sample_tbl <- export_cells %>%
  filter(!is.na(tcrdist_cluster), !is.na(sample)) %>%
  count(tcrdist_cluster, sample, name = "n_cells")

annotation_tcrdist_summary <- export_cells %>%
  filter(!is.na(annot)) %>%
  group_by(annot) %>%
  summarise(
    n_cells = n(),
    n_tcrdist = sum(!is.na(tcrdist_cluster)),
    frac_tcrdist = n_tcrdist / n_cells,
    mean_tcrdist = mean(tcrdist_score, na.rm = TRUE),
    median_tcrdist = median(tcrdist_score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(frac_tcrdist))

save_table_safe(cluster_summary, "tcrdist3_cluster_summary.tsv")
save_table_safe(annotation_tcrdist_summary, "annotation_tcrdist3_summary.tsv")
save_table_safe(cluster_annotation_tbl, "tcrdist3_cluster_annotation_composition.tsv")
save_table_safe(cluster_sample_tbl, "tcrdist3_cluster_sample_composition.tsv")
save_table_safe(export_cells, "tcrdist3_export_cells.tsv")



```

# overview
```{r}
#| label: overview
#| results: asis

cat("### High-Level Analysis Overview\n")

if (exists("summary_rollup") && nrow(summary_rollup) > 0) {
  print(
    kable(summary_rollup, caption = "High-level overview of TCRdist3 analysis.") %>%
      kable_styling(
        full_width = FALSE,
        bootstrap_options = c("striped", "hover", "condensed")
      )
  )
} else {
  cat("Overview table was not available because summary statistics were not generated.\n")
}


```

# score-density
```{r}
#| label: score-density
if (isTRUE(params$show_score_density) && any(!is.na(export_cells$tcrdist_score))) {
  p_density <- export_cells %>%
    filter(!is.na(tcrdist_score)) %>%
    ggplot(aes(x = tcrdist_score)) +
    geom_density(fill = "steelblue", alpha = 0.5, linewidth = 0.8) +
    labs(
      title = "Distribution of TCRdist3 scores",
      subtitle = "Mapped clone-level neighborhood degree.",
      x = "TCRdist3 score",
      y = "Density"
    ) +
    theme_scratch_pub(params$base_size)

  print(p_density)
  save_plot_safe(p_density, glue("tcrdist3_score_density.{params$figure_format}"))
}
```

# Cluster-umap
```{r}
#| label: cluster-umap
#| fig-width: 12
#| fig-height: 9

if (isTRUE(params$show_cluster_umap) && any(!is.na(export_cells$tcrdist_display))) {

  cells_dist <- colnames(seu)[!is.na(seu@meta.data$tcrdist_display) & seu@meta.data$tcrdist_display != ""]
  n_dist <- length(cells_dist)
  n_total <- ncol(seu)

  cat(sprintf(
    "**TCRdist3 cluster assignment note:** %s of %s cells (%.2f%%) received a TCRdist3 cluster assignment. Only assigned cells are highlighted by TCRdist3 display labels.\n\n",
    scales::comma(n_dist),
    scales::comma(n_total),
    100 * n_dist / max(1, n_total)
  ))

  p_cluster <- DimPlot(
    seu,
    reduction = reduction_to_use,
    group.by = "tcrdist_display",
    label = TRUE,
    repel = TRUE,
    raster = isTRUE(params$raster_large_umap)
  ) +
    ggtitle("Embedding overlay: TCRdist3 clusters") +
    theme_scratch_pub(params$base_size) +
    theme(
      legend.position = "none"
    )

  print(p_cluster)

  ggplot2::ggsave(
    filename = file.path(params$outdir, params$figures_dir, glue("tcrdist3_cluster_umap.{params$figure_format}")),
    plot = p_cluster,
    width = 12,
    height = 9,
    dpi = params$figure_dpi,
    bg = "white",
    limitsize = FALSE
  )
} else {
  cat("No TCRdist3 cluster UMAP was produced because no TCRdist3-assigned cells were available.\n")
}

# #| label: cluster-umap
# #| fig-width: 16
# #| fig-height: 12
# if (isTRUE(params$show_cluster_umap) && any(!is.na(export_cells$gliph_cluster_display))) {
# 
#   seu$gliph_cluster_display <- export_cells$gliph_cluster_display[
#     match(colnames(seu), export_cells$cell_id)
#   ]
# 
#   p_cluster <- DimPlot(
#     seu,
#     reduction = reduction_to_use,
#     group.by = "gliph_cluster_display",
#     label = TRUE,
#     repel = TRUE,
#     raster = isTRUE(params$raster_large_umap)
#   ) +
#     ggtitle("Embedding overlay: GLIPH2 clusters") +
#     theme_scratch_pub(params$base_size) +
#     theme(
#       legend.text = element_text(size = 7),
#       legend.title = element_text(size = 9)
#     )
# 
#   tryCatch({
#     print(p_cluster)
#   }, error = function(e) {
#     message("Inline patchwork rendering failed; figure will still be saved.")
#   })
# 
#   save_plot_safe(
#     p_cluster,
#     glue("gliph2_cluster_umap.{params$figure_format}"),
#     width = 16,
#     height = 12
#   )
# }
```

# Cluster size barplot
```{r}
#| label: cluster-size-barplot
if (nrow(cluster_summary) > 0) {
  p_bar <- cluster_summary %>%
    slice_head(n = 20) %>%
    mutate(tcrdist_display = fct_reorder(tcrdist_display, n_cells)) %>%
    ggplot(aes(x = tcrdist_display, y = n_cells)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    labs(title = "Top TCRdist3 Structural Clusters", x = NULL, y = "Cell Count") +
    theme_scratch_pub()
  
  print(p_bar)
  save_plot_safe(p_bar, glue("tcrdist3_cluster_size_barplot.{params$figure_format}"))
}


```

# annotation-heatmap
```{r}
#| label: annotation-heatmap
#| fig-width: 14
#| fig-height: 10
#| results: asis

if (exists("cluster_annotation_tbl") && nrow(cluster_annotation_tbl) > 0) {

  top_cls <- cluster_summary %>%
    slice_head(n = params$top_n_clusters) %>%
    pull(tcrdist_cluster)

  mat_df <- cluster_annotation_tbl %>%
    filter(tcrdist_cluster %in% top_cls) %>%
    left_join(cluster_label_map, by = "tcrdist_cluster") %>%
    pivot_wider(names_from = annot, values_from = n_cells, values_fill = 0) %>%
    as.data.frame()

  if (nrow(mat_df) > 0 && ncol(mat_df) > 2) {
    rownames(mat_df) <- mat_df$tcrdist_display
    mat <- as.matrix(
      mat_df[, !(colnames(mat_df) %in% c("tcrdist_cluster", "tcrdist_display")), drop = FALSE]
    )

    cond_anno <- export_cells %>%
      filter(tcrdist_cluster %in% top_cls, !is.na(condition)) %>%
      left_join(cluster_label_map, by = "tcrdist_cluster") %>%
      group_by(tcrdist_display) %>%
      summarise(
        dominant = tools::toTitleCase(tolower(names(which.max(table(condition))))),
        .groups = "drop"
      )

    cond_vec <- cond_anno$dominant[match(rownames(mat), cond_anno$tcrdist_display)]
    cond_vec[is.na(cond_vec)] <- "Unknown"

    ann_colors <- c(
      "Pre" = "#66c2a5",
      "Post" = "#fc8d62",
      "Normal" = "#8da0cb",
      "Unknown" = "grey80"
    )

    row_ha <- rowAnnotation(
      Clinical_State = cond_vec,
      col = list(Clinical_State = ann_colors),
      gp = gpar(col = "white", lwd = 1)
    )

    max_mat <- max(mat, na.rm = TRUE)
    if (!is.finite(max_mat) || max_mat <= 0) max_mat <- 1

    cat("### TCRdist3 Structural Specificity vs. Cell State\n")

    ht <- Heatmap(
      mat,
      name = "Cells",
      right_annotation = row_ha,
      col = colorRamp2(c(0, max_mat / 2, max_mat), c("#f7fbff", "#6baed6", "#08306b")),
      cluster_rows = TRUE,
      cluster_columns = TRUE,
      rect_gp = gpar(col = "white", lwd = 1),
      column_title = "Cell Type Composition of Top Structural Clusters",
      row_names_gp = gpar(fontsize = 10, fontface = "bold"),
      column_names_rot = 45
    )

    draw(ht, heatmap_legend_side = "right")
  } else {
    cat("*Insufficient data to generate the annotation heatmap.*")
  }

} else {
  cat("*Insufficient data in `cluster_annotation_tbl` to generate heatmap.*")
}

```

# sample-heatmap
```{r}
#| label: sample-heatmap
#| fig-width: 14
#| fig-height: 10
#| results: asis

if (isTRUE(params$show_sample_heatmap) && exists("cluster_sample_tbl") && nrow(cluster_sample_tbl) > 0) {

  top_clusters <- cluster_summary %>%
    filter(n_cells >= params$min_cluster_size_plot) %>%
    slice_head(n = params$top_n_clusters) %>%
    pull(tcrdist_cluster)

  mat_df <- cluster_sample_tbl %>%
    filter(tcrdist_cluster %in% top_clusters) %>%
    left_join(cluster_label_map, by = "tcrdist_cluster") %>%
    pivot_wider(names_from = sample, values_from = n_cells, values_fill = 0) %>%
    as.data.frame()

  if (nrow(mat_df) > 0 && ncol(mat_df) > 2) {
    rownames(mat_df) <- mat_df$tcrdist_display
    mat <- as.matrix(mat_df[, !(colnames(mat_df) %in% c("tcrdist_cluster", "tcrdist_display")), drop = FALSE])

    max_mat <- max(mat, na.rm = TRUE)
    if (!is.finite(max_mat) || max_mat <= 0) max_mat <- 1

    cat("### Sample-wise structural distribution\n")

    ht <- Heatmap(
      mat,
      name = "Cells",
      col = colorRamp2(
        c(0, max_mat / 2, max_mat),
        c("white", "skyblue", "navy")
      ),
      cluster_rows = TRUE,
      cluster_columns = TRUE,
      row_names_side = "left",
      row_names_gp = grid::gpar(fontsize = 9, fontface = "bold"),
      column_names_gp = grid::gpar(fontsize = 9),
      column_names_rot = 45,
      column_title = "Sample composition of TCRdist3 clusters",
      heatmap_legend_param = list(title = "Cells")
    )

    draw(ht, heatmap_legend_side = "right")
  } else {
    cat("*Insufficient cluster-by-sample matrix to generate the sample heatmap.*")
  }

} else {
  cat("*No sample heatmap was produced because no TCRdist3 cluster/sample composition table was available.*")
}

```

# Fraction with TCRdist3 annotation by grouping
```{r}
#| label: fraction-plots
#| fig-height: 8
#| fig-width: 10
#| results: asis

# 1. Advanced Plotting Function
# We use a robust internal check to ensure we don't crash on empty data
plot_tcrdist_enrichment <- function(df, col_name, title_txt, fill_color = "steelblue") {
  if (!col_name %in% colnames(df)) return(NULL)
  
  # Calculate summary on the fly to ensure fresh data
  plot_data <- df %>%
    filter(!is.na(.data[[col_name]])) %>%
    group_by(label = .data[[col_name]]) %>%
    summarise(
      n_total = n(),
      n_clust = sum(!is.na(tcrdist_cluster)),
      # Using our global safe_percent helper
      frac = safe_percent(n_clust, n_total),
      .groups = "drop"
    ) %>%
    filter(n_total >= params$min_cells_per_group)
  
  if (nrow(plot_data) < 2) return(NULL)

  p <- plot_data %>%
    mutate(label = fct_reorder(label, frac)) %>%
    ggplot(aes(x = label, y = frac)) +
    geom_col(fill = fill_color, width = 0.7) +
    coord_flip() +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(
      title = title_txt,
      subtitle = paste0("Minimum cells per group: ", params$min_cells_per_group),
      x = NULL, y = "Percentage of Clones in TCRdist3 Clusters"
    ) +
    theme_scratch_pub(params$base_size)
  
  return(p)
}

# 2. Execution Blocks
# We wrap these in 'print' and 'cat' to ensure Quarto places them correctly in the HTML
cat("### Structural Enrichment by Category\n")

# A. By Cell State (Annotation)
p1 <- plot_tcrdist_enrichment(export_cells, "annot", "Enrichment by Cell State", "darkseagreen")
if (!is.null(p1)) {
  print(p1)
  save_plot_safe(p1, "tcrdist3_enrichment_annotation.png")
}

# B. By Clinical Condition (The most important for your Ovarian cohort)
p2 <- plot_tcrdist_enrichment(export_cells, "condition", "Enrichment by Clinical Condition", "#fc8d62")
if (!is.null(p2)) {
  cat("\n\n")
  print(p2)
  save_plot_safe(p2, "tcrdist3_enrichment_condition.png")
}

# C. By Sample
p3 <- plot_tcrdist_enrichment(export_cells, "sample", "Enrichment by Sample", "#8da0cb")
if (!is.null(p3)) {
  cat("\n\n")
  print(p3)
  save_plot_safe(p3, "tcrdist3_enrichment_sample.png")
}
```

# association-plots
```{r}
#| label: association-plots
#| results: asis
#| fig-width: 10
#| fig-height: 6

cat("### TCRdist3 vs. Integration Scores\n")

# Check if TCRi scores exist in the original metadata
tcri_col <- intersect(c("tcri_score", "TCRi_score"), colnames(seu@meta.data))

if (length(tcri_col) > 0 && any(!is.na(export_cells$tcrdist_score))) {
  p_assoc <- export_cells %>%
    mutate(tcri_score = seu@meta.data[[tcri_col[1]]][match(cell_id, rownames(seu@meta.data))]) %>%
    filter(!is.na(tcri_score), !is.na(tcrdist_score)) %>%
    ggplot(aes(x = tcri_score, y = tcrdist_score)) +
    geom_point(alpha = 0.2, color = "midnightblue", size = 0.8) +
    geom_smooth(method = "lm", color = "firebrick", linewidth = 1) +
    labs(title = "TCRi vs. TCRdist3 Convergence", x = "TCRi Score", y = "TCRdist3 Neighborhood Score") +
    theme_scratch_pub()
  
  print(p_assoc)
} else {
  cat("*TCRi scores not found or insufficient overlap for association plotting.*")
}

```

# Neighbor summary
```{r}
#| label: neighbor-summary
#| results: asis

cat("### Structural Neighbor Connectivity\n")

if (exists("tcrdist_neighbors") && nrow(tcrdist_neighbors) > 0) {
  neigh_sum <- tibble(
    Metric = c("Total Structural Edges", "Unique Clones in Networks", "Maximum Neighborhood Degree"),
    Value = c(
      nrow(tcrdist_neighbors),
      length(unique(c(tcrdist_neighbors$source, tcrdist_neighbors$target))),
      max(export_cells$tcrdist_score, na.rm = TRUE)
    )
  )
  
  print(kable(neigh_sum, caption = "Summary of TCR structural connectivity.") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")))
} else {
  cat("*No structural neighbors identified within the defined radius.*")
}

```

# summary-tables-display
```{r}
#| label: summary-tables-display
#| results: asis

cat("### Top Structural Clusters\n")
print(kableExtra::kbl(cluster_summary %>% head(20), caption = "Top clusters by cell count") %>%
  kableExtra::kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed")))


# #| label: summary-tables-display
# #| results: asis
# 
# if (isTRUE(params$show_summary_tables)) {
#   
#   # Helper function for professional styling
#   render_adv_table <- function(df, cap) {
#     print(kableExtra::kbl(df, caption = cap) %>%
#       kableExtra::kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed")))
#   }
# 
#   cat("### TCRdist3 High-Level Overview\n")
#   render_adv_table(summary_rollup, "Analysis Statistics")
#   cat("\n\n<br>\n\n")
# 
#   cat("### Top Structural TCRdist3 Clusters\n")
#   render_adv_table(cluster_summary %>% dplyr::select(tcrdist_display, n_cells, frac) %>% head(20), 
#                    "Top Clusters by Structural Similarity")
#   cat("\n\n<br>\n\n")
# 
#   cat("### Integration Diagnostics\n")
#   render_adv_table(merge_diag, "Metadata Mapping Diagnostics")
# }

```

# warnings
```{r}
#| label: warnings
#| results: asis

warn_tbl <- tibble(
  warning = c(
    "Low TCRdist3 clustering",
    "Low neighborhood scores",
    "Missing Condition metadata",
    "Missing Patient metadata"
  ),
  triggered = c(
    mean(!is.na(export_cells$tcrdist_cluster)) < 0.2,
    mean(export_cells$tcrdist_score > 0, na.rm = TRUE) < 0.2,
    all(is.na(export_cells$condition)),
    all(is.na(export_cells$patient))
  ),
  interpretation = c(
    "Less than 20% of cells formed structural clusters. Consider increasing the TCRdist3 radius.",
    "Most clones are structural singletons. Neighborhood connectivity is low for this cohort.",
    "Condition metadata (e.g., Pre/Post) was not found in the final export table.",
    "Patient ID metadata was not found in the final export table."
  )
) %>%
  filter(triggered)

if (nrow(warn_tbl) == 0) {
  cat("### <span style='color:green;'>System Status: Optimized</span>\n")
  cat("No major warnings triggered for TCRdist3 structural analysis.")
} else {
  cat("### <span style='color:orange;'>Analysis Warnings</span>\n")
  print(kable(warn_tbl %>% select(-triggered), caption = "TCRdist3 specific diagnostics.") %>%
    kable_styling(full_width = TRUE, bootstrap_options = c("striped", "hover", "condensed")))
}

```

# session info
```{r}
#| label: session-info
writeLines(capture.output(sessionInfo()), file.path(params$outdir, "sessionInfo.tcrdist3.txt"))
sessionInfo()
```



